---
title: "Map the Meal Gap 2020 Projections"
author: "Xanna Burg"
date: "6/10/2020"
output: html_document
---

## Indicator 1: (Special Projection) Child food insecurity by county in 2020

**Created by:** Xanna Burg
**Date:** June 2020
**Updated by:**

**Data Source:** Feeding America, The Impact of the Coronavirus on Food Security
**Purpose:** Input the special report of data, downloaded from the Tableau dashboard

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
#input location ID file
MT_locationids <- read.csv("./Input/MT KC Location IDs.csv")
MT_locationids$Location <- as.character(paste(MT_locationids$Location))
ND_locationids <- read.csv("./Input/ND KC Location IDs.csv")
ND_locationids$Location <- as.character(paste(ND_locationids$Location))
SD_locationids <- read.csv("./Input/SD KC Location IDs.csv")
SD_locationids$Location <- as.character(paste(SD_locationids$Location))
```

## STEP 1: IMPORT THE DATA AND CHECK THAT ALL IS IMPORTED CORRECTLY
```{r}
df_food <- read.csv("./Input/economics/MMG_2020Projections_ALLDATA.csv")
df_food_state <- read.csv("./Input/economics/MMG_2020Projections_ALLDATA_state.csv")
```


## STEP 2: CLEAN THE IMPORTED DATA FOR COUNTY LEVEL AND STATE LEVEL
```{r}
############
#COUNTY DATA
#subset to only include Montana, North Dakota, and South Dakota and clean the data

county_food <- df_food %>%
  subset(State=='MT' | State=='ND' | State=='SD') %>%
  rename(state='State',
         county='County..State',
         childfi_rate_2020='X.C..2020.Child.Food.Insecurity..',
         childfi_rate_2018='X2018.Child.Food.Insecurity..') %>%
  mutate(county=sub("\\ County.*","",county)) %>% 
  select(c(state,county,childfi_rate_2020,childfi_rate_2018)) %>%
  mutate(state=case_when(
    state=='MT' ~ 'Montana',
    state=='ND' ~ 'North Dakota',
    state=='SD' ~ 'South Dakota')) %>%
  rename(location=county) %>%
  mutate(locationtype='County') %>%
  pivot_longer(cols=c(childfi_rate_2018,childfi_rate_2020),names_to='timeframe',values_to='data') %>%
  mutate(timeframe=case_when(
    timeframe=='childfi_rate_2020' ~ '2020',
    timeframe=='childfi_rate_2018' ~ '2018')) %>%
  
  #reformat percentage
  mutate(data=as.numeric(gsub("[\\%,]", "",data))) %>%
  mutate(data=data/100) %>%
  
  mutate(dataformat='Percent') %>%
  mutate(varname='childfoodinsecuritymmg_specialprojection2020')

#merge in location IDs with data

#MONTANA COUNTY DATA  
mt_county_food <- county_food %>%
  subset(state=='Montana') %>%
  mutate(location=replace(location, 
                          state=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  left_join(MT_locationids, by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#NORTH DAKOTA COUNTY DATA
nd_county_food <- county_food %>%
  subset(state=='North Dakota') %>%
  left_join(ND_locationids, by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#SOUTH DAKOTA COUNTY DATA
sd_county_food <- county_food %>%
  subset(state=='South Dakota') %>%
  left_join(SD_locationids, by=c('location'='Location')) %>%
  rename(locationid=LocationId)



###########
#STATE DATA
#subset to only include Montana, North Dakota, and South Dakota and clean the data

state_food <- df_food_state %>%
  subset(State=='Montana' | State=='North Dakota' | State=='South Dakota') %>%
  rename(state='State',
         childfi_rate_2020=X2020.Projected,
         childfi_rate_2018=X2018) %>%
  select(c(state,childfi_rate_2020,childfi_rate_2018)) %>%
   mutate(location=case_when(
    state=='Montana' ~ 'Montana',
    state=='North Dakota' ~ 'North Dakota',
    state=='South Dakota' ~ 'South Dakota')) %>%
  mutate(locationtype='State') %>%
  
  pivot_longer(cols=c(childfi_rate_2018,childfi_rate_2020),names_to='timeframe',values_to='data') %>%
  mutate(timeframe=case_when(
    timeframe=='childfi_rate_2020' ~ '2020',
    timeframe=='childfi_rate_2018' ~ '2018')) %>%
  
  #reformat percentage
  mutate(data=as.numeric(gsub("[\\%,]", "",data))) %>%
  mutate(data=data/100) %>%
  
  mutate(dataformat='Percent') %>%
  mutate(varname='childfoodinsecuritymmg_specialprojection2020')

#merge in location IDs with data

#MONTANA COUNTY DATA  
mt_state_food <- state_food %>%
  subset(state=='Montana') %>%
  mutate(location=replace(location, 
                          state=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  left_join(MT_locationids, by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#NORTH DAKOTA COUNTY DATA
nd_state_food <- state_food %>%
  subset(state=='North Dakota') %>%
  left_join(ND_locationids, by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#SOUTH DAKOTA COUNTY DATA
sd_state_food <- state_food %>%
  subset(state=='South Dakota') %>%
  left_join(SD_locationids, by=c('location'='Location')) %>%
  rename(locationid=LocationId)



#######################################################
#COMBINE COUNTY AND STATE DATA, SEPARATE FOR EACH STATE
mt_data <- mt_county_food %>%
  bind_rows(mt_state_food)

nd_data <- nd_county_food %>%
  bind_rows(nd_state_food)

sd_data <- sd_county_food %>%
  bind_rows(sd_state_food)

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

#MONTANA
# 1. Print name of location that has a mismatched location ID
if (sum(is.na(mt_data$locationid))>=1) {
  print(mt_data$location[is.na(mt_data$locationid)])
} else if (sum(is.na(mt_data$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- mt_data %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
View(mt_data)

#NORTH DAKOTA
# 1. Print name of location that has a mismatched location ID
if (sum(is.na(nd_data$locationid))>=1) {
  print(nd_data$location[is.na(nd_data$locationid)])
} else if (sum(is.na(nd_data$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- nd_data %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
View(nd_data)

#SOUTH DAKOTA
# 1. Print name of location that has a mismatched location ID
if (sum(is.na(sd_data$locationid))>=1) {
  print(sd_data$location[is.na(sd_data$locationid)])
} else if (sum(is.na(sd_data$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- sd_data %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
View(sd_data)
```

## STEP 2: COMMIT TO DATABASE
```{r}
#CHECK DATASETS NAMED mt_data, nd_data, AND sd_data TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',mt_data,append=TRUE,row.names=FALSE)
dbWriteTable(con,'northdakota',nd_data,append=TRUE,row.names=FALSE)
dbWriteTable(con,'southdakota',sd_data,append=TRUE,row.names=FALSE)
```

## STEP 3: OUTPUT DATA FOR UPLOADING TO DATA CENTER 
**Child food insecurity by county**
```{r}
#write query from database to get needed format for KC data center for food insecurity only

mt_food_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE varname='childfoodinsecuritymmg_specialprojection2020';")

mt_upload_data_food <- dbGetQuery(con,mt_food_sql)

mt_upload_data_food2 <- mt_upload_data_food %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

nd_food_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM northdakota WHERE varname='childfoodinsecuritymmg_specialprojection2020';")

nd_upload_data_food <- dbGetQuery(con,nd_food_sql)

nd_upload_data_food2 <- nd_upload_data_food %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

sd_food_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE varname='childfoodinsecuritymmg_specialprojection2020';")

sd_upload_data_food <- dbGetQuery(con,sd_food_sql)

sd_upload_data_food2 <- sd_upload_data_food %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(mt_upload_data_food2,file=paste0("./Output/economics/montana_childfoodinsecuritymmg_specialprojection2020.csv"),row.names=FALSE)

write.csv(nd_upload_data_food2,file=paste0("./Output/economics/northdakota_childfoodinsecuritymmg_specialprojection2020.csv"),row.names=FALSE)

write.csv(sd_upload_data_food2,file=paste0("./Output/economics/southdakota_childfoodinsecuritymmg_specialprojection2020.csv"),row.names=FALSE)
```

