---
title: "montanawic"
output: html_document
---

## Indicator 1: Participants in the supplemental nutrition program for women, infants, and children (WIC)


**Created by:** Xanna Burg
**Date:** September 2020
**Updated by:**

**Data Source:** Montana Department of Public Health and Human Services, Early Childhood and Family Services Division
**Purpose:** Input the WIC data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)

**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year' and 'fullyear'
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '2019'

statefile <- 'montana'
statename <- 'Montana'

#input location ID file for MT
locationids <- read.csv("./Input/MT KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)

#input region ID file for MT
regionids <- read.csv("./Input/MT KC Region List.csv")
regionids$county <- as.character(regionids$county)
```


## IMPORT THE DATA FILES THAT WILL BE USED
```{r}
wic_countydata <- read_excel(path=paste0("./Input/economics/montana_",year,"_wicparticipants.xlsx"),sheet='By County') 

wic_statedata <- read_excel(path=paste0("./Input/economics/montana_",year,"_wicparticipants.xlsx"),sheet='Statewide') 
```


## CLEAN THE IMPORTED DATA
```{r}
#COUNTY DATA
wic_county <- wic_countydata %>% 
  
  rename(timeframe=FFY,
         data=`Total Participants`) %>%
  
  #fix the formatting of the county names
  #lowercase county names
  mutate(County=tolower(County)) %>%
  mutate(location=str_to_title(County)) %>%
  select(-c(County)) %>%
  
  subset(location != 'Null') %>%

  mutate(locationtype='County') %>%
  
  #fix the name on some counties
  mutate(location=replace(location,location=='Lewis And Clark','Lewis & Clark')) %>%
  mutate(location=replace(location,location=='Mccone','McCone')) %>%
  
  #subset if it is not a Montana county
  subset(location!='Mckenzie')
  

#STATE DATA
wic_state <- wic_statedata %>% 
  
  rename(timeframe=FFY,
         data=`Total Participants`) %>%
  
  mutate(location='Montana') %>%
  mutate(locationtype='State') 
  

#COMBINE GEOGRAPHIES
wic_all <- wic_county %>%
  bind_rows(wic_state) %>%
  
  mutate(state='Montana',
         dataformat='Number',
         varname='wicparticipants') %>%
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(wic_all$locationid))>=1) {
  print(wic_all$location[is.na(wic_all$locationid)])
} else if (sum(is.na(wic_all$locationid))==0) {
  'all locations match'
}

# 4. Visually inspect output data
View(wic_all)
```

## STEP 2: ADD TO DATABASE
```{r}
#CHECK DATASET NAMED wic_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',wic_all,append=TRUE,row.names=FALSE)

```

## STEP 3: OUTPUT FILE FOR KC
```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",year,"' AND varname='wicparticipants';")

upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/economics/",statefile,"_",year,"_wicparticipants.csv"),row.names=FALSE)
```


