---
title: "montanawic"
output: html_document
---

## Indicator 1: Children and families receiving evidence-based home visiting services


**Created by:** Xanna Burg
**Date:** October 2020
**Updated by:**

**Data Source:** Montana Department of Public Health and Human Services, Early Childhood and Family Services Division
**Purpose:** Input the home visiting data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)

**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year' and 'fullyear'
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '2025'


statefile <- 'montana'
statename <- 'Montana'

#input location ID file for MT
locationids <- read.csv("./Input/MT KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)

#input region ID file for MT
regionids <- read.csv("./Input/MT KC Region List.csv")
regionids$county <- as.character(regionids$county)
```


## IMPORT THE DATA FILES THAT WILL BE USED
```{r}
homevisitingdata <- read_excel(path=paste0("./Input/health/montana_",year,"_HMF Families Served by County.xlsx"),sheet='Sheet1',skip=2) 
```


## CLEAN THE IMPORTED DATA
```{r}
homevisitingdata2 <- homevisitingdata %>% 
  
  rename(location=COUNTY,
        families_2025=`FAMILIES SERVED...6`,
         children_2025=`CHILDREN SERVED...7`) %>%
  
  #remove footnotes
  subset(location!='Notes:') %>%

  #select only most recent year
  select(c(location,paste0("families_",year),paste0("children_",year))) %>%
  
  #update locations
  mutate(location=replace(location,location=='TOTAL','Montana')) %>%
  mutate(locationtype=ifelse(location=='Montana','State','County')) %>%

  mutate(state='Montana',
         dataformat='Number',
         varname='childrenandfamilieshomevisiting') %>%
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  

  #pivot longer
  pivot_longer(cols=c(paste0("families_",year),paste0("children_",year)),names_to='category',values_to='data') %>%
  
  #split category to include the date
  separate(category,into=c('category','timeframe'),sep=-4) %>%
  mutate(category=case_when(
    category=='families_' ~ 'Families',
    category=='children_' ~ 'Children')) %>%
  
  #delineate NA and S
  mutate(data=case_when(
    data=='*' ~ 'S',
    is.na(data) ~'NA',
    TRUE ~ data))
  



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(homevisitingdata2$locationid))>=1) {
  print(homevisitingdata2$location[is.na(homevisitingdata2$locationid)])
} else if (sum(is.na(homevisitingdata2$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(homevisitingdata2)
```

## STEP 2: ADD TO DATABASE
```{r}
#CHECK DATASET NAMED homevisitingdata2 TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',homevisitingdata2,append=TRUE,row.names=FALSE)

```

## STEP 3: OUTPUT FILE FOR KC
```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, category, data FROM montana WHERE timeframe='",year,"' AND varname='childrenandfamilieshomevisiting';")

upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Category=category)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/health/",statefile,"_",year,"_childrenandfamilieshomevisiting.csv"),row.names=FALSE)
```


