---
title: "Montana Juvenile Offenses"
author: "Xanna Burg"
date: "6/5/2020"
output: html_document
---

## Indicator 1: Youth referred to Youth Court
## Indicator 2: Referrals to Youth Court
## Indicator 3: Juvenile offenses  

**Created by:** Xanna Burg
**Date:** June 2020
**Updated by:**

**Data Source:** Montana Judicial Branch, Youth Court Services
**Purpose:** Input the youth court data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)

**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year' and 'fullyear'
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '21'
fullyear <- '2021' 


#input location ID file
locationids <- read.csv("./Input/MT KC Location IDs.csv")
locationids$Location <- as.character(paste(locationids$Location))

#input county reference file
MT_counties <- read.csv("./Input/MT County List.csv")
MT_counties$Location <- as.character(paste(MT_counties$Location))

```


## ############################# ##
## YOUTH REFERRED TO YOUTH COURT ##
## REFERRALS TO YOUTH COURT      ##
## ############################# ##

## STEP 1: CLEAN THE COUNTS OF REFERRALS AND YOUTH
```{r}
#import the referral and unduplicated youth data file
youthdata <- read_excel(path=paste0("./Input/safety/montana_",fullyear,"_juvenilereferralsandyouth.xlsx"),sheet=fullyear,skip=3)

############
#COUNTY DATA
youthdata_county <- youthdata %>%
  subset(County!="Unduplicated Statewide") %>%
  
  #merge in all county names
  mutate(County=replace(County, 
                          County=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
    mutate(County=replace(County, 
                          County=='Silverbow', 
                          'Silver Bow')) %>%
  full_join(MT_counties,by=c('County'='Location')) %>%
  
  rename(location=County,
         undup_youth='Undup Youth') %>%
  mutate(locationtype='County') %>%
  mutate(dataformat='Number') %>%
  mutate(state='Montana') %>%
  mutate(timeframe=fullyear) %>%
  
  #pivot wide to long
  pivot_longer(cols=c('Referrals','undup_youth'), names_to='varname',values_to='data') %>%
  mutate(varname=case_when(
    varname=='Referrals' ~ 'referralstojuvenilecourt',
    varname=='undup_youth' ~ 'childrenreferredtojuvenilecourt'))

############
#STATE DATA
youthdata_state <- youthdata %>%
  subset(County=="Unduplicated Statewide") %>%
  
  mutate(location='Montana') %>%
  select(-c(County)) %>%
  
  rename(undup_youth='Undup Youth') %>%
  mutate(locationtype='State') %>%
  mutate(dataformat='Number') %>%
  mutate(state='Montana') %>%
  mutate(timeframe=fullyear) %>%
  
  #pivot wide to long
  pivot_longer(cols=c('Referrals','undup_youth'), names_to='varname',values_to='data') %>%
  mutate(varname=case_when(
    varname=='Referrals' ~ 'referralstojuvenilecourt',
    varname=='undup_youth' ~ 'childrenreferredtojuvenilecourt'))
  

#COMBINE COUNTY AND STATE
youthdata_all <- youthdata_county %>%
  bind_rows(youthdata_state) %>%
  
  mutate(suppress=case_when(
    data<5 ~ 1,
    data>=5 ~ 0)) %>%
  
  #suppress if NA or less than 5
  mutate(data=as.character(paste(data))) %>%
  mutate(data=case_when(
    suppress==1 ~ '<5',
     data=='NA' ~ '0',
     suppress==0 ~ data)) %>%
  select(-c(suppress)) %>%
  
  #merge in location ids
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

####################
###################
#DATA QUALITY CHECKS
# 1. Print name of location that has a mismatched location ID
if (sum(is.na(youthdata_all$locationid))>=1) {
  print(youthdata_all$location[is.na(youthdata_all$locationid)])
} else if (sum(is.na(youthdata_all$locationid))==0) {
  'all locations match'
}


# 3. Visually inspect output data
View(youthdata_all)

```

## STEP 2: COMMIT COUNTS TO DATABASE
```{r}
#CHECK DATASETS NAMED youthdata_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',youthdata_all,append=TRUE,row.names=FALSE)
```

## STEP 3: OUTPUT DATASETS FOR UPLOADING TO DATA CENTER
```{r}
#write query from database to get needed format for KC data center 

#REFERRALS
referrals_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",fullyear,"' AND varname='referralstojuvenilecourt';")

referrals_upload <- dbGetQuery(con,referrals_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(referrals_upload,file=paste0("./Output/safety/montana_",fullyear,"_referralstojuvenilecourt.csv"),row.names=FALSE)

#YOUTH
youth_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",fullyear,"' AND varname='childrenreferredtojuvenilecourt';")

youth_upload <- dbGetQuery(con,youth_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(youth_upload,file=paste0("./Output/safety/montana_",fullyear,"_childrenreferredtojuvenilecourt.csv"),row.names=FALSE)
```


## ################# ##
## JUVENILE OFFENSES ##
## ################# ##

## STEP 1: IMPORT AND CLEAN DATA
```{r}
#import the data
offensedata <- read_excel(path=paste0("./Input/safety/montana_",fullyear,"_juvenileoffenses.xlsx"),sheet=fullyear,skip=1)

############
#COUNTY DATA
offenses_county <- offensedata %>%
  subset(County!='NULL'& !is.na(County)) %>%
  
  #sum by county
  group_by(County) %>%
  summarise(data=sum(Count)) %>%
  ungroup %>%
  
  rename(location=County) %>%
  
  #merge in all county names
  mutate(location=replace(location, 
                          location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
    mutate(location=replace(location, 
                          location=='Silverbow', 
                          'Silver Bow')) %>%
    mutate(location=replace(location, 
                          location=='MINERAL', 
                          'Mineral')) %>%
  full_join(MT_counties,by=c('location'='Location')) %>%
    mutate(locationtype='County') 


###########
#STATE DATA
offenses_state <- offensedata %>%
  subset(County!='NULL'& !is.na(County)) %>%
  
  mutate(location='Montana') %>%
  
  #sums for state
  group_by(location) %>%
  summarise(data=sum(Count)) %>%
  ungroup %>%
  
  mutate(locationtype='State') 


##############################
#COMBINE COUNTY AND STATE DATA
offenses_all <- offenses_county %>%
  bind_rows(offenses_state) %>%
  
  mutate(timeframe=fullyear) %>%
  mutate(state='Montana') %>%
  mutate(varname='juvenileoffenses') %>%
  
  #merge in location ids
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  
  #suppress if less than 5 and change NA to zero
  mutate(suppress=case_when(
    data<5 ~ 1,
    data>=5 ~ 0)) %>%
  mutate(data=as.character(paste(data))) %>%
  mutate(data=case_when(
    suppress==1 ~ '<5',
    suppress==0 ~ data,
    data=='NA' ~ '0')) %>%
  select(-c(suppress)) %>%
  
  mutate(dataformat='Number')


####################
###################
#DATA QUALITY CHECKS
# 1. Print name of location that has a mismatched location ID
if (sum(is.na(offenses_all$locationid))>=1) {
  print(offenses_all$location[is.na(offenses_all$locationid)])
} else if (sum(is.na(offenses_all$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(offenses_all)
```

## STEP 2: COMMIT COUNTS TO DATABASE
```{r}
#CHECK DATASETS NAMED offenses_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',offenses_all,append=TRUE,row.names=FALSE)
```

## STEP 3: OUTPUT DATASETS FOR UPLOADING TO DATA CENTER
```{r}
#write query from database to get needed format for KC data center 

offenses_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",fullyear,"' AND varname='juvenileoffenses';")

offenses_upload <- dbGetQuery(con,offenses_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(offenses_upload,file=paste0("./Output/safety/montana_",fullyear,"_juvenileoffenses.csv"),row.names=FALSE)
```

