---
title: "Montana Youth Risk Behavior Survey"
author: "Xanna Burg"
date: "2/24/2020"
output: html_document
---

## Indicator 1: Cigarette smoking among high school students
## Indicator 2: Smokeless tobacco use among high school students
## Indicator 3: Alcohol use among high school students
## Indicator 4: Marijuana use among high school students
## Indicator 5: Suicide attempts among high school students
## Indicator 6: High school students feeling sad or hopeless
## Indicator 7: High school students who drink and drive
## Indicator 8: High school students not wearing seat belts
## Indicator 9: High school students bullied at school
## Indicator 10: Electronic vapor product use among high school students

**Created by:** Xanna Burg
**Date:** February, 2020
**Updated by:**

**Data Source:** MT Office of Public Instruction, Montana Youth Risk Behavior Survey
**Purpose:** Input the statewide YRBS data by race and ethnicity
**Description:** this program calculates eight indicators 


* Geographies: state total
* Percent - The percent of youth reporting behavior

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)

**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year'
* Update the year in the data query from the database for each of the five indicators, prior to exporting to .csv for data center upload. 


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)

```

```{r}
#UPDATE to reflect the current year data working with
year <- '2023'
statefile <- 'montana'
statename <- 'Montana'
```


```{r}

xl_data <- "./Input/safety/MT YRBS/montana_allyears_yrbs.xlsx"

#input yrbs data
df_allstudents <- read_excel(path=xl_data, sheet='All Montana Students')
df_AIreservation <- read_excel(path=xl_data, sheet='AI Reservation')
df_AItotal <- read_excel(path=xl_data, sheet='AI Total')
df_white <- read_excel(path=xl_data, sheet='White Total')
df_hispanic <- read_excel(path=xl_data, sheet='Hispanic total')
df_multiple <- read_excel(path=xl_data, sheet='Multiple race total')
df_black <- read_excel(path=xl_data, sheet='Black total')


#input location ID file for MT
locationids <- read.csv("./Input/MT KC Location IDs.csv")

```


##Before running the current year, if any information was added from the prior year for Multiple Race, that needs to be updated too##
```{r}
year_prior <- as.character(as.numeric(year)-2)



#delete the old rows of data
dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_sadhopelessamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_suicideattemptsamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_schoolbullyingamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_cigarettesmokingamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_smokelesstobaccoamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_alcoholuseamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_marijuanauseamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_electronicvaporuseamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_drinkdriveamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))

dbGetQuery(con,paste0("DELETE FROM montana WHERE (varname='yrbs_noseatbeltamonghighschoolstudents' AND race='Multiple Race students' AND timeframe='",year_prior,"');"))


#get the replacement data from the excel file
cigarette_mr <-  df_multiple %>%
  select(c(locationid,location,race,year,dataformat,cigarette)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=cigarette) %>%
  mutate(varname='yrbs_cigarettesmokingamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))

smokeless_mr <-  df_multiple %>%
  select(c(locationid,location,race,year,dataformat,smokeless)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=smokeless) %>%
  mutate(varname='yrbs_smokelesstobaccoamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  alcohol_mr <- df_multiple %>%
    select(c(locationid,location,race,year,dataformat,alcohol)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=alcohol) %>%
  mutate(varname='yrbs_alcoholuseamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  marijuana_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,marijuana)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=marijuana) %>%
  mutate(varname='yrbs_marijuanauseamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  suicide_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,suicide)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=suicide) %>%
  mutate(varname='yrbs_suicideattemptsamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  noseatbelt_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,noseatbelt)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=noseatbelt) %>%
  mutate(varname='yrbs_noseatbeltamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  drinkdrive_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,drinkdrive)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=drinkdrive) %>%
  mutate(varname='yrbs_drinkdriveamonghighschoolstudents') %>% 
  subset(timeframe==as.numeric(year_prior))
  
  hopeless_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,hopeless)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=hopeless) %>%
  mutate(varname='yrbs_sadhopelessamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  bullied_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,bullying)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=bullying) %>%
  mutate(varname='yrbs_schoolbullyingamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))
  
  ecigarette_mr <- df_multiple %>%
  select(c(locationid,location,race,year,dataformat,ecigarette)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=ecigarette) %>%
  mutate(varname='yrbs_electronicvaporuseamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year_prior))

```

```{r}
#add to database 
dbWriteTable(con,'montana',cigarette_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',smokeless_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',alcohol_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',marijuana_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',suicide_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',noseatbelt_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',drinkdrive_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',hopeless_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',bullied_mr,append=TRUE,row.names=FALSE)
dbWriteTable(con,'montana',ecigarette_mr,append=TRUE,row.names=FALSE)

```

#the updated data, if there is any can be pulled as an entire new dataset with all years from the output code.


**CIGARETTE USE**

```{r}
#format the data to add to the database

cigarette <-  df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_AIreservation) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,cigarette)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=cigarette) %>%
  mutate(varname='yrbs_cigarettesmokingamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(cigarette)
```

```{r}
#add to database percent of yrbs_cigarettesmokingamonghighschoolstudents
dbWriteTable(con,'montana',cigarette,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_CIGARETTESMOKINGAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_cigarette <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE timeframe = '",year,"' AND varname = 'yrbs_cigarettesmokingamonghighschoolstudents';"))

#all years
upload_data_cigarette <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_cigarettesmokingamonghighschoolstudents';"))

upload_data_cigarette2 <- upload_data_cigarette %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_cigarette2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_cigarettesmokingamonghighschoolstudents.csv"),row.names=FALSE)

```


**SMOKELESS TOBACCO**

```{r}
#format the data to add to the database
smokeless <-  df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_AIreservation) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,smokeless)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=smokeless) %>%
  mutate(varname='yrbs_smokelesstobaccoamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(smokeless)
```

```{r}
#add to database percent of yrbs_smokelesstobaccoamonghighschoolstudents
dbWriteTable(con,'montana',smokeless,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_SMOKELESSTOBACCOAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_smokeless<- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_smokelesstobaccoamonghighschoolstudents';"))

#all years
upload_data_smokeless<- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_smokelesstobaccoamonghighschoolstudents';"))

upload_data_smokeless2 <- upload_data_smokeless %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_smokeless2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_smokelesstobaccoamonghighschoolstudents.csv"),row.names=FALSE)

```



**ALCOHOL**

```{r}
#format the data to add to the database
alcohol <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_AIreservation) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,alcohol)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=alcohol) %>%
  mutate(varname='yrbs_alcoholuseamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(alcohol)
```

```{r}
#add to database percent of yrbs_alcoholuseamonghighschoolstudents
dbWriteTable(con,'montana',alcohol,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_ALCOHOLUSEAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_alcohol<- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_alcoholuseamonghighschoolstudents';"))

#all years
upload_data_alcohol<- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_alcoholuseamonghighschoolstudents';"))

upload_data_alcohol2 <- upload_data_alcohol %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_alcohol2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_alcoholuseamonghighschoolstudents.csv"),row.names=FALSE)

```




**MARIJUANA**

```{r}
#format the data to add to the database
marijuana <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_AIreservation) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,marijuana)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=marijuana) %>%
  mutate(varname='yrbs_marijuanauseamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(marijuana)
```

```{r}
#add to database percent of yrbs_marijuanauseamonghighschoolstudents
dbWriteTable(con,'montana',marijuana,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_MARIJUANAUSEAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_marijuana<- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_marijuanauseamonghighschoolstudents';"))

#all years
upload_data_marijuana<- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_marijuanauseamonghighschoolstudents';"))

upload_data_marijuana2 <- upload_data_marijuana %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_marijuana2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_marijuanauseamonghighschoolstudents.csv"),row.names=FALSE)

```


**SUICIDE ATTEMPTS**

```{r}
#format the data to add to the database
suicide <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_AIreservation) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,suicide)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=suicide) %>%
  mutate(varname='yrbs_suicideattemptsamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(suicide)
```

```{r}
#add to database percent of yrbs_suicideattemptsamonghighschoolstudents
dbWriteTable(con,'montana',suicide,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_SUICIDEATTEMPTSAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_suicide <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_suicideattemptsamonghighschoolstudents';"))

#all years
upload_data_suicide <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_suicideattemptsamonghighschoolstudents';"))

upload_data_suicide2 <- upload_data_suicide %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_suicide2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_suicideattemptsamonghighschoolstudents.csv"),row.names=FALSE)

```



**RIDER NOT WEARING SEATBELT**

```{r}
#format the data to add to the database
noseatbelt <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,noseatbelt)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=noseatbelt) %>%
  mutate(varname='yrbs_noseatbeltamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(noseatbelt)
```

```{r}
#add to database percent of yrbs_noseatbeltamonghighschoolstudents
dbWriteTable(con,'montana',noseatbelt,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_NOSEATBELTAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_noseatbelt <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND  varname = 'yrbs_noseatbeltamonghighschoolstudents';"))

#all years
upload_data_noseatbelt <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_noseatbeltamonghighschoolstudents';"))

upload_data_noseatbelt2 <- upload_data_noseatbelt %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_noseatbelt2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_noseatbeltamonghighschoolstudents.csv"),row.names=FALSE)


```



**DRINK AND DRIVE**

```{r}
#format the data to add to the database
drinkdrive <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,drinkdrive)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=drinkdrive) %>%
  mutate(varname='yrbs_drinkdriveamonghighschoolstudents') %>% 
  subset(timeframe==as.numeric(year))

View(drinkdrive)
```

```{r}
#add to database percent of yrbs_drinkdriveamonghighschoolstudents
dbWriteTable(con,'montana',drinkdrive,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - YRBS_DRINKDRIVEAMONGHIGHSCHOOLSTUDENTS

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_drinkdrive <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_drinkdriveamonghighschoolstudents';"))

#all years
upload_data_drinkdrive <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_drinkdriveamonghighschoolstudents';"))

upload_data_drinkdrive2 <- upload_data_drinkdrive %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_drinkdrive2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_drinkdriveamonghighschoolstudents.csv"),row.names=FALSE)


```


**FEELING SAD OR HOPELESS**

```{r}
#format the data to add to the database
hopeless <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,hopeless)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=hopeless) %>%
  mutate(varname='yrbs_sadhopelessamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(hopeless)
```

```{r}
#add to database percent of yrbs_sadhopelessamonghighschoolstudents
dbWriteTable(con,'montana',hopeless,append=TRUE,row.names=FALSE)

```
```{r}
#write query from database to get needed format for KC data center
#PERCENT - yrbs_sadhopelessamonghighschoolstudents

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_sadhopeless <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_sadhopelessamonghighschoolstudents';"))

#all years
upload_data_sadhopeless <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_sadhopelessamonghighschoolstudents';"))

upload_data_sadhopeless2 <- upload_data_sadhopeless %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_sadhopeless2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_sadhopelessamonghighschoolstudents.csv"),row.names=FALSE)


```



**BULLIED AT SCHOOL**

```{r}
#format the data to add to the database
bullied <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,bullying)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=bullying) %>%
  mutate(varname='yrbs_schoolbullyingamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(bullied)
```

```{r}
#add to database percent of yrbs_schoolbullyingamonghighschoolstudents
dbWriteTable(con,'montana',bullied,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - yrbs_schoolbullyingamonghighschoolstudents

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_bullied <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_schoolbullyingamonghighschoolstudents';"))

#all years
upload_data_bullied <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_schoolbullyingamonghighschoolstudents';"))

upload_data_bullied2 <- upload_data_bullied %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_bullied2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_schoolbullyingamonghighschoolstudents.csv"),row.names=FALSE)


```



**ELECTRONIC VAPOR USE**

```{r}
#format the data to add to the database
ecigarette <- df_allstudents %>%
  bind_rows(df_AItotal) %>%
  bind_rows(df_white) %>%
  bind_rows(df_hispanic) %>%
  bind_rows(df_multiple) %>%
  bind_rows(df_black) %>%
  select(c(locationid,location,race,year,dataformat,ecigarette)) %>%
  mutate(locationtype='State') %>%
  mutate(state='Montana') %>%
  rename(timeframe=year) %>%
  rename(data=ecigarette) %>%
  mutate(varname='yrbs_electronicvaporuseamonghighschoolstudents') %>%
  subset(timeframe==as.numeric(year))

View(ecigarette)
```

```{r}
#add to database percent of yrbs_electronicvaporuseamonghighschoolstudents
dbWriteTable(con,'montana',ecigarette,append=TRUE,row.names=FALSE)

```

```{r}
#write query from database to get needed format for KC data center
#PERCENT - yrbs_electronicvaporuseamonghighschoolstudents

#UPDATE THE YEAR IN THE SQL QUERY

upload_data_ecigarette <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race
                     FROM montana
                     WHERE timeframe = '",year,"' AND varname = 'yrbs_electronicvaporuseamonghighschoolstudents';"))

#all years
upload_data_ecigarette <- dbGetQuery(con,paste0("SELECT locationid,location,timeframe,dataformat,data,race FROM montana WHERE varname = 'yrbs_electronicvaporuseamonghighschoolstudents';"))

upload_data_ecigarette2 <- upload_data_ecigarette %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_ecigarette2,file=paste0("./Output/safety/",statefile,"_",year,"_yrbs_electronicvaporuseamonghighschoolstudents.csv"),row.names=FALSE)


```
