---
title: "MT Motor Vehicle"
author: "Xanna Burg"
date: "March 2020"
output: html_document
---

## Indicator 1: Motor vehicle crashes with driver under age 18

**Created by:** Xanna Burg
**Date:** March 2020
**Updated by:**

**Data Source:** Montana Department of Transportation
**Purpose:** Input the motor vehicle data.

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)

**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year' and 'fullyear'
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '19'

fullyear <- '2019' 
statefile <- 'montana'
statename <- 'Montana'

#input file for region IDs
regionids <- read.csv("./Input/MT KC Region List.csv")
regionids$county <- as.character(regionids$county)

#input location ID file for MT
locationids <- read.csv("./Input/MT KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)
```

```{r}
#read in the data
crashes <- read.csv(paste0('./Input/safety/montana_',year,'_motorvehicle.csv'))
```

```{r}
############
#COUNTY DATA
crashes_county <- crashes %>%
  rename(location=County,
         data=paste0("X",fullyear)) %>%
  mutate(location=as.character(paste(location))) %>%
  mutate(locationtype='County')
  

###########################
#STATE PLANNING REGION DATA
crashes_region <- crashes_county %>%

  #add in regions
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location,statename=='Montana' & location=='Sweetgrass','Sweet Grass')) %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region) %>%
  summarise(data=sum(data)) %>%
  rename(location=region) %>%
  mutate(location=as.character(paste(location))) %>%
  mutate(locationtype='Planning Region') 
  

###########
#STATE DATA
crashes_state <- crashes_county %>%
  
  mutate(location='Montana') %>%
  group_by(location) %>%
  summarise(data=sum(data)) %>%
  mutate(locationtype='State') 
 

##################################################
#COMBINE COUNTY, STATE, AND STATE PLANNING REGIONS
crashesdata <- crashes_county %>%
  bind_rows(crashes_region) %>%
  bind_rows(crashes_state) %>%
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location,statename=='Montana' & location=='Sweetgrass','Sweet Grass')) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  
  mutate(timeframe=fullyear) %>%
  mutate(state='Montana') %>%
  mutate(dataformat='Number') %>%
  mutate(varname='motorvehiclecrashesunder18') 



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(crashesdata$locationid))>=1) {
  print(crashesdata$location[is.na(crashesdata$locationid)])
} else if (sum(is.na(crashesdata$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(crashesdata)

```


```{r}
#CHECK DATASET NAMED crashesdata TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',crashesdata,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

crashes_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",fullyear,"' AND varname='motorvehiclecrashesunder18';")

upload_data_crashes <- dbGetQuery(con,crashes_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_crashes,file=paste0("./Output/safety/",statefile,"_",year,"_motorvehiclecrashesunder18.csv"),row.names=FALSE)
```



