---
title: "Montana Foster Care Data"
output: html_document
---

## Indicator 1: Children ages 0 to 17 in foster care by county
## Indicator 2: Children ages 0 to 5 in foster care by county
## Indicator 3: Average days a child is in state care before exiting


**Created by:** Xanna Burg
**Date:** November 2020
**Updated by:**

**Data Source:** Montana Department of Public Health and Human Services, Child and Family Services Division
**Purpose:** Input the foster care data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)

**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year' and 'fullyear'
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}


#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '2022'
year_short <- '22'

statefile <- 'montana'
statename <- 'Montana'

#input location ID file for MT
locationids <- read.csv("./Input/MT KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)

#input region ID file for MT
regionids <- read.csv("./Input/MT KC Region List.csv")
regionids$county <- as.character(regionids$county)
```


## IMPORT THE DATA FILES THAT WILL BE USED
```{r}
#fostercaredata_0to17 <- read_excel(path=paste0("./Input/safety/montana_",year_short,"_fostercaredata.xlsx"),sheet='Ages 0-17',skip=3) 

#fostercaredata_0to5 <- read_excel(path=paste0("./Input/safety/montana_",year_short,"_fostercaredata.xlsx"),sheet='Ages 0-5',skip=3) 

fostercaredata <- read_excel(path=paste0("./Input/safety/montana_",year_short,"_fostercaredata.xlsx"),sheet='2022 Unduplicated Kids in Care') 
```


## -- FOR CHILDREN AGES 0 TO 17 -- ##

## CLEAN THE IMPORTED DATA
```{r}
fostercaredata_0to17 <- fostercaredata %>% 
  
  rename(location=`...1`) %>%
  #select(c(location,all_of(paste0("CY ",year)))) %>%
  #rename(data=all_of(paste0("CY ",year))) %>%
  
  #new format in 2021
  rename(data=`Unduplicated kids in care 0-17`) %>%
  select(location,data) %>%
  subset(!is.na(location)) %>%
  
  
  
  #update locations
  mutate(location=replace(location,location=='State (including tribal)','Montana')) %>%
  filter(!location %in% c("Blackfeet Tribal SS","Crow Tribal SS","Flathead Tribal SS","Fort Belknap Tribal SS","Northern Cheyenne Tribal SS","Rocky Boy Tribal SS")) %>%
  mutate(locationtype=ifelse(location=='Montana','State','County')) %>%
  mutate(location=replace(location,location=='Choteau','Chouteau')) %>%
  mutate(location=replace(location,location=='Lewis and Clark','Lewis & Clark')) %>%
  mutate(location=replace(location,location=='Power River','Powder River')) %>%
  mutate(location=replace(location,location=='Sweetgrass','Sweet Grass')) %>%

  mutate(state='Montana',
         dataformat='Number',
         varname='childrenages0to17infostercare') %>%
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=year) %>%  
  
  
  #suppress if less than 4
  mutate(data=replace(data,data<4,NA))
  



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(fostercaredata_0to17$locationid))>=1) {
  print(fostercaredata_0to17$location[is.na(fostercaredata_0to17$locationid)])
} else if (sum(is.na(fostercaredata_0to17$locationid))==0) {
  'all locations match'
}


#2. Check that no values are less than 4
temp_suppresscheck <- fostercaredata_0to17 %>% subset(data<4)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- fostercaredata_0to17 %>%
  subset(locationtype=='County') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
} else if (temp_rows==1) {
  View(temp_testsuppression)
}

#if there is only one county, print the counties in ascending order
#temp_check <- fostercaredata_0to17 %>%
#  subset(locationtype=='County') %>%
#  arrange(data)
#View(temp_check)

#******
#CHANGE TO THE CORRECT COUNTY NAME HERE
#**Choose the county with the next lowest count of births; if there is a tie, suppress both
#fostercaredata_0to17$data[fostercaredata_0to17$location=='Billings'] <- NA

# 4. Visually inspect output data
View(fostercaredata_0to17)
```

## STEP 2: ADD TO DATABASE
```{r}
#CHECK DATASET NAMED fostercaredata_0to17 TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',fostercaredata_0to17,append=TRUE,row.names=FALSE)
```

## STEP 3: CALCULATE THE RATE
```{r}
##########################################
#CALCULATE THE PERCENT FROM THE POPULATION

#write the sql code needed to pull population data from the database
childpop_sql <- paste0("SELECT location, timeframe, age_group, data FROM montana WHERE (timeframe='",year,"' AND vintageyear='",year,"') AND varname='childpopulationbysingleyearofage';")


childpop <- dbGetQuery(con,childpop_sql) %>%
  mutate(age_group=as.numeric(age_group)) %>%
  subset(age_group<18) %>%
  
  #create 0 to 17 age group sums
  mutate(data=as.numeric(paste(data))) %>%
  group_by(location,timeframe) %>%
  summarise(totalpop0to17=sum(data),.groups='keep') %>%
  ungroup


#write the sql code needed to pull the medicaid data that has already been published to the database
fostercare_sql <- paste0("SELECT location, locationtype, locationid, state, timeframe, dataformat, data, varname FROM montana WHERE timeframe='",year,"' AND varname='childrenages0to17infostercare' AND dataformat='Number';")

fostercare_pop <- dbGetQuery(con,fostercare_sql)


##############################
#CALCULATE PERCENTS FOR COUNTY, STATE
fostercare_percent <- fostercare_pop %>%
  left_join(childpop,by=c('location'='location','timeframe'='timeframe')) %>%
  rename(childinfostercare=data) %>%
  mutate(childinfostercare=as.numeric(paste(childinfostercare))) %>%
  mutate(data=(childinfostercare/totalpop0to17)*1000) %>%
  select(-c(childinfostercare,totalpop0to17)) %>%
  
  mutate(dataformat='Rate')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(fostercare_percent$locationid))>=1) {
  print(fostercare_percent$location[is.na(fostercare_percent$locationid)])
} else if (sum(is.na(fostercare_percent$locationid))==0) {
  'all locations match'
}

# 2. Visually inspect output data
view(fostercare_percent)


```

## STEP 4: ADD RATE TO DATABASE
```{r}
#CHECK DATASET NAMED fostercare_percent TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',fostercare_percent,append=TRUE,row.names=FALSE)

```

## STEP 3: OUTPUT FILE FOR KC
```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",year,"' AND varname='childrenages0to17infostercare';")


upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/safety/",statefile,"_",year,"_childrenages0to17infostercare.csv"),row.names=FALSE)
```




## -- FOR CHILDREN AGES 0 TO 5 -- ##

## CLEAN THE IMPORTED DATA
```{r}
fostercaredata_0to5 <- fostercaredata %>% 
  
  rename(location=`...1`) %>%
  #select(c(location,all_of(paste0("CY ",year)))) %>%
  #rename(data=all_of(paste0("CY ",year))) %>%
  
  #select(c(location,all_of(paste0("CY ",year)))) %>%
  #rename(data=all_of(paste0("CY ",year))) %>%
  
  #new format in 2021
  rename(data=`Unduplicated kids in care 0-5`) %>%
  select(location,data) %>%
  subset(!is.na(location)) %>%
  
  
  
  #update locations
  mutate(location=replace(location,location=='State (including tribal)','Montana')) %>%
  filter(!location %in% c("Blackfeet Tribal SS","Crow Tribal SS","Flathead Tribal SS","Fort Belknap Tribal SS","Northern Cheyenne Tribal SS","Rocky Boy Tribal SS")) %>%
  mutate(locationtype=ifelse(location=='Montana','State','County')) %>%
  mutate(location=replace(location,location=='Choteau','Chouteau')) %>%
  mutate(location=replace(location,location=='Lewis and Clark','Lewis & Clark')) %>%
  mutate(location=replace(location,location=='Power River','Powder River')) %>%
  mutate(location=replace(location,location=='Sweetgrass','Sweet Grass')) %>%

  mutate(state='Montana',
         dataformat='Number',
         varname='childrenages0to5infostercare') %>%
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
    
  mutate(timeframe=year) %>%

  #suppress if less than 4
  mutate(data=replace(data,data<4,NA))
  



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(fostercaredata_0to5$locationid))>=1) {
  print(fostercaredata_0to5$location[is.na(fostercaredata_0to5$locationid)])
} else if (sum(is.na(fostercaredata_0to5$locationid))==0) {
  'all locations match'
}


#2. Check that no values are less than 4
temp_suppresscheck <- fostercaredata_0to5 %>% subset(data<4)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- fostercaredata_0to5 %>%
  subset(locationtype=='County') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
} else if (temp_rows==1) {
  View(temp_testsuppression)
}

#if there is only one county, print the counties in ascending order
#temp_check <- fostercaredata_0to5 %>%
#  subset(locationtype=='County') %>%
#  arrange(data)
#View(temp_check)

#******
#CHANGE TO THE CORRECT COUNTY NAME HERE
#**Choose the county with the next lowest count of births; if there is a tie, suppress both
#fostercaredata_0to5$data[fostercaredata_0to5$location=='Billings'] <- NA

# 4. Visually inspect output data
View(fostercaredata_0to5)
```

## STEP 2: ADD TO DATABASE
```{r}
#CHECK DATASET NAMED fostercaredata_0to5 TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',fostercaredata_0to5,append=TRUE,row.names=FALSE)
```

## STEP 3: CALCULATE THE RATE
```{r}
##########################################
#CALCULATE THE PERCENT FROM THE POPULATION

#write the sql code needed to pull population data from the database
childpop_sql <- paste0("SELECT location, timeframe, age_group, data FROM montana WHERE (timeframe='",year,"' AND vintageyear='",year,"') AND varname='childpopulationbysingleyearofage';")


childpop <- dbGetQuery(con,childpop_sql) %>%
  mutate(age_group=as.numeric(age_group)) %>%
  subset(age_group<=5) %>%
  
  #create 0 to 5 age group sums
  mutate(data=as.numeric(paste(data))) %>%
  group_by(location,timeframe) %>%
  summarise(totalpop0to5=sum(data),.groups='keep') %>%
  ungroup


#write the sql code needed to pull the medicaid data that has already been published to the database
fostercare_sql <- paste0("SELECT location, locationtype, locationid, state, timeframe, dataformat, data, varname FROM montana WHERE timeframe='",year,"' AND varname='childrenages0to5infostercare' AND dataformat='Number';")

fostercare_pop <- dbGetQuery(con,fostercare_sql)


##############################
#CALCULATE PERCENTS FOR COUNTY, STATE
fostercare_percent <- fostercare_pop %>%
  left_join(childpop,by=c('location'='location','timeframe'='timeframe')) %>%
  rename(childinfostercare=data) %>%
  mutate(childinfostercare=as.numeric(paste(childinfostercare))) %>%
  mutate(data=(childinfostercare/totalpop0to5)*1000) %>%
  select(-c(childinfostercare,totalpop0to5)) %>%
  
  mutate(dataformat='Rate')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(fostercare_percent$locationid))>=1) {
  print(fostercare_percent$location[is.na(fostercare_percent$locationid)])
} else if (sum(is.na(fostercare_percent$locationid))==0) {
  'all locations match'
}

# 2. Visually inspect output data
view(fostercare_percent)


```

## STEP 4: ADD RATE TO DATABASE
```{r}
#CHECK DATASET NAMED fostercare_percent TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',fostercare_percent,append=TRUE,row.names=FALSE)

```

## STEP 3: OUTPUT FILE FOR KC
```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",year,"' AND varname='childrenages0to5infostercare';")

upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/safety/",statefile,"_",year,"_childrenages0to5infostercare.csv"),row.names=FALSE)
```







## CREATE DATA FILE TO COPY TO EXCEL FOR HMHB
```{r}
excel0to17_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE varname='childrenages0to17infostercare';")

excel0to17_county <- dbGetQuery(con,excel0to17_sql) %>%
  pivot_wider(names_from='timeframe',values_from='data') %>%
  pivot_wider(names_from='dataformat', values_from=c('2017','2018','2019','2020','2021')) %>%
  select(c(`location`,`2021_Number`,`2021_Rate`,
           `2020_Number`,`2020_Rate`,`2019_Number`,`2019_Rate`,
           `2018_Number`,`2018_Rate`,`2017_Number`,`2017_Rate`)) %>%
  subset(location != 'Montana') %>%
  arrange(location)

excel0to17_state <- dbGetQuery(con,excel0to17_sql) %>%
  pivot_wider(names_from='timeframe',values_from='data') %>%
  pivot_wider(names_from='dataformat', values_from=c('2017','2018','2019','2020','2021')) %>%
  select(c(`location`,`2021_Number`,`2021_Rate`,
           `2020_Number`,`2020_Rate`,`2019_Number`,`2019_Rate`,
           `2018_Number`,`2018_Rate`,`2017_Number`,`2017_Rate`)) %>%
  subset(location == 'Montana') %>%
  arrange(location)

excel0to17 <- excel0to17_county %>%
  bind_rows(excel0to17_state)

write.csv(excel0to17,'./Output/safety/MT Foster Care Excel/mt_fostercare_tocopy_0to17.csv',row.names=FALSE)


excel0to5_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE varname='childrenages0to5infostercare';")

excel0to5_county <- dbGetQuery(con,excel0to5_sql) %>%
  pivot_wider(names_from='timeframe',values_from='data') %>%
  pivot_wider(names_from='dataformat', values_from=c('2017','2018','2019','2020','2021')) %>%
  select(c(`location`,`2021_Number`,`2021_Rate`,
           `2020_Number`,`2020_Rate`,`2019_Number`,`2019_Rate`,
           `2018_Number`,`2018_Rate`,`2017_Number`,`2017_Rate`)) %>%
  subset(location != 'Montana') %>%
  arrange(location)

excel0to5_state <- dbGetQuery(con,excel0to5_sql) %>%
  pivot_wider(names_from='timeframe',values_from='data') %>%
  pivot_wider(names_from='dataformat', values_from=c('2017','2018','2019','2020','2021')) %>%
  select(c(`location`,`2021_Number`,`2021_Rate`,
           `2020_Number`,`2020_Rate`,`2019_Number`,`2019_Rate`,
           `2018_Number`,`2018_Rate`,`2017_Number`,`2017_Rate`)) %>%
  subset(location == 'Montana') %>%
  arrange(location)

excel0to5 <- excel0to5_county %>%
  bind_rows(excel0to5_state)
  
  
write.csv(excel0to5,'./Output/safety/MT Foster Care Excel/mt_fostercare_tocopy_0to5.csv',row.names=FALSE)


```



## -- Average days a child is in state care before exiting -- ##
## STEP 1: create the data file manually
```{r}
#edit the data only
data <- 592

#----
location <- 'Montana'
locationtype <- 'State'
locationid <- '28'
state <- 'Montana'
timeframe <- all_of(year)
dataformat <- 'Number'
varname <- 'averagedaysachildisinstatecarebeforeexiting'


fostercareexit <- data.frame(location,locationtype,locationid,state,timeframe,dataformat,varname,data)

```

## STEP 2: Add to KC database
```{r}
#CHECK DATASET NAMED fostercareexit TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'montana',fostercareexit,append=TRUE,row.names=FALSE)

```

## STEP 3: OUTPUT FILE FOR KC
```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM montana WHERE timeframe='",year,"' AND varname='averagedaysachildisinstatecarebeforeexiting';")

upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/safety/",statefile,"_",year,"_averagedaysachildisinstatecarebeforeexiting.csv"),row.names=FALSE)
```

