---
title: "North Dakota Public Assistance"
author: "Xanna Burg"
date: "2/20/2020"
output: html_document
---

## Indicator 1: TANF recipients ages 0 to 19
## Indicator 2: SNAP recipients ages 0 to 18
## Indicator 3: Child care assistance recipients ages 0 to 13
## Indicator 4: Medicaid recipients ages 0 to 20
## Indicator 5: Children ages 0 to 18 enrolled in Healthy Steps

**Created by:** Xanna Burg
**Date:** February, 2020
**Updated by:**

**Data Source:** North Dakota Department of Human Services 
**Purpose:** Clean the public assistance provided by the Department of Human Services and update KIDS COUNT data for North Dakota.
**Description:** this program calculates five indicators 
**TANF recipients ages 0 to 19**
**SNAP recipients age 0 to 18**
**Child care assitance recipients ages 0 to 13**
**Medicaid recipients ages 0 to 20**
**Chidren ages 0 to 18 enrolled in Healthy Steps**

* Geographies: state total, county, state planning regions (where applicable, which are groups of counties)
* Number - count of individuals participating in program
* Percent - The percent of individuals participating in program out of estimated population in age range. [Numerator: count of participation, Denominator: children in corresponding age group from Census estimates] 

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
```

```{r}
#UPDATE to reflect the current year data working with
year <- '2024'
fiscalyear <- 'FY24'
state <- 'northdakota'


year_minus1 <- as.character(as.numeric(year)-1)
```


```{r}
#read in the data in Excel format with multiple sheets

df_tanf <- read_excel(paste0("./Input/economics/northdakota_",fiscalyear,"_economicassistance_tanf.xlsx"))
df_snap <- read_excel(paste0("./Input/economics/northdakota_",fiscalyear,"_economicassistance_snap.xlsx"))
df_ccap <- read_excel(paste0("./Input/economics/northdakota_",fiscalyear,"_economicassistance_ccap.xlsx"))
df_medicaid <- read_excel(paste0("./Input/economics/northdakota_",fiscalyear,"_economicassistance_medicaid.xlsx"))
df_healthysteps <- read_excel(paste0("./Input/economics/northdakota_",fiscalyear,"_economicassistance_chip.xlsx"))

#input location ID file for ND
locationids <- read.csv("./Input/ND KC Location IDs.csv")

#input region ID file for ND
regionids <- read.csv("./Input/ND KC Region List.csv")
regionids$county <- as.character(regionids$county)

#input county list file for ND
countylist <- read.csv("./Input/ND County List.csv")

```


## -- SNAP -- ##

```{r}
#FOR SNAP
#clean the county data
df_snap2 <- df_snap %>%
  rename(location=County,
         data=`Unduplicated Count of SNAP Recipients age 0-18`) %>%
  #remove extra rows or counties not a ND county
  subset(!is.na(location)) %>%
  subset(location != 'County Not Available / Out of State County') %>%
  #remove the word "County"
  mutate(location=gsub(' County','',location)) %>%
  #merge in county list, as some counties with 0 are not in the dataset
  full_join(countylist, by=c('location'='Location')) %>%
  mutate(data=replace(data,is.na(data),0)) %>%
  #add in Kids Count variables
  mutate(locationtype='County') %>%
  mutate(state='North Dakota') %>%
  mutate(dataformat='Number') %>%
  mutate(timeframe=year) %>%
  mutate(varname='snapages0to18') 

#create state summary
df_snap_state <- df_snap2 %>%
  group_by(state,timeframe,dataformat,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(location='North Dakota') %>%
  mutate(locationtype='State') 

#create region summaries
df_snap_regions <- df_snap2 %>% 
  left_join(regionids,by=c('location'='county')) %>%
  select(-c(location)) %>%
  rename(location=region) %>%
  
  group_by(location,state,dataformat,timeframe,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(locationtype='Planning Region') 

#union all three datasets together and add location ID
snapages0to18 <- df_snap2 %>%
  bind_rows(df_snap_state) %>%
  bind_rows(df_snap_regions) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(snapages0to18$locationid))>=1) {
  print(snapages0to18$location[is.na(snapages0to18$locationid)])
} else if (sum(is.na(snapages0to18$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(snapages0to18)
```

```{r}
#add to database count of snapages0to18
dbWriteTable(con,'northdakota',snapages0to18,append=TRUE,row.names=FALSE)
```

```{r}
#calculate percent of population 0 to 18 on snap

#create dataset that has population merged with geography for 0 to 18 year olds
#population is for the beginning of the FY, which is minus 1 from the year that is reported
snap_pop_sql <- paste0("SELECT locationid,location,age_group,data
                     FROM northdakota
                     WHERE timeframe = '",year_minus1,"' AND vintageyear='",year_minus1,"' AND varname = 'childpopulationbysingleyearofage' AND (age_group='0' OR age_group='1' OR age_group='2' OR age_group='3' OR age_group='4' OR age_group='5' OR age_group='6' OR age_group='7' OR age_group='8' OR age_group='9' OR age_group='10' OR age_group='11' OR age_group='12' OR age_group='13' OR age_group='14' OR age_group='15' OR age_group='16' OR age_group='17' OR age_group='18') ;")
pop0_18 <- dbGetQuery(con,snap_pop_sql) %>%
  mutate(data=as.numeric(paste(data)))

#sum the population across all age groups 0 to 18
pop0_18_sums <- pop0_18 %>%
  group_by(location) %>%
  summarise(total_pop=sum(data),.groups='keep')

#merge population with snap data
snap_percent <- snapages0to18 %>%
  left_join(pop0_18_sums,by='location') %>%
  mutate(percent_snap=(data/total_pop)) %>%
  select(-c(data,total_pop)) %>%
  rename(data=percent_snap) %>%
  mutate(dataformat='Percent')



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(snap_percent$locationid))>=1) {
  print(snap_percent$location[is.na(snap_percent$locationid)])
} else if (sum(is.na(snap_percent$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- snap_percent %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}


# 3. Visually inspect output data
View(snap_percent)
```

```{r}
#add to database percent of snapages0to18
dbWriteTable(con,'northdakota',snap_percent,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#COUNT AND PERCENT OF SNAP

upload_data_snap_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'snapages0to18';")

upload_data_snap <- dbGetQuery(con,upload_data_snap_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_snap,file=paste0("./Output/economics/",state,"_",year,"_snapages0to18.csv"),row.names=FALSE)
```





## -- TANF -- ##

```{r}
#FOR TANF
#clean the county data
df_tanf2 <- df_tanf %>%
  rename(location=County,
         data=`Unduplicated Count of TANF Recipients age 0-19`) %>%
  #remove extra rows or counties not a ND county
  subset(!is.na(location)) %>%
  subset(location != 'County Not Available / Out of State County') %>%
  #remove the word "County"
  mutate(location=gsub(' County','',location)) %>%
  #merge in county list, as some counties with 0 are not in the dataset
  full_join(countylist, by=c('location'='Location')) %>%
  mutate(data=replace(data,is.na(data),0)) %>%
  #add in Kids Count variables %>%
  mutate(locationtype='County') %>%
  mutate(state='North Dakota') %>%
  mutate(dataformat='Number') %>%
  mutate(timeframe=year) %>%
  mutate(varname='tanfages0to19') 
 

#create state summary
df_tanf_state <- df_tanf2 %>%
  group_by(state,timeframe,dataformat,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(location='North Dakota',
         locationtype='State') 

#create region summaries
df_tanf_regions <- df_tanf2 %>% 
  left_join(regionids,by=c('location'='county')) %>%
  select(-c(location)) %>%
  rename(location=region) %>%
  
  group_by(location,state,dataformat,timeframe,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(locationtype='Planning Region') 
  

#union all three datasets together and add location ID
tanfages0to19 <- df_tanf2 %>%
  bind_rows(df_tanf_state) %>%
  bind_rows(df_tanf_regions) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  
  
####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(tanfages0to19$locationid))>=1) {
  print(tanfages0to19$location[is.na(tanfages0to19$locationid)])
} else if (sum(is.na(tanfages0to19$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(tanfages0to19)
```

```{r}
#add to database count of tanfages0to19
dbWriteTable(con,'northdakota',tanfages0to19,append=TRUE,row.names=FALSE)
```

```{r}
#calculate percent of population 0 to 19 on tanf

#create dataset that has population merged with geography for 0 to 19 year olds
#population is for the beginning of the FY, which is minus 1 from the year that is reported
tanf_pop_sql <- paste0("SELECT locationid,location,age_group,data
                     FROM northdakota
                     WHERE timeframe = '",year_minus1,"' AND vintageyear='",year_minus1,"' AND varname = 'childpopulationbysingleyearofage' AND (age_group='0' OR age_group='1' OR age_group='2' OR age_group='3' OR age_group='4' OR age_group='5' OR age_group='6' OR age_group='7' OR age_group='8' OR age_group='9' OR age_group='10' OR age_group='11' OR age_group='12' OR age_group='13' OR age_group='14' OR age_group='15' OR age_group='16' OR age_group='17' OR age_group='18' OR age_group='19') ;")

pop0_19 <- dbGetQuery(con,tanf_pop_sql) %>%
  mutate(data=as.numeric(paste(data)))

#sum the population across all age groups 0 to 19
pop0_19_sums <- pop0_19 %>%
  group_by(location) %>%
  summarise(total_pop=sum(data),.groups='keep')

#merge population with snap data
tanf_percent <- tanfages0to19 %>%
  left_join(pop0_19_sums,by='location') %>%
  mutate(percent_tanf=(data/total_pop)) %>%
  select(-c(data,total_pop)) %>%
  rename(data=percent_tanf) %>%
  mutate(dataformat='Percent')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(tanf_percent$locationid))>=1) {
  print(tanf_percent$location[is.na(tanf_percent$locationid)])
} else if (sum(is.na(tanf_percent$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- tanf_percent %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}


# 3. Visually inspect output data
View(tanf_percent)
```

```{r}
#add to database percent of tanfages0to19
dbWriteTable(con,'northdakota',tanf_percent,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#COUNT AND PERCENT OF TANF

upload_data_tanf_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'tanfages0to19';")
upload_data_tanf <- dbGetQuery(con,upload_data_tanf_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_tanf,file=paste0("./Output/economics/",state,"_",year,"_tanfages0to19.csv"),row.names=FALSE)
```


## -- CCAP -- ##

```{r}
#FOR CCAP
#clean the county data
df_ccap2 <- df_ccap %>%
  rename(location=County) %>%
  mutate(data=`Unduplicated count of CCAP recipients of age 0-5`+`Unduplicated count of CCAP recipients of age 6+`) %>%
  select(-c(`Unduplicated count of CCAP recipients of age 0-5`,`Unduplicated count of CCAP recipients of age 6+`)) %>%
  #remove extra rows or counties not a ND county
  subset(!is.na(location)) %>%
  subset(location != 'County Not Available / Out of State County') %>%
  #remove the word "County"
  mutate(location=gsub(' County','',location)) %>%
  #merge in county list, as some counties with 0 are not in the dataset
  full_join(countylist, by=c('location'='Location')) %>%
  mutate(data=replace(data,is.na(data),0)) %>%
  #add in Kids Count variables %>%
  mutate(locationtype='County') %>%
  mutate(state='North Dakota') %>%
  mutate(dataformat='Number') %>%
  mutate(timeframe=year) %>%
  mutate(varname='ccapages0to13') 
  

#create state summary
df_ccap_state <- df_ccap2 %>%
  group_by(state,timeframe,dataformat,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(location='North Dakota',
         locationtype='State') 

#create region summaries
df_ccap_regions <- df_ccap2 %>% 
  left_join(regionids,by=c('location'='county')) %>%
  select(-c(location)) %>%
  rename(location=region) %>%
  
  group_by(location,state,dataformat,timeframe,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(locationtype='Planning Region') 


#union all three datasets together and add location ID
ccapages0to13 <- df_ccap2 %>%
  bind_rows(df_ccap_state) %>%
  bind_rows(df_ccap_regions) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  
 
####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(ccapages0to13$locationid))>=1) {
  print(ccapages0to13$location[is.na(ccapages0to13$locationid)])
} else if (sum(is.na(ccapages0to13$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(ccapages0to13)
```

```{r}
#add to database count of ccapages0to13
dbWriteTable(con,'northdakota',ccapages0to13,append=TRUE,row.names=FALSE)
```

```{r}
#calculate percent of population 0 to 13 on ccap

#create dataset that has population merged with geography for 0 to 13 year olds
#population is for the beginning of the FY, which is minus 1 from the year that is reported
ccap_pop_sql <- paste0("SELECT locationid,location,age_group,data
                     FROM northdakota
                     WHERE timeframe = '",year_minus1,"' AND vintageyear='",year_minus1,"' AND varname = 'childpopulationbysingleyearofage' AND (age_group='0' OR age_group='1' OR age_group='2' OR age_group='3' OR age_group='4' OR age_group='5' OR age_group='6' OR age_group='7' OR age_group='8' OR age_group='9' OR age_group='10' OR age_group='11' OR age_group='12' OR age_group='13') ;")

pop0_13 <- dbGetQuery(con,ccap_pop_sql) %>%
  mutate(data=as.numeric(paste(data)))

#sum the population across all age groups 0 to 13
pop0_13_sums <- pop0_13 %>%
  group_by(location) %>%
  summarise(total_pop=sum(data),.groups='keep')

#merge population with ccap data
ccap_percent <- ccapages0to13 %>%
  left_join(pop0_13_sums,by='location') %>%
  mutate(percent_ccap=(data/total_pop)) %>%
  select(-c(data,total_pop)) %>%
  rename(data=percent_ccap) %>%
  mutate(dataformat='Percent')

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(ccap_percent$locationid))>=1) {
  print(ccap_percent$location[is.na(ccap_percent$locationid)])
} else if (sum(is.na(ccap_percent$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- ccap_percent %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}


# 3. Visually inspect output data
View(ccap_percent)
```

```{r}
#add to database percent of ccapages0to13
dbWriteTable(con,'northdakota',ccap_percent,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#COUNT AND PERCENT OF CCAP

upload_data_ccap_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'ccapages0to13';")
upload_data_ccap <- dbGetQuery(con,upload_data_ccap_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_ccap,file=paste0("./Output/economics/",state,"_",year,"_ccapages0to13.csv"),row.names=FALSE)
```



## -- MEDICAID -- ##
```{r}
#FOR MEDICAID
#clean the county data
df_medicaid2 <- df_medicaid %>%
  rename(location=County,
         data=`MED_EX_CHIP`) %>%
  #remove extra rows or counties not a ND county
  subset(!is.na(location)) %>%
  subset(location != 'County Not Available / Out of State County') %>%
  #remove the word "County"
  mutate(location=gsub(' County','',location)) %>%
  #merge in county list, as some counties with 0 are not in the dataset
  full_join(countylist, by=c('location'='Location')) %>%
  mutate(data=replace(data,is.na(data),0)) %>%
  #add in Kids Count variables %>%
  mutate(locationtype='County') %>%
  mutate(state='North Dakota') %>%
   mutate(dataformat='Number') %>%
  mutate(timeframe=year) %>%
  mutate(varname='medicaidages0to20') 


#create state summary
df_medicaid_state <- df_medicaid2 %>%
  group_by(state,timeframe,dataformat,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(location='North Dakota',
         locationtype='State') 

#create region summaries
df_medicaid_regions <- df_medicaid2 %>% 
  left_join(regionids,by=c('location'='county')) %>%
  select(-c(location)) %>%
  rename(location=region) %>%
  
  group_by(location,state,dataformat,timeframe,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(locationtype='Planning Region') 


#union all three datasets together and add location ID
medicaidages0to20 <- df_medicaid2 %>%
  bind_rows(df_medicaid_state) %>%
  bind_rows(df_medicaid_regions) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
 
####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(medicaidages0to20$locationid))>=1) {
  print(medicaidages0to20$location[is.na(medicaidages0to20$locationid)])
} else if (sum(is.na(medicaidages0to20$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(medicaidages0to20)
```

```{r}
#add to database count of medicaidages0to20
dbWriteTable(con,'northdakota',medicaidages0to20,append=TRUE,row.names=FALSE)
```

```{r}
#calculate percent of population 0 to 20 on medicaid

#create dataset that has population merged with geography for 0 to 20 year olds
#population is for the beginning of the FY, which is minus 1 from the year that is reported
medicaid_pop_sql <- paste0("SELECT locationid,location,age_group,data
                     FROM northdakota
                     WHERE timeframe = '",year_minus1,"' AND vintageyear='",year_minus1,"' AND varname = 'childpopulationbysingleyearofage';")



pop0_20 <- dbGetQuery(con,medicaid_pop_sql) %>%
  mutate(age_group=as.numeric(paste(age_group))) %>%
  subset(age_group<=20) %>%
  mutate(data=as.numeric(paste(data)))

#sum the population across all age groups 0 to 20
pop0_20_sums <- pop0_20 %>%
  group_by(location) %>%
  summarise(total_pop=sum(data),.groups='keep')

#merge population with snap data
medicaid_percent <- medicaidages0to20 %>%
  left_join(pop0_20_sums,by='location') %>%
  mutate(percent_medicaid=(data/total_pop)) %>%
  select(-c(data,total_pop)) %>%
  rename(data=percent_medicaid) %>%
  mutate(dataformat='Percent') %>%
  mutate(data=if_else(data>1,1,data))

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(medicaid_percent$locationid))>=1) {
  print(medicaid_percent$location[is.na(medicaid_percent$locationid)])
} else if (sum(is.na(medicaid_percent$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- medicaid_percent %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}


# 3. Visually inspect output data
View(medicaid_percent)
```

```{r}
#add to database percent of medicaidages0to20
dbWriteTable(con,'northdakota',medicaid_percent,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#COUNT AND PERCENT OF MEDICAID

upload_data_medicaid_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'medicaidages0to20';")
upload_data_medicaid <- dbGetQuery(con,upload_data_medicaid_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_medicaid,file=paste0("./Output/economics/",state,"_",year,"_medicaidages0to20.csv"),row.names=FALSE)
```



## -- HEALTHY STEPS -- ##

```{r}
#FOR HEALTHY STEPS
#clean the county data
df_healthysteps2 <- df_healthysteps %>%
 rename(location=County,
         data=`Unduplicated Health Steps/CHIP (optional children’s group) recipients age 0-18`) %>%
  #remove extra rows or counties not a ND county
  subset(!is.na(location)) %>%
  subset(location != 'County Not Available / Out of State County') %>%
  #remove the word "County"
  mutate(location=gsub(' County','',location)) %>%
  #merge in county list, as some counties with 0 are not in the dataset
  full_join(countylist, by=c('location'='Location')) %>%
  mutate(data=replace(data,is.na(data),0)) %>%
  #add in Kids Count variables %>%
  mutate(locationtype='County') %>%
  mutate(state='North Dakota') %>%
    mutate(dataformat='Number') %>%
  mutate(timeframe=year) %>%
  mutate(varname='healthystepsages0to18') 



#create state summary
df_healthysteps_state <- df_healthysteps2 %>%
  group_by(state,timeframe,dataformat,varname) %>%
  summarise(data=sum(data,na.rm=TRUE),.groups='keep') %>%
  mutate(location='North Dakota',
         locationtype='State') 

#create region summaries
df_healthysteps_regions <- df_healthysteps2 %>%
  left_join(regionids,by=c('location'='county')) %>%
  select(-c(location)) %>%
  rename(location=region) %>%
  
  group_by(location,state,dataformat,timeframe,varname) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(locationtype='Planning Region') 


#union all three datasets together and add location ID
healthystepsages0to18 <- df_healthysteps2 %>%
  bind_rows(df_healthysteps_state) %>%
  bind_rows(df_healthysteps_regions) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  
 
####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(healthystepsages0to18$locationid))>=1) {
  print(healthystepsages0to18$location[is.na(healthystepsages0to18$locationid)])
} else if (sum(is.na(healthystepsages0to18$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(healthystepsages0to18)
```

```{r}
#add to database count of healthystepsages0to18
dbWriteTable(con,'northdakota',healthystepsages0to18,append=TRUE,row.names=FALSE)
```

```{r}
#calculate percent of population 0 to 18 on healthysteps

#use the population dataset created for SNAP
#population is for the beginning of the FY, which is minus 1 from the year that is reported

#merge population with snap data
healthysteps_percent <- healthystepsages0to18 %>%
  left_join(pop0_18_sums,by='location') %>%
  mutate(percent_healthysteps=(data/total_pop)) %>%
  select(-c(data,total_pop)) %>%
  rename(data=percent_healthysteps) %>%
  mutate(dataformat='Percent')

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(healthysteps_percent$locationid))>=1) {
  print(healthysteps_percent$location[is.na(healthysteps_percent$locationid)])
} else if (sum(is.na(healthysteps_percent$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- healthysteps_percent %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}


# 3. Visually inspect output data
View(healthysteps_percent)
```

```{r}
#add to database percent of healthystepsages0to18
dbWriteTable(con,'northdakota',healthysteps_percent,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#COUNT AND PERCENT OF HEALTHY STEPS

upload_data_healthysteps_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'healthystepsages0to18';")
upload_data_healthysteps <- dbGetQuery(con,upload_data_healthysteps_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_healthysteps,file=paste0("./Output/economics/",state,"_",year,"_healthystepsages0to18.csv"),row.names=FALSE)
```