---
title: "Child Care Providers by Type"
author: "Xanna Burg"
output: html_document
---

## Indicator 1: Child Care Providers by Type
## Indicator 2: Child Care Capacity by Type
## Indicator 3: Child Care Providers by QRIS
**Created by:** Xanna Burg
**Date:** February, 2020
**Updated by:**

**Data Source:** North Dakota Department of Human Services 
**Purpose:** Clean the child care data provided by the Department of Human Services and update KIDS COUNT data for North Dakota.

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), Category (character), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
#UPDATE to reflect the current year data working with
year <- '2026'
state <- 'northdakota'

year_shortened <- substr(year,3,4)

#only need to change if using a different month than February
month <- '2'

#read in the codes for county name
countymatch <- read.csv("./Documentation/Indicator Documentation/North Dakota Data Requests/ND County Codes_for childcare.csv")
```

```{r}
#read in the child care data
child_data <- read_excel(paste0("./Input/education/northdakota_",year,"_childcaredata.xlsx"))

#read in the quality data
quality_data <- read_excel(path=paste0("./Input/education/northdakota_",year,"_childcarequality.xlsx"))




#create a dataset that has all combinations of county and category for final provider type categories
#this is to account for the fact that some combinations will be 0
providertype_kc_category <- c('Licensed','Licensed School Age','Self-declaration Providers')
providertype_kc_categorylist <- as.data.frame(providertype_kc_category)
county_provider_categorylist <- crossing(countymatch,providertype_kc_categorylist) %>%
  select(-c(ndcounty_number))




#create all possible combos of quality categories and counties
quality <- c('Step 1','Step 2','Step 3','Step 4')
qualitylist <- as.data.frame(quality)
county_quality_categorylist <-crossing(countymatch,qualitylist) %>%
  select(-c(ndcounty_number))

#input location ID file for ND
locationids <- read.csv("./Input/ND KC Location IDs.csv")

#input region ID file for ND
regionids <- read.csv("./Input/ND KC Region List.csv")

#input ID file for county to city
countyids <- read_excel("./Documentation/Indicator Documentation/North Dakota Data Requests/nd_city_county child care key v2.xlsx")

```

```{r}
child_data_2 <- child_data %>%
  
  #clean up the data
  #use county reported in updated file (some years had to use city or another location variable)
  rename(county=County) %>%
  
  #create new variable matching to KC categories
  mutate(category=case_when(
    `Program Type Name`=='Child Care Center' | `Program Type Name`=='Family Child Care' | 
      `Program Type Name`=='Group Child Care Facility' | `Program Type Name`=='Group Child Care Home' | 
      `Program Type Name`=='Multiple Licensed Facility' | 
      `Program Type Name`=='Preschool Education Facility' ~ 'Licensed',
    `Program Type Name`=='School Age Care' ~ 'Licensed School Age',
    `Program Type Name`=='Self-Declaration' ~ 'Self-declaration Providers')) %>%
  
  subset(City != 'Test') %>%
  subset(City !='	test city') %>%
  subset(`Program Name` != 'ted') %>%
  subset(License != '0008153C') %>%
  subset(License != '001157E') %>%
  subset(License != '0001162C') %>%
  subset(License !='09857775F') %>%
  
  mutate(capacity=as.numeric(paste(`Capacity`))) 

```



## -- PROVIDERS AND CAPACITY BY TYPE --##
NOTE: SCHOOL AGE INCLUDED
```{r}
#COUNTY DATA
countysums_providerscapacity <- child_data_2 %>%

  group_by(county,category) %>%
  summarise(providers = n(),
            capacity=sum(capacity),.groups='keep') %>%
  
  #fill in missing county/group combinations
  right_join(county_provider_categorylist,by=c('county'='county','category'='providertype_kc_category')) %>%
  
  #for the missing combinations that are NA, make them zero
  mutate(providers = replace_na(providers, 0),
         capacity= replace_na(capacity,0)) %>%
  
  #fix kc naming
  rename(location=county) %>%
  mutate(locationtype='County') 


##STATE DATA
#start with county data
statesums_providerscapacity <- countysums_providerscapacity %>%
  group_by(category) %>%
  summarise(providers = sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  
  mutate(location='North Dakota') %>%
  mutate(locationtype='State') 


##STATE PLANNING REGIONS 
#start with county data
regionsums_providerscapacity <- countysums_providerscapacity %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region,category) %>%
  summarise(providers = sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  
  rename(location=region) %>%
  mutate(locationtype='Planning Region') 



#format to database count child care
childcareprovidersbytype <- countysums_providerscapacity %>%
  bind_rows(statesums_providerscapacity) %>%
  bind_rows(regionsums_providerscapacity) %>%
  select(-c(capacity)) %>%
  rename(data=providers) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcareprovidersbytype') 


childcarecapacitybytype <- countysums_providerscapacity %>%
  bind_rows(statesums_providerscapacity) %>%
  bind_rows(regionsums_providerscapacity) %>%
  select(-c(providers)) %>%
  
  #remove unlicensed facilities, we can no longer report
  subset(category=='Licensed' | category=='Licensed School Age') %>%
  
  rename(data=capacity) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcarecapacitybytype')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childcareprovidersbytype$locationid))>=1) {
  print(childcareprovidersbytype$location[is.na(childcareprovidersbytype$locationid)])
} else if (sum(is.na(childcareprovidersbytype$locationid))==0) {
  'all locations match'
}


if (sum(is.na(childcarecapacitybytype$locationid))>=1) {
  print(childcarecapacitybytype$location[is.na(childcarecapacitybytype$locationid)])
} else if (sum(is.na(childcarecapacitybytype$locationid))==0) {
  'all locations match'
}

#2. Visually inspect data
View(childcareprovidersbytype)
View(childcarecapacitybytype)
```


```{r}
#add to database providers and capacity of child care
dbWriteTable(con,'northdakota',childcareprovidersbytype,append=TRUE,row.names=FALSE)
dbWriteTable(con,'northdakota',childcarecapacitybytype,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#PROVIDERS OF CHILD CARE

upload_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname='childcareprovidersbytype';")
upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/education/",state,"_",year,"_childcareprovidersbytype.csv"),row.names=FALSE)

#CAPACITY OF CHILD CARE

upload_sql2 <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname='childcarecapacitybytype' AND dataformat='Number';")
upload_data2 <- dbGetQuery(con,upload_sql2) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data2,file=paste0("./Output/education/",state,"_",year,"_childcarecapacitybytype.csv"),row.names=FALSE)
```






## -- Child Care Providers and Capacity by QRIS -- ##
NOTE: SCHOOL AGE NOT INCLUDED
```{r}
####CREATE COUNTY SUMMARY COUNT OF CHILD CARE
countysums_quality <- quality_data %>%

  rename(category=`Quality Rating`,
         location=County,
         capacity=`Licensed Capacity`) %>%
  mutate(capacity=as.numeric(paste(capacity))) %>%
  group_by(location,category) %>%
  summarise(providers = n(),
            capacity=sum(capacity),.groups='keep') %>%
  ungroup() %>%
  
  #join with shell to make sure all county/quality categories are represented
  right_join(county_quality_categorylist,by=c('location'='county','category'='quality')) %>%


  mutate(providers = replace_na(providers, 0),
         capacity= replace_na(capacity,0)) %>%
  mutate(locationtype='County') %>%
  arrange(location,category)




####CREATE STATEWIDE SUMMARY COUNT OF CHILD CARE
statesums_quality <- countysums_quality %>%
  group_by(category) %>%
  summarise(providers=sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  ungroup() %>%
  mutate(location='North Dakota',
         locationtype='State')



####CREATE STATE PLANNING REGIONS SUMMARY COUNT OF CHILD CARE
#create blank shell in order to capture when 0 occurs
location <- c('Planning Region 1','Planning Region 1','Planning Region 1',
              'Planning Region 1',
              'Planning Region 2','Planning Region 2','Planning Region 2',
              'Planning Region 2',
              'Planning Region 3','Planning Region 3','Planning Region 3',
              'Planning Region 3',
              'Planning Region 4','Planning Region 4','Planning Region 4',
              'Planning Region 4',
              'Planning Region 5','Planning Region 5','Planning Region 5',
              'Planning Region 5',
              'Planning Region 6','Planning Region 6','Planning Region 6',
              'Planning Region 6',
              'Planning Region 7','Planning Region 7','Planning Region 7',
              'Planning Region 7',
              'Planning Region 8','Planning Region 8','Planning Region 8',
              'Planning Region 8')
category <- c('Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4',
              'Step 1','Step 2','Step 3','Step 4')
regionshell <- data.frame(location,category)

regionsums_quality <- countysums_quality %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region,category) %>%
  summarise(providers = sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  ungroup() %>%
  
  right_join(regionshell,by=c('region'='location','category'='category')) %>%
  mutate(providers = replace_na(providers, 0),
         capacity = replace_na(capacity,0)) %>%
  mutate(locationtype='Planning Region') %>%
  rename(location=region)

#combine all location types
quality_data_combined <- countysums_quality %>%
  bind_rows(statesums_quality) %>%
  bind_rows(regionsums_quality)

```

#Use licensed data to double check data quality, but no longer reporting percentages due to difference in reporting systems
```{r}
licensedproviders_sql <- paste0("SELECT location,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcareprovidersbytype' AND category='Licensed' AND dataformat='Number';")
licensedproviders <- dbGetQuery(con,licensedproviders_sql) %>%
  rename(totallicensedproviders=data) %>%
  mutate(totallicensedproviders=as.numeric(paste(totallicensedproviders)))

licensedcapacity_sql <- paste0("SELECT location,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcarecapacitybytype' AND category='Licensed' AND dataformat='Number';")
licensedcapacity <- dbGetQuery(con,licensedcapacity_sql) %>%
  rename(totallicensedcapacity=data) %>%
  mutate(totallicensedcapacity=as.numeric(paste(totallicensedcapacity)))

compare_dataset <- quality_data_combined %>%
  #create total
  group_by(location) %>%
  summarise(providers=sum(providers),
            capacity=sum(capacity)) %>%
  
  left_join(licensedproviders) %>%
  left_join(licensedcapacity) %>%
  
  #create new variables that indicate where quality data is greater than licensed data
  mutate(providerdiff=totallicensedproviders-providers) %>%
  mutate(capacitydiff=totallicensedcapacity-capacity) %>%
  subset(providerdiff<0 | capacitydiff<0)

View(compare_dataset)

```

Format the Kids Count Data Center
```{r}
#format to database count child care
childcareprovidersbyquality <- quality_data_combined %>%
  select(-c(capacity)) %>%
  rename(data=providers) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcareprovidersbyquality')


childcarecapacitybyquality <- quality_data_combined %>%
  select(-c(providers)) %>%
  rename(data=capacity) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcarecapacitybyquality')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childcareprovidersbyquality$locationid))>=1) {
  print(childcareprovidersbyquality$location[is.na(childcareprovidersbyquality$locationid)])
} else if (sum(is.na(childcareprovidersbyquality$locationid))==0) {
  'all locations match'
}


if (sum(is.na(childcarecapacitybyquality$locationid))>=1) {
  print(childcarecapacitybyquality$location[is.na(childcarecapacitybyquality$locationid)])
} else if (sum(is.na(childcarecapacitybyquality$locationid))==0) {
  'all locations match'
}

#2. Visually inspect data
View(childcareprovidersbyquality)
View(childcarecapacitybyquality)
```

```{r}
#add to database count of child care
dbWriteTable(con,'northdakota',childcareprovidersbyquality,append=TRUE,row.names=FALSE)
dbWriteTable(con,'northdakota',childcarecapacitybyquality,append=TRUE,row.names=FALSE)
```

Export to upload to Data Center
```{r}
#write query from database to get needed format for KC data center
#PROVIDERS
upload_data_qualityproviders_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'childcareprovidersbyquality';")
upload_data_qualityproviders <- dbGetQuery(con,upload_data_qualityproviders_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_qualityproviders,file=paste0("./Output/education/",state,"_",year,"_childcareprovidersbyquality.csv"),row.names=FALSE)

#CAPACITY
upload_data_qualitycapacity_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'childcarecapacitybyquality';")
upload_data_qualitycapacity <- dbGetQuery(con,upload_data_qualitycapacity_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_qualitycapacity,file=paste0("./Output/education/",state,"_",year,"_childcarecapacitybyquality.csv"),row.names=FALSE)
```
