---
title: "Child Care Providers by Type"
author: "Xanna Burg"
output: html_document
---

## Indicator 1: Child Care Providers by Type
## Indicator 2: Child Care Capacity by Type
## Indicator 3: Child Care Providers by QRIS
## Indicator 4: Child Care Capacity by QRIS
**Created by:** Xanna Burg
**Date:** February, 2020
**Updated by:**

**Data Source:** North Dakota Department of Human Services 
**Purpose:** Clean the child care data provided by the Department of Human Services and update KIDS COUNT data for North Dakota.

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), Category (character), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
#UPDATE to reflect the current year data working with
year <- '2020'
state <- 'northdakota'
```

```{r}
#read in the most recent year of child care data
child_data <- read.csv(paste0("./Input/education/northdakota_",year,"_childcaredata.csv"))

#read in the quality data
quality_data <- read_excel(path=paste0("./Input/education/northdakota_",year,"_childcarequality.xlsx"),sheet='February 2020')

#read in the codes for county name
countymatch <- read.csv("./Documentation/Indicator Documentation/North Dakota Data Requests/ND County Codes_for childcare.csv")

#create dataset to match the provider type
provider_code <- c('C','E','F','G','H','I','K','M','P','Q','R','S','Z')
provider_name <- c('Child Care Center','Preschool','Family Child Care',
             'Group Child Care in a home','Group Child Care in a facility',
             'In-home registered provider','School Age Care',
             'Multiple License (most often center and Preschool)',
             'Early Head Start-Child Care Partnerships','Approved Relative',
             'Tribal Registration','Self Certification Affidavit',
             'Early Head Start-Child Care Partnerships')
providertype <- data.frame(provider_code,provider_name)


#create a dataset that has all combinations of county and category for final provider type categories
#this is to account for the fact that some combinations will be 0
providertype_kc_category <- c('Licensed','Licensed School Age','Standard Compliance Certification (SCC) and In-Home','Registered Tribal','Approved Relative')
providertype_kc_categorylist <- as.data.frame(providertype_kc_category)
county_provider_categorylist <- crossing(countymatch,providertype_kc_categorylist) %>%
  select(-c(ndcounty_number))




#create all possible combos of quality categories and counties
quality <- c('Step 1','Step 2','Step 3','Step 4')
qualitylist <- as.data.frame(quality)
county_quality_categorylist <-crossing(countymatch,qualitylist) %>%
  select(-c(ndcounty_number))

#input location ID file for ND
locationids <- read.csv("./Input/ND KC Location IDs.csv")

#input region ID file for ND
regionids <- read.csv("./Input/ND KC Region List.csv")
```

```{r}
#merge in full names for coded variables
child_data_2 <- child_data %>%
  #create a county_code variable using provider license
  mutate(county_code=as.numeric(str_sub(Provider.License..,start=1,end=2))) %>%
  #create a license_code variable using provider license
  mutate(license_code=str_sub(Provider.License..,start=-1,end=-1)) %>%
  left_join(countymatch,by=c('county_code'='ndcounty_number')) %>%
  left_join(providertype, by=c('license_code'='provider_code')) %>%
  rename(providertype_name=provider_name,
         providertype=license_code,
         capacity=`Number.of.Children`) %>%
  
  #do not include if missing provider license (bad data)
  subset(Provider.License..!='') %>%

#create new variable matching type on KC data center
  mutate(providertype_group = case_when(
  providertype=='C' | providertype=='E' | providertype=='F' | providertype=='G' | 
    providertype=='H' | providertype=='M' ~ 'Licensed',
  providertype=='K' ~ 'Licensed School Age',
  providertype=='S' | 
    providertype=='I' ~ 'Standard Compliance Certification (SCC) and In-Home',
  providertype=='R' ~ 'Registered Tribal',
  providertype=='Q' ~ 'Approved Relative',
  providertype=='P' | providertype=='Z' ~ 'Early Head Start-Child Care Partnerships'
))


## DATA QUALITY CHECK
#see if any capacity is missing or zero
temp_capacitycheck <- child_data_2 %>%
  subset(is.na(capacity) | capacity==0)
View(temp_capacitycheck)
```



## -- PROVIDERS AND CAPACITY BY TYPE --##
NOTE: SCHOOL AGE INCLUDED
```{r}
#COUNTY DATA
countysums_providerscapacity <- child_data_2 %>%
  #some rows are not matched to a county, for example reservations. 
  #Remove these from the county data
  drop_na(county) %>%
  group_by(county,providertype_group) %>%
  summarise(providers = n(),
            capacity=sum(capacity),.groups='keep') %>%
  
  #fill in missing county/group combinations
  right_join(county_provider_categorylist,by=c('county'='county','providertype_group'='providertype_kc_category')) %>%
  
  #for the missing combinations that are NA, make them zero
  mutate(providers = replace_na(providers, 0),
         capacity= replace_na(capacity,0)) %>%
  
  #fix kc naming
  rename(location=county,
         category=providertype_group) %>%
  mutate(locationtype='County') 


##STATE DATA
#create a shell of all provider group/location combinations
location_state <- c('North Dakota','North Dakota','North Dakota','North Dakota','North Dakota')
stateshell_providers <- data.frame(location_state,providertype_kc_category)


statesums_providerscapacity <- child_data_2 %>%
  group_by(providertype_group) %>%
  summarise(providers=n(),
            capacity=sum(capacity),.groups='keep') %>%
  mutate(location='North Dakota') %>%
  
  #merge in any missing categories
  right_join(stateshell_providers,by=c('location'='location_state','providertype_group'='providertype_kc_category')) %>%
  
  #for the missing combinations that are NA, make them zero
  mutate(providers = replace_na(providers, 0),
         capacity= replace_na(capacity,0)) %>%
  
  #fix kc naming
  rename(category=providertype_group) %>%
  mutate(locationtype='State') 


##STATE PLANNING REGIONS 
#create a shell of all provider group/location combinations
location_region <- c('Planning Region 1','Planning Region 1','Planning Region 1',
                     'Planning Region 1','Planning Region 1',
                     'Planning Region 2','Planning Region 2','Planning Region 2',
                     'Planning Region 2','Planning Region 2',
                     'Planning Region 3','Planning Region 3','Planning Region 3',
                     'Planning Region 3','Planning Region 3',
                     'Planning Region 4','Planning Region 4','Planning Region 4',
                     'Planning Region 4','Planning Region 4',
                     'Planning Region 5','Planning Region 5','Planning Region 5',
                     'Planning Region 5','Planning Region 5',
                     'Planning Region 6','Planning Region 6','Planning Region 6',
                     'Planning Region 6','Planning Region 6',
                     'Planning Region 7','Planning Region 7','Planning Region 7',
                     'Planning Region 7','Planning Region 7',
                     'Planning Region 8','Planning Region 8','Planning Region 8',
                     'Planning Region 8','Planning Region 8')
regionshell_providers <- data.frame(location_region,providertype_kc_category)


regionsums_providerscapacity <- countysums_providerscapacity %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region,category) %>%
  summarise(providers = sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  
   #merge in any missing categories
  right_join(regionshell_providers,by=c('region'='location_region','category'='providertype_kc_category')) %>%
  
  #for the missing combinations that are NA, make them zero
  mutate(providers = replace_na(providers, 0),
         capacity= replace_na(capacity,0)) %>%
  
  #fix kc naming
  rename(location=region) %>%
  mutate(locationtype='Planning Region') 

#format to database count child care
childcareprovidersbytype <- countysums_providerscapacity %>%
  bind_rows(statesums_providerscapacity) %>%
  bind_rows(regionsums_providerscapacity) %>%
  select(-c(capacity)) %>%
  rename(data=providers) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcareprovidersbytype') 


childcarecapacitybytype <- countysums_providerscapacity %>%
  bind_rows(statesums_providerscapacity) %>%
  bind_rows(regionsums_providerscapacity) %>%
  select(-c(providers)) %>%
  rename(data=capacity) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcarecapacitybytype')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childcareprovidersbytype$locationid))>=1) {
  print(childcareprovidersbytype$location[is.na(childcareprovidersbytype$locationid)])
} else if (sum(is.na(childcareprovidersbytype$locationid))==0) {
  'all locations match'
}


if (sum(is.na(childcarecapacitybytype$locationid))>=1) {
  print(childcarecapacitybytype$location[is.na(childcarecapacitybytype$locationid)])
} else if (sum(is.na(childcarecapacitybytype$locationid))==0) {
  'all locations match'
}

#2. Visually inspect data
View(childcareprovidersbytype)
View(childcarecapacitybytype)
```


```{r}
#add to database providers and capacity of child care
dbWriteTable(con,'northdakota',childcareprovidersbytype,append=TRUE,row.names=FALSE)
dbWriteTable(con,'northdakota',childcarecapacitybytype,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#PROVIDERS OF CHILD CARE

upload_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname='childcareprovidersbytype';")
upload_data <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("./Output/education/",state,"_",year,"_childcareprovidersbytype.csv"),row.names=FALSE)

#CAPACITY OF CHILD CARE

upload_sql2 <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname='childcarecapacitybytype' AND dataformat='Number';")
upload_data2 <- dbGetQuery(con,upload_sql2) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data2,file=paste0("./Output/education/",state,"_",year,"_childcarecapacitybytype.csv"),row.names=FALSE)
```





## -- Child Care Providers and Capacity by QRIS -- ##
NOTE: SCHOOL AGE NOT INCLUDED
```{r}
####CREATE COUNTY SUMMARY COUNT OF CHILD CARE
countysums_quality <- quality_data %>%

  #remove rows not assigned to a county
  rename(category=`Rated QRIS Level Name`,
         location=County,
         capacity=`Total Capacity`) %>%
  mutate(capacity=as.numeric(paste(capacity))) %>%
  group_by(location,category) %>%
  summarise(providers = n(),
            capacity=sum(capacity),.groups='keep') %>%
  ungroup() %>%
  
  #join with shell to make sure all county/quality categories are represented
  right_join(county_quality_categorylist,by=c('location'='county','category'='quality')) %>%

  mutate(providers = replace_na(providers, 0),
         capacity= replace_na(capacity,0)) %>%
  mutate(locationtype='County') 

#create an overall category that shows total participating in QRIS, so we can compute the opposite later
countysums_quality2 <- countysums_quality %>%
  group_by(location,locationtype) %>%
  summarise(providers=sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  mutate(category='Total participating in QRIS')

#combine all groups
countysums_quality_combined <- countysums_quality %>%
  bind_rows(countysums_quality2) %>%
  arrange(location,category)




####CREATE STATEWIDE SUMMARY COUNT OF CHILD CARE
statesums_quality <- countysums_quality_combined %>%
  group_by(category) %>%
  summarise(providers=sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  ungroup() %>%
  mutate(location='North Dakota',
         locationtype='State')



####CREATE STATE PLANNING REGIONS SUMMARY COUNT OF CHILD CARE
#create blank shell in order to capture when 0 occurs
location <- c('Planning Region 1','Planning Region 1','Planning Region 1',
              'Planning Region 1','Planning Region 1',
              'Planning Region 2','Planning Region 2','Planning Region 2',
              'Planning Region 2','Planning Region 2',
              'Planning Region 3','Planning Region 3','Planning Region 3',
              'Planning Region 3','Planning Region 3',
              'Planning Region 4','Planning Region 4','Planning Region 4',
              'Planning Region 4','Planning Region 4',
              'Planning Region 5','Planning Region 5','Planning Region 5',
              'Planning Region 5','Planning Region 5',
              'Planning Region 6','Planning Region 6','Planning Region 6',
              'Planning Region 6','Planning Region 6',
              'Planning Region 7','Planning Region 7','Planning Region 7',
              'Planning Region 7','Planning Region 7',
              'Planning Region 8','Planning Region 8','Planning Region 8',
              'Planning Region 8','Planning Region 8')
category <- c('Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS',
              'Step 1','Step 2','Step 3','Step 4','Total participating in QRIS')
regionshell <- data.frame(location,category)

regionsums_quality <- countysums_quality_combined %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region,category) %>%
  summarise(providers = sum(providers),
            capacity=sum(capacity),.groups='keep') %>%
  ungroup() %>%
  
  right_join(regionshell,by=c('region'='location','category'='category')) %>%
  mutate(providers = replace_na(providers, 0),
         capacity = replace_na(capacity,0)) %>%
  mutate(locationtype='Planning Region') %>%
  rename(location=region)
```

Combine with licensed data from Kids Count database to calculated those not participating
```{r}
licensedproviders_sql <- paste0("SELECT location,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcareprovidersbytype' AND category='Licensed' AND dataformat='Number';")
licensedproviders <- dbGetQuery(con,licensedproviders_sql) %>%
  rename(totallicensedproviders=data) %>%
  mutate(totallicensedproviders=as.numeric(paste(totallicensedproviders)))

licensedcapacity_sql <- paste0("SELECT location,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcarecapacitybytype' AND category='Licensed' AND dataformat='Number';")
licensedcapacity <- dbGetQuery(con,licensedcapacity_sql) %>%
  rename(totallicensedcapacity=data) %>%
  mutate(totallicensedcapacity=as.numeric(paste(totallicensedcapacity)))

#join with full quality data
notparticipatinginquality <- countysums_quality_combined %>%
  bind_rows(statesums_quality) %>%
  bind_rows(regionsums_quality) %>%
  
  #only look at total
  subset(category=='Total participating in QRIS') %>%
  
  left_join(licensedproviders,by=c('location'='location')) %>%
  left_join(licensedcapacity,by=c('location'='location')) %>%
  
  #create column that is those not participating
  mutate(providers_notparticipating=totallicensedproviders-providers,
         capacity_notparticipating=totallicensedcapacity-capacity) %>%
  select(c(location,locationtype,providers_notparticipating,capacity_notparticipating)) %>%
  rename(providers=providers_notparticipating,
         capacity=capacity_notparticipating) %>%
  mutate(category='Not participating in QRIS')
  
  


#COMBINE ALL DATA
quality_data_combined <- countysums_quality_combined %>%
  bind_rows(statesums_quality) %>%
  bind_rows(regionsums_quality) %>%
  bind_rows(notparticipatinginquality) %>%
  subset(category != 'Total participating in QRIS') %>%
  arrange(location,category)


#format to database count child care
childcareprovidersbyquality <- quality_data_combined %>%
  select(-c(capacity)) %>%
  rename(data=providers) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcareprovidersbyquality')


childcarecapacitybyquality <- quality_data_combined %>%
  select(-c(providers)) %>%
  rename(data=capacity) %>%
  mutate(state='North Dakota') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(timeframe=paste0(year)) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='childcarecapacitybyquality')


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childcareprovidersbyquality$locationid))>=1) {
  print(childcareprovidersbyquality$location[is.na(childcareprovidersbyquality$locationid)])
} else if (sum(is.na(childcareprovidersbyquality$locationid))==0) {
  'all locations match'
}


if (sum(is.na(childcarecapacitybyquality$locationid))>=1) {
  print(childcarecapacitybyquality$location[is.na(childcarecapacitybyquality$locationid)])
} else if (sum(is.na(childcarecapacitybyquality$locationid))==0) {
  'all locations match'
}

#2. Visually inspect data
View(childcareprovidersbyquality)
View(childcarecapacitybyquality)
```

```{r}
#add to database count of child care
dbWriteTable(con,'northdakota',childcareprovidersbyquality,append=TRUE,row.names=FALSE)
dbWriteTable(con,'northdakota',childcarecapacitybyquality,append=TRUE,row.names=FALSE)
```


```{r}
#calculate percentages out of all licensed facilities
qualityproviders_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcareprovidersbyquality';")
quality_providers <- dbGetQuery(con,qualityproviders_sql) %>%
  rename(quality_providers=data)
  
qualitycapacity_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcarecapacitybyquality';")
quality_capacity <- dbGetQuery(con,qualitycapacity_sql) %>%
  rename(quality_capacity=data)


licensedproviders_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcareprovidersbytype' AND category='Licensed';")
licensed_providers <- dbGetQuery(con,licensedproviders_sql) %>%
  rename(licensed_providers=data)

licensedcapacity_sql <- paste0("SELECT locationid,location,timeframe,dataformat,data FROM northdakota WHERE timeframe = '",year,"' AND varname='childcarecapacitybytype' AND category='Licensed';")
licensed_capacity <- dbGetQuery(con,licensedcapacity_sql) %>%
  rename(licensed_capacity=data)


quality_percents <- quality_providers %>%
  left_join(quality_capacity) %>%
  left_join(licensed_providers) %>%
  left_join(licensed_capacity) %>%
  arrange(location,category) %>%
  mutate(quality_providers=as.numeric(paste(quality_providers)),
         quality_capacity=as.numeric(paste(quality_capacity)),
         licensed_providers=as.numeric(paste(licensed_providers)),
         licensed_capacity=as.numeric(paste(licensed_capacity))) %>%
  #calculate percentages
  mutate(percent_providers=quality_providers/licensed_providers,
         percent_capacity=quality_capacity/licensed_capacity) %>%
  
  #add in KC variables
  mutate(dataformat='Percent')


quality_percent_providers <- quality_percents %>%
  rename(data=percent_providers) %>%
  select(c(locationid,location,category,timeframe,dataformat,data)) %>%
  mutate(varname='childcareprovidersbyquality')

quality_percent_capacity <- quality_percents %>%
  rename(data=percent_capacity) %>%
  select(c(locationid,location,category,timeframe,dataformat,data)) %>%
  mutate(varname='childcarecapacitybyquality')



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(quality_percent_providers$locationid))>=1) {
  print(quality_percent_providers$location[is.na(quality_percent_providers$locationid)])
} else if (sum(is.na(quality_percent_providers$locationid))==0) {
  'all locations match'
}


if (sum(is.na(quality_percent_capacity$locationid))>=1) {
  print(quality_percent_capacity$location[is.na(quality_percent_capacity$locationid)])
} else if (sum(is.na(quality_percent_capacity$locationid))==0) {
  'all locations match'
}

#2. Visually inspect data
View(quality_percent_providers)
View(quality_percent_capacity)

```



```{r}
#add to database count of child care
dbWriteTable(con,'northdakota',quality_percent_providers,append=TRUE,row.names=FALSE)
dbWriteTable(con,'northdakota',quality_percent_capacity,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#PROVIDERS
upload_data_qualityproviders_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'childcareprovidersbyquality';")
upload_data_qualityproviders <- dbGetQuery(con,upload_data_qualityproviders_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_qualityproviders,file=paste0("./Output/education/",state,"_",year,"_childcareprovidersbyquality.csv"),row.names=FALSE)

#CAPACITY
upload_data_qualitycapacity_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'childcarecapacitybyquality';")
upload_data_qualitycapacity <- dbGetQuery(con,upload_data_qualitycapacity_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_qualitycapacity,file=paste0("./Output/education/",state,"_",year,"_childcarecapacitybyquality.csv"),row.names=FALSE)
```



## OLD CODE USED TO CALCULATE THE PERCENTAGES FOR 1998-2020
```{r}
#FOR  CHILD CARE CAPACITY
#create percentages using ACS data in database
#variable needed for denominator calculation is childrenages0to13parentsinlaborforce
#need denominator variable for two years prior
year_minus1 <- as.character(as.numeric(year)-1)
year_minus5 <- as.character(as.numeric(year)-5)

children_estimates_sql <- paste0("SELECT locationid,data FROM northdakota WHERE timeframe='",year_minus5,"-",year_minus1,"' AND varname='childrenages0to13parentsinlaborforce' AND dataformat='Number' AND age_group='0 to 13';")
children_estimates <- dbGetQuery(con,children_estimates_sql) %>%
  rename(children_estimate=data) %>%
  mutate(children_estimate=as.numeric(paste(children_estimate)))

childcarecapacitybytype2 <- childcarecapacitybytype %>%
  rename(childcare_estimate=data) %>%
  mutate(dataformat='Percent') %>%
  left_join(children_estimates,by='locationid') %>%
  mutate(data=(childcare_estimate/children_estimate)) %>%
  mutate(state='North Dakota') %>%
  select(-children_estimate) %>%
  select(-childcare_estimate)

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childcarecapacitybytype2$locationid))>=1) {
  print(childcarecapacitybytype2$location[is.na(childcarecapacitybytype2$locationid)])
} else if (sum(is.na(childcarecapacitybytype2$locationid))==0) {
  'all locations match'
}

#2. Visually inspect data
View(childcarecapacitybytype2)

```

```{r}
#add to database
dbWriteTable(con,'northdakota',childcarecapacitybytype2,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_data_capacity_sql <- paste0("SELECT locationid,location,category,timeframe,dataformat,data
                     FROM northdakota
                     WHERE timeframe = '",year,"' AND varname = 'childcarecapacitybytype';")
upload_data_capacity <- dbGetQuery(con,upload_data_capacity_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         Category=category,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_capacity,file=paste0("./Output/education/",state,"_",year,"_childcarecapacitybytype.csv"),row.names=FALSE)
```



