---
title: "North Dakota WIC"
author: "Xanna Burg"
date: "July 2020"
output: html_document
---

## Indicator 1: Women, infants, and children (WIC) program participants

**Created by:** Xanna Burg
**Date:** August 2020
**Updated by:**

**Data Source:** North Dakota Department of Health, Division of Nutrition and Physical Activity, WIC Program
**Purpose:** Input the WIC data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage, rate), DataFormat (character: "number"), LocationId (numeric: assigned for KIDS COUNT system)


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(tidycensus)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
library(sas7bdat)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '2025'

statefile <- 'northdakota'
statename <- 'North Dakota'

#input location ID file for ND
locationids <- read.csv("./Input/ND KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)

#input region ID file for ND
regionids <- read.csv("./Input/ND KC Region List.csv")
regionids$county <- as.character(regionids$county)

#input county list file
countylist <- read.csv("./Input/ND County List.csv")
countylist$Location <- as.character(countylist$Location)
```


## STEP 1: IMPORT AND CLEAN THE DATA
```{r}
wicdata_state <- read_xlsx(paste0("./Input/Health/northdakota_",year,"_wicparticipants.xlsx"),range="A4:Z5") %>%
  select(c(`Grand Total`)) %>% 
  mutate(location='North Dakota') %>%
  mutate(locationtype='State') %>%
  rename(data=`Grand Total`) 

wicdata_county <- read_xlsx(paste0("./Input/Health/northdakota_",year,"_wicparticipants.xlsx"),range="A8:Z61") %>%
  select(c(`County`, `Grand Total`)) %>% 
  rename(location=`County`,
         data=`Grand Total`) %>%
  mutate(locationtype='County') %>%
  
  mutate(location=replace(location,location=='Mclean','McLean')) %>%
  mutate(location=replace(location,location=='Lamoure','LaMoure')) %>%
  mutate(location=replace(location,location=='Mchenry','McHenry')) %>%
  mutate(location=replace(location,location=='Mcintosh','McIntosh')) %>%
  mutate(location=replace(location,location=='Mckenzie','McKenzie')) 
  

#region data
wicdata_region <- wicdata_county %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region) %>%
  summarise(data=sum(data,na.rm=TRUE)) %>%
  ungroup %>%
  rename(location=region) %>%
  mutate(location=as.character(paste(location))) %>%
  mutate(locationtype='Planning Region')


#combine all geographies and add in KC variables
wicdata_all <- wicdata_county %>%
  bind_rows(wicdata_state) %>%
  bind_rows(wicdata_region) %>%
  mutate(state='North Dakota') %>%
  mutate(timeframe=year) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='wicparticipants') %>%
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(wicdata_all$locationid))>=1) {
  print(wicdata_all$location[is.na(wicdata_all$locationid)])
} else if (sum(is.na(wicdata_all$locationid))==0) {
  'all locations match'
}

# 2. View data
View(wicdata_all)

```

# STEP 2: COMMIT TO DATABASE
```{r}
#CHECK DATASET NAMED wicdata_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'northdakota',wicdata_all,append=TRUE,row.names=FALSE)

```

# STEP 3: OUTPUT FILE TO UPLOAD
```{r}
#write query from database to get needed format for KC data center

wicparticipants_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM northdakota WHERE timeframe='",year,"' AND varname='wicparticipants';")

upload_data_wicparticipants <- dbGetQuery(con,wicparticipants_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_wicparticipants,file=paste0("./Output/health/",statefile,"_",year,"_wicparticipants.csv"),row.names=FALSE)
```
