---
title: "southdakotaeducation"
author: "Xanna Burg"
date: "March, 2020"
output: html_document
---

## Indicator 1: Children enrolled in special education in public schools
## Indicator 2: Children enrolled in special education by type of impairment
## Indicator 3: Children enrolled in special education by age
## Indicator 4: Children enrolled in special education by grade
## Indicator 5: Public school enrollment grades Pre-K through 12
## Indicator 6: Public school enrollment grades K through 12
## Indicator 7: Public school enrollment for pre-K
## Non-public school enrollment grades pre-K through 12
## Home school enrollment
## Grades 7 to 12 event drop-out rate
## High school completion rate
## Four-year cohort graduation rate
## Average expenditures per student in public schools
## Average ACT composite score
## Eligible recipients of free or reduced-price lunch (state and district only)
## Children enrolled in special education in public schools by race/ethnicity
## High school completion rate by race/ethnicity
## Four-year cohort graduation rate by race/ethnicity

**Created by:** Xanna Burg
**Date:** March 2020
**Updated by:**

**Data Source:** South Dakota Department of Education - Statistical Digest, Child Count, Report Card, and special request
**Purpose:** Input the statewide and district-level data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Race (for three indicators broken down by race; character) Data (numeric: number, percentage), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '23_24'
fullyear <- '2023/24' 
year_prior <- '2022/23' #for dropout data
year_dropout <- '22_23'
calyearformat <- '2024'
statefile <- 'southdakota'
statename <- 'South Dakota'

#input location ID file for SD
locationids <- read.csv("./Input/SD KC Location IDs.csv")
```



**PUBLIC SCHOOL ENROLLMENT GRADES PRE-K THROUGH 12**
**PUBLIC SCHOOL ENROLLMENT GRADES K THROUGH 12**
**PUBLIC SCHOOL ENROLLMENT FOR PRE-K**
#The same code can be used to run both enrollment indicators by changing the objects at the beginning of the first chunk
```{r}
#requested dataset for enrollment and dropouts
xl_data_enroll_public <- paste0("./Input/education/southdakota_",year,"_publicenrollment.xlsx")
df_enrollment <- read_excel(path=xl_data_enroll_public, sheet=paste0('Pubdisgr-',calyearformat),skip=5)


#UPDATE THESE TWO VARIABLES PRIOR TO RUNNING THE CODE
#RUN THE CODE IN THESE CHUNKS THROUGH FOR EACH OF THE INDICATORS
columnname <- 'TOTAL KG-12'
input_dataformat <- 'Number'
variablename <- 'publicschoolk12enrollment'


enrollment_district <- df_enrollment %>%
  rename(location='District Name',
         data=all_of(columnname)) %>%
  select(c(location,data)) %>%
  mutate(locationtype='School District') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename) %>%
  mutate(location=sub("[ ]+(?=[0-9])", 
                      ' School District ', location,perl=TRUE)) %>%
  mutate(location=ifelse(location=='Wolsey-Wessington Sch District 02-6',
                         'Wolsey-Wessington School District 02-6',location)) %>%
    mutate(location=ifelse(location=='Deubrook Area School District 05-6',
                         'Deubrook School District 05-6',location)) %>%
    mutate(location=ifelse(location=='Wagner Community School District 11-4',
                         'Wagner School District 11-4',location)) %>%
    mutate(location=ifelse(location=='Webster Area School District 18-5',
                         'Webster School District 18-5',location)) %>%
    mutate(location=ifelse(location=='Ipswich Public School District 22-6',
                         'Ipswich School District 22-6',location)) %>%
    mutate(location=ifelse(location=='Faulkton Area Schools School District 24-4',
                         'Faulkton School District 24-4',location)) %>%
    mutate(location=ifelse(location=='Miller School District 29-4',
                         'Miller Area School District 29-4',location)) %>%
    mutate(location=ifelse(location=='Kadoka Area School District 35-2',
                         'Kadoka School District 35-2',location)) %>%
    mutate(location=ifelse(location=='Chester Area School District 39-1',
                         'Chester Area District 39-1',location)) %>%
    mutate(location=ifelse(location=='Langford Area School District 45-5',
                         'Langford School District 45-5',location)) %>%
    mutate(location=ifelse(location=='Rapid City Area School District 51-4',
                         'Rapid City School District 51-4',location)) %>%
    mutate(location=ifelse(location=='Viborg-Hurley School District 60-6',
                         'Viborg Hurley 60-6',location)) %>%
    mutate(location=ifelse(location=='Selby Area School District 62-5',
                         'Selby School District 62-5',location)) %>%
    mutate(location=ifelse(location=='Shannon County School District 65-1',
                         'Oglala Lakota County School District 65-1',location)) %>%
    mutate(location=ifelse(location=='Oglala Lakota School District 65-1',
                         'Oglala Lakota County School District 65-1',location)) %>%
    mutate(location=ifelse(location=='Colome Consolidated',
                         'Colome Consolidated School District 59-3',location)) %>%
    mutate(location=ifelse(location=='Groton Area',
                         'Groton Area School District 06-6',location)) %>%
    mutate(location=ifelse(location=='Highmore-Harrold',
                         'Highmore-Harrold School District 34-2',location)) %>%
    mutate(location=ifelse(location=='Hitchcock-Tulare',
                         'Hitchcock-Tulare School District 56-6',location)) %>%
   mutate(location=ifelse(location=='Oldham-Ramona-Rutland School District 39-6',
                         'Oldham-Ramona School District 39-5',location)) %>%
  
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  subset(!is.na(data)) %>%
  subset(location!='State Total')



#calculate the state data from the district data
enrollment_state <- enrollment_district %>%
  group_by(state) %>%
  summarise(data=sum(data),.groups='keep') %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(varname=variablename) %>%
  mutate(locationid=43) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(timeframe=fullyear)

enrollment <- enrollment_district %>%
  bind_rows(enrollment_state)

```

```{r}
#CHECK DATASET NAMED enrollment TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',enrollment,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

enroll_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='",variablename,"';")

upload_data_enroll <-
  dbGetQuery(con,enroll_sql)

upload_data_enroll2 <- upload_data_enroll %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_enroll,file=paste0("./Output/education/southdakota_",year,"_",variablename,".csv"),row.names=FALSE)
```



**PUBLIC SCHOOL ENROLLMENT GRADES PREK THROUGH 12 FOR TRIBAL AREAS**
**NON PUBLIC SCHOOL ENROLLMENT GRADES PREK THROUGH 12 FOR TRIBAL AREAS**
```{r}
#import the tribal schools crosswalk
tribalschools <- read.csv("./Documentation/Indicator Documentation/South Dakota Data Requests/southdakota_education_tribalschoolcrosswalk.csv") %>%
  subset(!is.na(school_specialednumber))


#requested dataset for enrollment and dropouts
xl_data_enroll_nonpublic <- paste0("./Input/education/southdakota_",year,"_nonpublicenrollment.xlsx")
df_enrollment <- read_excel(path=xl_data_enroll_public, sheet=paste0('Pubdisgr-',calyearformat),skip=5)
df_enrollment_nonpublic <- read_excel(path=xl_data_enroll_nonpublic, sheet=paste0('NPdisgr-',calyearformat),skip=6)


#UPDATE THESE TWO VARIABLES PRIOR TO RUNNING THE CODE
#RUN THE CODE IN THESE CHUNKS THROUGH FOR EACH OF THE INDICATORS
columnname <- 'TOTAL PK-12'
input_dataformat <- 'Number'
variablename <- 'prek12enrollment'

#for public schools
enrollment_district <- df_enrollment %>%
  rename(fulllocation_name='District Name',
         districtid='District Number',
         data=all_of(columnname)) %>%
  select(c(fulllocation_name,data,districtid)) %>%
  mutate(districtid=as.numeric(paste(districtid))) %>%
  
  #join with crosswalk to only include public schools matched to a tribal area
  inner_join(tribalschools,by=c('districtid'='school_specialednumber')) %>%
  select(-c(school,fulllocation_name,districtid,category)) %>%
  
  
  mutate(locationtype='Tribal Area') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename) 

#for non public schools
enrollment_district_nonpublic <- df_enrollment_nonpublic %>%
  rename(fulllocation_name='District Name',
         districtid='District Number',
         data=all_of(columnname)) %>%
  select(c(fulllocation_name,data,districtid)) %>%
  mutate(districtid=as.numeric(paste(districtid))) %>%
  
  #join with crosswalk to only include public schools matched to a tribal area
  inner_join(tribalschools,by=c('districtid'='school_specialednumber')) %>%
  select(-c(school,fulllocation_name,districtid,category)) %>%
  
  
  mutate(locationtype='Tribal Area') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename) 

  

#combine public and nonpublic dataframes
enrollment <- enrollment_district %>%
  bind_rows(enrollment_district_nonpublic)

```

```{r}
#CHECK DATASET NAMED enrollment TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',enrollment,append=TRUE,row.names=FALSE)
```



**NON-PUBLIC SCHOOL ENROLLMENT GRADES PRE-K THROUGH 12**
```{r}
state_data <- 15775
#######

#CALCULATE THE STATEWIDE DATA
nonpublic_state <- as.data.frame(state_data) %>%
  rename(data=state_data) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(locationid=43) %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='nonpublicschoolprek12enrollment')

```

```{r}
#CHECK DATASET NAMED nonpublic_state TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',nonpublic_state,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

nonpublic_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='nonpublicschoolprek12enrollment';")

upload_data_nonpublic <-
  dbGetQuery(con,nonpublic_sql)

upload_data_nonpublic2 <- upload_data_nonpublic %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_nonpublic,file=paste0("./Output/education/southdakota_",year,"_nonpublicschoolprek12enrollment.csv"),row.names=FALSE)
```


**DROP OUT RATES FOR PUBLIC AND NON PUBLIC SCHOOLS FOR TRIBAL AREAS**
*NOTE: drop out for public KIDS COUNT data is from statistical digest, data used in tribal data books comes from Excel file for both*
```{r}
#import the tribal schools crosswalk
tribalschools <- read.csv("./Documentation/Indicator Documentation/South Dakota Data Requests/southdakota_education_tribalschoolcrosswalk.csv") %>%
  subset(!is.na(school_specialednumber))


#requested dataset for dropouts
xl_data_enroll <- paste0("./Input/education/southdakota_",year_dropout,"_dropout.xlsx")
df_enrollment <- read_excel(path=xl_data_enroll, sheet='Public District Report - Suppre',skip=5)
df_enrollment_nonpublic <- read_excel(path=xl_data_enroll, sheet='Non-Public, Tribal Report - Sup',skip=5)


#UPDATE THESE TWO VARIABLES PRIOR TO RUNNING THE CODE
#RUN THE CODE IN THESE CHUNKS THROUGH FOR EACH OF THE INDICATORS
columnname <- 'Dropout Rate'
input_dataformat <- 'Percent'
variablename <- 'grades7to12eventdropoutrate'

#for public schools
enrollment_district <- df_enrollment %>%
  rename(fulllocation_name='District Name',
         data=all_of(columnname)) %>%
  
  
  #split district name to get id
  mutate(districtid_part1=str_sub(fulllocation_name,-4,-3)) %>%
  mutate(districtid_part2=str_sub(fulllocation_name,-1,-1)) %>%
  mutate(districtid=paste0(districtid_part1,"00",districtid_part2)) %>%
  select(c(fulllocation_name,data,districtid)) %>%
  mutate(data=as.numeric(paste(data))) %>%
  
  #join with crosswalk to only include public schools matched to a tribal area
  mutate(districtid=as.numeric(paste(districtid))) %>%
  inner_join(tribalschools,by=c('districtid'='school_specialednumber')) %>%
  select(-c(school,fulllocation_name,districtid,category)) %>%
  
  #change percent into decimal
  mutate(data=data/100) %>%
  
  
  mutate(locationtype='Tribal Area') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=year_prior) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename) 

#for non public schools
enrollment_district_nonpublic <- df_enrollment_nonpublic %>%
  rename(fulllocation_name='District Name',
         data=all_of(columnname)) %>%
  mutate(districtid=str_sub(fulllocation_name,-5,-1)) %>%
  select(c(fulllocation_name,data,districtid)) %>%
  mutate(districtid=as.numeric(paste(districtid))) %>%
  mutate(data=as.numeric(paste(data))) %>%
  
  #join with crosswalk to only include public schools matched to a tribal area
  inner_join(tribalschools,by=c('districtid'='school_specialednumber')) %>%
  select(-c(school,fulllocation_name,districtid,category)) %>%
  
  #change percent into decimal
  mutate(data=data/100) %>%
  
  
  mutate(locationtype='Tribal Area') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=year_prior) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename) 

  

#combine public and nonpublic dataframes
enrollment <- enrollment_district %>%
  bind_rows(enrollment_district_nonpublic)

```

```{r}
#CHECK DATASET NAMED enrollment TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',enrollment,append=TRUE,row.names=FALSE)
```



**GRADES 7 TO 12 EVENT DROP-OUT RATE**
**AVERAGE EXPENDITURES PER STUDENT IN PUBLIC SCHOOLS**
**AVERAGE ACT COMPOSITE SCORE**
**HOME SCHOOL ENROLLMENT GRADES K THROUGH 12**
**ELIGIBLE RECIPIENTS OF FREE OR REDUCED PRICE LUNCH - district and school**
#The same code can be used to run all indicators by changing the objects at the beginning of the first chunk
```{r}
#read in data from statistical digest - may need to confirm the number of rows to skip before columns begin
df_digest <- read_excel(path=paste0("./Input/education/southdakota_",year,"_educationstatisticaldigest.xlsx"),skip=3)

####### UPDATE THESE VARIABLES PRIOR TO RUNNING THE CODE
#verify column names prior to assigning, should be: 'Dropout Rate\r\n (%)' and 'Expenditure per ADM' and 'ACT Composite Score' and 'Fall Count of Open Enrolled Students' and 'Free & Reduced Lunch Eligibility Percentage'
#in 20/21 home school varname is 'Fall ADM Count of Home School Students'
#variable.names(df_digest)

columnname <- 'Fall Count of Open Enrolled Students'
input_dataformat <- 'Number'
variablename <- 'homeschoolk12enrollment'
#pull state data from the school profile on the statistical digest
state_data <- 9135
#######


#CALCULATE THE STATEWIDE DATA
digestdata_state <- as.data.frame(state_data) %>%
  rename(data=state_data) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(locationid=43) %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename)

#CALCULATE THE DISTRICT DATA
digestdata_district <- df_digest %>%
  rename(location='District Name',
         data=all_of(columnname)) %>%
  select(c(location,data)) %>%
  mutate(locationtype='School District') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat=input_dataformat) %>%
  mutate(varname=variablename) %>%
  mutate(location=ifelse(location=='Wolsey-Wessington Sch District 02-6',
                         'Wolsey-Wessington School District 02-6',location)) %>%
    mutate(location=ifelse(location=='Deubrook Area School District 05-6',
                         'Deubrook School District 05-6',location)) %>%
    mutate(location=ifelse(location=='Wagner Community School District 11-4',
                         'Wagner School District 11-4',location)) %>%
    mutate(location=ifelse(location=='Webster Area School District 18-5',
                         'Webster School District 18-5',location)) %>%
    mutate(location=ifelse(location=='Ipswich Public School District 22-6',
                         'Ipswich School District 22-6',location)) %>%
    mutate(location=ifelse(location=='Faulkton Area Schools District 24-4',
                         'Faulkton School District 24-4',location)) %>%
    mutate(location=ifelse(location=='Miller School District 29-4',
                         'Miller Area School District 29-4',location)) %>%
    mutate(location=ifelse(location=='Kadoka Area School District 35-2',
                         'Kadoka School District 35-2',location)) %>%
    mutate(location=ifelse(location=='Chester Area School District 39-1',
                         'Chester Area District 39-1',location)) %>%
    mutate(location=ifelse(location=='Langford Area School District 45-5',
                         'Langford School District 45-5',location)) %>%
    mutate(location=ifelse(location=='Rapid City Area School District 51-4',
                         'Rapid City School District 51-4',location)) %>%
    mutate(location=ifelse(location=='Viborg-Hurley School District 60-6',
                         'Viborg Hurley 60-6',location)) %>%
    mutate(location=ifelse(location=='Selby Area School District 62-5',
                         'Selby School District 62-5',location)) %>%
  mutate(location=ifelse(location=='Shannon County School District 65-1',
                         'Oglala Lakota County School District 65-1',location)) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  subset(!is.na(locationid)) #%>%
  #mutate(data=as.numeric(paste(data)))
  #mutate(data=(as.numeric(str_remove(data,"%")))/100) #for frl

#COMBINE STATE AND DISTRICT DATA
digestdata <- digestdata_state %>%
  bind_rows(digestdata_district)

#DATA CHECKS
# 1. Print name of location that has a mismatched location ID
if (sum(is.na(digestdata$locationid))>=1) {
  print(digestdata$location[is.na(digestdata$locationid)])
} else if (sum(is.na(digestdata$locationid))==0) {
  'all locations match'
}

# 2. View data
View(digestdata)
```

```{r}
#CHECK DATASET NAMED digestdata TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',digestdata,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center


#for FRL only, which has county data for same year represented in statistical digest
#digest_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='",variablename,"' AND (locationtype='School District' OR locationtype='State');")

digest_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='",variablename,"' AND locationtype <> 'Tribal Area';")

upload_data_digest <- dbGetQuery(con,digest_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_digest,file=paste0("./Output/education/southdakota_",year,"_",variablename,".csv"),row.names=FALSE)
```



```{r}
#READ IN THE SPECIAL EDUCATION DATA

#UPDATE THE RANGES TO INCLUDE THE NEW COLUMN

#special education state data
xl_data <- paste0("./Input/education/southdakota_",year,"_specialeducationstate.xlsx")
df_specialed_impairment <- read_excel(path=xl_data, range="A4:Z23")
df_specialed_grade <- read_excel(path=xl_data, range="A26:Z42")
df_specialed_age <- read_excel(path=xl_data, range="A67:Z87")
df_specialed_race <- read_excel(path=xl_data, range="A49:Z57")

#special education district data
xl_data_specialed_district <- paste0("./Input/education/southdakota_",year,"_specialeducationdistrict.xlsx")
df_specialed_district <- read_excel(path=xl_data_specialed_district,skip=3)
```


**CHILDREN ENROLLED IN SPECIAL EDUCATION IN PUBLIC SCHOOLS**
```{r}
specialed_count <- df_specialed_district %>%
  subset(!is.na(`District Number`)) %>%
  select(c('District Name','2023 TOTAL Child Count (ages 3-21)')) %>%
  rename(location='District Name',
         data='2023 TOTAL Child Count (ages 3-21)') %>%
  subset(!is.na(location) & location != 'TOTAL' & 
           location != 'contact Susan Woodmansey,') %>%
  mutate(locationtype='School District') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducation') %>%
  mutate(data=as.numeric(paste(data))) %>%
  mutate(location=sub("[ ]+(?=[0-9])", 
                      ' School District ', location,perl=TRUE)) %>%
  mutate(location=ifelse(location=='Chester Area School District 39-1', 
        'Chester Area District 39-1',location)) %>%
  mutate(location=ifelse(location=='Deubrook Area School District 05-6', 
        'Deubrook School District 05-6',location)) %>%
  mutate(location=ifelse(location=='Faulkton Area Schools School District 24-4','Faulkton School District 24-4',location)) %>%
  mutate(location=ifelse(location=='Ipswich Public School District 22-6', 
        'Ipswich School District 22-6',location)) %>%
  mutate(location=ifelse(location=='Kadoka Area School District 35-2',
        'Kadoka School District 35-2',location)) %>%
  mutate(location=ifelse(location=='Langford Area School District 45-5',
        'Langford School District 45-5',location)) %>%
  mutate(location=ifelse(location=='Miller School District 29-4', 
        'Miller Area School District 29-4',location)) %>%
  mutate(location=ifelse(location=='Oglala Lakota School District 65-1', 
        'Oglala Lakota County School District 65-1',location)) %>%
  mutate(location=ifelse(location=='Rapid City Area School District 51-4',
        'Rapid City School District 51-4',location)) %>%
  mutate(location=ifelse(location=='Selby Area School District 62-5',
        'Selby School District 62-5',location)) %>%
  mutate(location=ifelse(location=='Viborg-Hurley School District 60-6',
        'Viborg Hurley 60-6',location)) %>%
  mutate(location=ifelse(location=='Wagner Community School District 11-4',
        'Wagner School District 11-4',location)) %>%
  mutate(location=ifelse(location=='Webster Area School District 18-5',
        'Webster School District 18-5',location)) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

specialed_percent <- df_specialed_district %>%
  subset(!is.na(`District Number`)) %>%
  select(c('District Name','Child Count as % of Fall PK-12 Enrollment')) %>%
  rename(location='District Name',
         data='Child Count as % of Fall PK-12 Enrollment') %>%
  subset(!is.na(location) & location != 'TOTAL' & 
           location != 'contact Susan Woodmansey,') %>%
  mutate(locationtype='School District') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducation') %>%
  mutate(location=sub("[ ]+(?=[0-9])", 
                      ' School District ', location,perl=TRUE)) %>%
  mutate(location=ifelse(location=='Chester Area School District 39-1', 
        'Chester Area District 39-1',location)) %>%
  mutate(location=ifelse(location=='Deubrook Area School District 05-6', 
        'Deubrook School District 05-6',location)) %>%
  mutate(location=ifelse(location=='Faulkton Area Schools School District 24-4','Faulkton School District 24-4',location)) %>%
  mutate(location=ifelse(location=='Ipswich Public School District 22-6', 
        'Ipswich School District 22-6',location)) %>%
  mutate(location=ifelse(location=='Kadoka Area School District 35-2',
        'Kadoka School District 35-2',location)) %>%
  mutate(location=ifelse(location=='Langford Area School District 45-5',
        'Langford School District 45-5',location)) %>%
  mutate(location=ifelse(location=='Miller School District 29-4', 
        'Miller Area School District 29-4',location)) %>%
  mutate(location=ifelse(location=='Oglala Lakota School District 65-1', 
        'Oglala Lakota County School District 65-1',location)) %>%
  mutate(location=ifelse(location=='Rapid City Area School District 51-4',
        'Rapid City School District 51-4',location)) %>%
  mutate(location=ifelse(location=='Selby Area School District 62-5',
        'Selby School District 62-5',location)) %>%
  mutate(location=ifelse(location=='Viborg-Hurley School District 60-6',
        'Viborg Hurley 60-6',location)) %>%
  mutate(location=ifelse(location=='Wagner Community School District 11-4',
        'Wagner School District 11-4',location)) %>%
  mutate(location=ifelse(location=='Webster Area School District 18-5',
        'Webster School District 18-5',location)) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


specialed_count_state <- df_specialed_district %>%
  rename(data='2023 TOTAL Child Count (ages 3-21)',
         location='District Number') %>%
  subset(location=='TOTAL') %>%
  select(c(data))  %>%
  mutate(location='South Dakota') %>%
  mutate(state='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(locationid=43) %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducation') %>%
  mutate(data=as.numeric(paste(data)))

specialed_percent_state <- df_specialed_district %>%
  rename(data='Child Count as % of Fall PK-12 Enrollment',
         location='District Number') %>%
  subset(location=='TOTAL') %>%
  select(c(data))  %>%
  mutate(location='South Dakota') %>%
  mutate(state='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(locationid=43) %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducation')

  
upload_specialeducation_district <- specialed_count %>%
  bind_rows(specialed_percent) %>%
  bind_rows(specialed_count_state) %>%
  bind_rows(specialed_percent_state)


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(upload_specialeducation_district$locationid))>=1) {
  print(upload_specialeducation_district$location[is.na(upload_specialeducation_district$locationid)])
} else if (sum(is.na(upload_specialeducation_district$locationid))==0) {
  'all locations match'
}

# 2. Visually inspect output data
View(upload_specialeducation_district)
```

```{r}
#CHECK DATASET NAMED upload_specialeducation_district TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',upload_specialeducation_district,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

specialed_districtenrollment_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND varname='totalspecialeducation';")

upload_data_specialed_district <-
  dbGetQuery(con,specialed_districtenrollment_sql)

upload_data_specialed_district2 <- upload_data_specialed_district %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_specialed_district2,file=paste0("./Output/education/",statefile,"_",year,"_totalspecialeducation.csv"),row.names=FALSE)
```


**CHILDREN ENROLLED IN SPECIAL EDUCATION IN PUBLIC SCHOOLS**
**PUBLIC SCHOOLS MATCHED TO TRIBAL AREA**
```{r}
#import the tribal schools crosswalk
tribalschools <- read.csv("./Documentation/Indicator Documentation/South Dakota Data Requests/southdakota_education_tribalschoolcrosswalk.csv") %>%
  subset(!is.na(school_specialednumber))
tribalschools$school_specialednumber <- as.character(tribalschools$school_specialednumber)


specialed_count <- df_specialed_district %>%
  subset(!is.na(`District Number`)) %>%
  select(c('District Name','2023 TOTAL Child Count (ages 3-21)','District Number')) %>%
  rename(fulllocation_name='District Name',
         districtid='District Number',
         data='2023 TOTAL Child Count (ages 3-21)') %>%
  
  #match to tribal area using crosswalk
  inner_join(tribalschools,by=c('districtid'='school_specialednumber')) %>%
  select(-c(fulllocation_name,districtid,school,category)) %>%
  
  
  mutate(locationtype='Tribal Area') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducation') %>%
  mutate(data=as.numeric(paste(data))) 




specialed_percent <- df_specialed_district %>%
  subset(!is.na(`District Number`)) %>%
  select(c('District Name','Child Count as % of Fall PK-12 Enrollment','District Number')) %>%
  rename(fulllocation_name='District Name',
         districtid='District Number',
         data='Child Count as % of Fall PK-12 Enrollment') %>%
  
  #match to tribal area using crosswalk
  inner_join(tribalschools,by=c('districtid'='school_specialednumber')) %>%
  select(-c(fulllocation_name,districtid,school,category)) %>%
  
  mutate(locationtype='Tribal Area') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducation') 


  
upload_specialeducation_district <- specialed_count %>%
  bind_rows(specialed_percent) 

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Visually inspect output data
View(upload_specialeducation_district)

```

```{r}
#CHECK DATASET NAMED upload_specialeducation_district TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',upload_specialeducation_district,append=TRUE,row.names=FALSE)
```




**CHILDREN ENROLLED IN SPECIAL EDUCATION IN PUBLIC SCHOOLS BY TYPE OF IMPAIRMENT**
```{r}
fullyear_file <- '2023'
year_variable <- paste0("December ",fullyear_file)


specialed_impairment_count <- df_specialed_impairment %>%
  rename(data=all_of(year_variable),
         category_original='Disabling  Condition') %>%
  select(c(data,category_original)) %>%
  subset(!is.na(category_original) & 
           category_original != 'Total Ages 3-5' &
           category_original != 'Total Ages 6-21' & 
           category_original != 'TOTAL CHILD COUNT (Ages 3-21)') %>%
  mutate(category=case_when(
    category_original=='Autism Spectrum Disorder (560)' ~ 'Autism Spectrum Disorder',
    category_original=='Cognitive Disability (510)' ~ 'Cognitive Disability',
    category_original=='Deaf (545)' ~ 'Deafness',
    category_original=='Emotional Behavioral Disability (505)' ~ 
      'Emotionally Disturbed',
    category_original=='Hearing Loss (515)' ~ 'Hearing Loss',
    category_original=='Multiple Disabilities (530)' ~ 
      'Multiple Disabilities',
    category_original=='Orthopedic Impairments (535)' ~ 
      'Orthopedic Impairments',
    category_original=='Other Health Impaired (555)' ~ 
      'Other Health Impairments',
    category_original=='Developmental Delay (570)' ~ 
      'Developmental Delay (Preschool Ages 3-5)',
    category_original=='Prolonged Assistance (Ages 0-2)' ~ 
      'Prolonged Assistance (Ages 0-2)',
    category_original=='Specific Learning Disabled (525)' ~ 
      'Specific Learning Disabled',
    category_original=='Speech/Language Impairments (550)' ~ 
      'Speech/Language Disorder',
    category_original=='Traumatic Brain Injury (565)' ~ 
      'Traumatic Brain Injury',
    category_original=='Vision Loss (540)' ~ 'Vision Loss')) %>%
  select(-c(category_original)) %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducationbyimpairment') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


specialed_impairment_percent <- df_specialed_impairment %>%
  rename(data_number=all_of(year_variable),
         category_original='Disabling  Condition') %>%
  select(c(data_number,category_original)) %>%
  subset(!is.na(category_original) & 
           category_original != 'Total Ages 3-5' &
           category_original != 'Total Ages 6-21') %>%
   mutate(total=data_number[category_original=='TOTAL CHILD COUNT (Ages 3-21)']) %>%
  subset(category_original != 'TOTAL CHILD COUNT (Ages 3-21)') %>%
  mutate(category=case_when(
    category_original=='Autism Spectrum Disorder (560)' ~ 'Autism Spectrum Disorder',
    category_original=='Cognitive Disability (510)' ~ 'Cognitive Disability',
    category_original=='Deaf (545)' ~ 'Deafness',
    category_original=='Emotional Behavioral Disability (505)' ~ 
      'Emotionally Disturbed',
    category_original=='Hearing Loss (515)' ~ 'Hearing Loss',
    category_original=='Multiple Disabilities (530)' ~ 
      'Multiple Disabilities',
    category_original=='Orthopedic Impairments (535)' ~ 
      'Orthopedic Impairments',
    category_original=='Other Health Impaired (555)' ~ 
      'Other Health Impairments',
    category_original=='Developmental Delay (570)' ~ 
      'Developmental Delay (Preschool Ages 3-5)',
    category_original=='Prolonged Assistance (Ages 0-2)' ~ 
      'Prolonged Assistance (Ages 0-2)',
    category_original=='Specific Learning Disabled (525)' ~ 
      'Specific Learning Disabled',
    category_original=='Speech/Language Impairments (550)' ~ 
      'Speech/Language Disorder',
    category_original=='Traumatic Brain Injury (565)' ~ 
      'Traumatic Brain Injury',
    category_original=='Vision Loss (540)' ~ 'Vision Loss')) %>%
  mutate(data=data_number/total) %>%
  select(-c(category_original,data_number,total)) %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducationbyimpairment') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


upload_specialeducation <- specialed_impairment_count %>%
  bind_rows(specialed_impairment_percent)

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Visually inspect output data
View(upload_specialeducation)
```

```{r}
#CHECK DATASET NAMED upload_specialeducation TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',upload_specialeducation,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

specialed_impairment_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data, category FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='totalspecialeducationbyimpairment';")

upload_data_specialed_impairment <- dbGetQuery(con,specialed_impairment_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Category=category)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_specialed_impairment,file=paste0("./Output/education/",statefile,"_",year,"_totalspecialeducationbyimpairment.csv"),row.names=FALSE)
```


**CHILDREN ENROLLED IN SPECIAL EDUCATION IN PUBLIC SCHOOLS BY GRADE LEVEL**
```{r}
#clean up the data

specialed_grade_count <- df_specialed_grade %>%
  rename(data=all_of(fullyear_file),
         category_original=Grade) %>%
  select(c(data,category_original)) %>%
  subset(!is.na(category_original) & 
           category_original != 'TOTAL') %>%
  mutate(category=case_when(
    category_original=='Early Childhood' ~ 'EC',
    category_original=='Eighth Grade' ~ '8th Grade',
    category_original=='Eleventh Grade' ~ '11th Grade',
    category_original=='Fifth Grade' ~ '5th Grade',
    category_original=='First Grade' ~ '1st Grade',
    category_original=='Fourth Grade' ~ '4th Grade',
    category_original=='Kindergarten' ~ 'K',
    category_original=='Ninth Grade' ~ '9th Grade',
    category_original=='Pre-Kindergarten' ~ 'PK',
    category_original=='Second Grade' ~ '2nd Grade',
    category_original=='Seventh Grade' ~ '7th Grade',
    category_original=='Sixth Grade' ~ '6th Grade',
    category_original=='Tenth Grade' ~ '10th Grade',
    category_original=='Third Grade' ~ '3rd Grade',
    category_original=='Twelfth Grade' ~ '12th Grade')) %>%
  select(-c(category_original)) %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducationbygrade') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


specialed_grade_percent <- df_specialed_grade %>%
  rename(data_number=all_of(fullyear_file),
         category_original='Grade') %>%
  select(c(data_number,category_original)) %>%
  subset(!is.na(category_original)) %>%
   mutate(total=data_number[category_original=='TOTAL']) %>%
  subset(category_original != 'TOTAL') %>%
  mutate(category=case_when(
    category_original=='Early Childhood' ~ 'EC',
    category_original=='Eighth Grade' ~ '8th Grade',
    category_original=='Eleventh Grade' ~ '11th Grade',
    category_original=='Fifth Grade' ~ '5th Grade',
    category_original=='First Grade' ~ '1st Grade',
    category_original=='Fourth Grade' ~ '4th Grade',
    category_original=='Kindergarten' ~ 'K',
    category_original=='Ninth Grade' ~ '9th Grade',
    category_original=='Pre-Kindergarten' ~ 'PK',
    category_original=='Second Grade' ~ '2nd Grade',
    category_original=='Seventh Grade' ~ '7th Grade',
    category_original=='Sixth Grade' ~ '6th Grade',
    category_original=='Tenth Grade' ~ '10th Grade',
    category_original=='Third Grade' ~ '3rd Grade',
    category_original=='Twelfth Grade' ~ '12th Grade')) %>%
  mutate(data=data_number/total) %>%
  select(-c(category_original,data_number,total)) %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducationbygrade') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


upload_specialeducation_grade <- specialed_grade_count %>%
  bind_rows(specialed_grade_percent)

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Visually inspect output data
View(upload_specialeducation_grade)
```

```{r}
#CHECK DATASET NAMED upload_specialeducation_grade TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',upload_specialeducation_grade,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

specialed_grade_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data, category FROM southdakota WHERE timeframe='",fullyear,"' AND   varname='totalspecialeducationbygrade';")

upload_data_specialed_grade <- dbGetQuery(con,specialed_grade_sql)

upload_data_specialed_grade2 <- upload_data_specialed_grade %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         'Grade Level'=category)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_specialed_grade2,file=paste0("./Output/education/",statefile,"_",year,"_totalspecialeducationbygrade.csv"),row.names=FALSE)
```



**CHILDREN ENROLLED IN SPECIAL EDUCATION IN PUBLIC SCHOOLS BY AGE**
```{r}
#clean up the data

specialed_age_count <- df_specialed_age %>%
  rename(data=all_of(fullyear_file),
         category=Age) %>%
  select(c(data,category)) %>%
  subset(!is.na(category) & 
           category != 'TOTAL') %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducationbyage') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


specialed_age_percent <- df_specialed_age %>%
  rename(data_number=all_of(fullyear_file),
         category=Age) %>%
  select(c(data_number,category)) %>%
  subset(!is.na(category)) %>%
   mutate(total=data_number[category=='TOTAL']) %>%
  subset(category != 'TOTAL') %>%
  mutate(data=data_number/total) %>%
  select(-c(data_number,total)) %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducationbyage') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


upload_specialeducation_age <- specialed_age_count %>%
  bind_rows(specialed_age_percent) %>%
  rename(age_group=category)

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Visually inspect output data
View(upload_specialeducation_age)
```

```{r}
#CHECK DATASET NAMED upload_specialeducation_age TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',upload_specialeducation_age,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

specialed_age_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data, age_group FROM southdakota WHERE timeframe='",fullyear,"' AND varname='totalspecialeducationbyage';")

upload_data_specialed_age <- dbGetQuery(con,specialed_age_sql)

upload_data_specialed_age2 <- upload_data_specialed_age %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         'Age group'=age_group)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_specialed_age2,file=paste0("./Output/education/",statefile,"_",year,"_totalspecialeducationbyage.csv"),row.names=FALSE)
```



**CHILDREN ENROLLED IN SPECIAL EDUCATION IN PUBLIC SCHOOLS BY RACE**
```{r}
#clean up the data

specialed_race_count <- df_specialed_race %>%
  rename(data=all_of(fullyear_file),
         category_original=Ethnicity) %>%
  select(c(data,category_original)) %>%
  subset(!is.na(category_original) & 
           category_original != 'TOTAL') %>%
  mutate(category=case_when(
    category_original=='Asian' ~ 'Asian',
    category_original=='Black' ~ 'Black or African American',
    category_original=='Hispanic' ~ 'Hispanic or Latino',
    category_original=='Multiple Race' ~ 'Multi-Race',
    category_original=='Native American' ~ 'Native American',
    category_original=='Pacific Islander' ~ 'Pacific Islander',
    category_original=='White' ~ 'White')) %>%
  subset(!is.na(category)) %>%
  select(-c(category_original)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalspecialeducationbyrace') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


specialed_race_percent <- df_specialed_race %>%
  rename(data_number=all_of(fullyear_file),
         category_original=Ethnicity) %>%
  select(c(data_number,category_original)) %>%
  subset(!is.na(category_original)) %>%
   mutate(total=data_number[category_original=='TOTAL']) %>%
  subset(category_original != 'TOTAL') %>%
  mutate(data=data_number/total) %>%
  mutate(category=case_when(
    category_original=='Asian' ~ 'Asian',
    category_original=='Black' ~ 'Black or African American',
    category_original=='Hispanic' ~ 'Hispanic or Latino',
    category_original=='Multiple Race' ~ 'Multi-Race',
    category_original=='Native American' ~ 'Native American',
    category_original=='Pacific Islander' ~ 'Pacific Islander',
    category_original=='White' ~ 'White')) %>%
  select(-c(data_number,total,category_original)) %>%
  subset(!is.na(category)) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='totalspecialeducationbyrace') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


upload_specialeducation_race <- specialed_race_count %>%
  bind_rows(specialed_race_percent) %>%
  rename(race=category)

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Visually inspect output data
View(upload_specialeducation_race)
```

```{r}
#CHECK DATASET NAMED upload_specialeducation_race TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',upload_specialeducation_race,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

specialed_race_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data, race FROM southdakota WHERE timeframe='",fullyear,"' AND varname='totalspecialeducationbyrace';")

upload_data_specialed_race <- dbGetQuery(con,specialed_race_sql)

upload_data_specialed_race2 <- upload_data_specialed_race %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_specialed_race2,file=paste0("./Output/education/",statefile,"_",year,"_totalspecialeducationbyrace.csv"),row.names=FALSE)
```



**HIGH SCHOOL COMPLETION RATE**
**FOUR-YEAR COHORT GRADUATION RATE**
#The same code can be used to run both indicators
```{r}
#read in data from report card data - may need to confirm the number of rows to skip before columns begin
df_hsdata <- read_excel(path=paste0("./Input/education/southdakota_",year,"_educationhighschooldata.xlsx"),skip=2)

####### UPDATE THESE VARIABLES PRIOR TO RUNNING THE CODE
#verify column names prior to assigning
#The dataset isn't set up very nice in that there are three rows to uniquely identify each variable. Before using this data, need to verify the order of the columns as the order is how it is matched to the correct group.

year1 <- substr(fullyear,1,4)
year2 <- as.numeric(paste(year1))+1

#rename columns
df_hsdata2 <- df_hsdata %>%
  rename(completion_percentage=Percentage...10,
         graduation_percentage=Percentage...16) %>%
  subset(School=="All Schools") %>%
  subset(`Academic Year`==paste0(year1,"-",year2)) %>%
  select(District, completion_percentage,graduation_percentage)


#CALCULATE THE STATEWIDE AND DISTRICT DATA FOR BOTH VARIABLES
hsdata <- df_hsdata2 %>%
  mutate(location=case_when(
    District=="All Districts" ~ "South Dakota",
    District!="All Districts" ~ District)) %>%
  mutate(locationtype=case_when(
    location=='South Dakota' ~ 'State',
    location!='South Dakota' ~ 'County')) %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  select(-c(District)) %>%
  

  #add in school district to the locations for matching to KIDS COUNT
  mutate(location=sub("[ ]+(?=[0-9])", 
                      ' School District ', location,perl=TRUE)) %>%
  mutate(location=ifelse(location=='Chester Area School District 39-1',
                         'Chester Area District 39-1',location)) %>%
  mutate(location=ifelse(location=='Deubrook Area School District 05-6',
                         'Deubrook School District 05-6',location)) %>%
  mutate(location=ifelse(location=='Faulkton Area Schools School District 24-4',
                         'Faulkton School District 24-4',location)) %>%
  mutate(location=ifelse(location=='Ipswich Public School District 22-6',
                         'Ipswich School District 22-6',location)) %>%
  mutate(location=ifelse(location=='Webster Area School District 18-5',
                         'Webster School District 18-5',location)) %>%
  mutate(location=ifelse(location=='Kadoka Area School District 35-2',
                         'Kadoka School District 35-2',location)) %>%
  mutate(location=ifelse(location=='Langford Area School District 45-5',
                         'Langford School District 45-5',location)) %>%
  mutate(location=ifelse(location=='Miller School District 29-4',
                         'Miller Area School District 29-4',location)) %>%
  mutate(location=ifelse(location=='Selby Area School District 62-5',
                         'Selby School District 62-5',location)) %>%
  mutate(location=ifelse(location=='Rapid City Area School District 51-4',
                         'Rapid City School District 51-4',location)) %>%
  mutate(location=ifelse(location=='Viborg-Hurley School District 60-6',
                         'Viborg Hurley 60-6',location)) %>%
  mutate(location=ifelse(location=='Wagner Community School District 11-4',
                         'Wagner School District 11-4',location)) %>%
  
  #add in locationid
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)



#CREATE THE HS COMPLETION DATASET
hs_completion <- hsdata %>%
  select(-c(graduation_percentage)) %>%
  rename(data=completion_percentage) %>%
  
  #assign asterisk to NA
  mutate(data=replace(data,data=='*',NA)) %>%
  
  mutate(varname='highschoolcompletion')


#CREATE THE HS GRADUATION DATASET
hs_graduation <- hsdata %>%
  select(-c(completion_percentage)) %>%
  rename(data=graduation_percentage) %>%
  
  #assign asterisk to NA
  mutate(data=replace(data,data=='*',NA)) %>%
  
  mutate(varname='graduationrate4cohort')

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(hs_completion$locationid))>=1) {
  print(hs_completion$location[is.na(hs_completion$locationid)])
} else if (sum(is.na(hs_completion$locationid))==0) {
  'all locations match'
}

# 2. Visually inspect output data
View(hs_completion)


# 1. Print name of location that has a mismatched location ID
if (sum(is.na(hs_graduation$locationid))>=1) {
  print(hs_graduation$location[is.na(hs_graduation$locationid)])
} else if (sum(is.na(hs_graduation$locationid))==0) {
  'all locations match'
}

# 2. Visually inspect output data
View(hs_graduation)

```

```{r}
#CHECK DATASET NAMED hs_completion TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',hs_completion,append=TRUE,row.names=FALSE)


#CHECK DATASET NAMED hs_graduation TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',hs_graduation,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#code will pull data for both HS completion and graduation

completion_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='highschoolcompletion';")

graduation_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='graduationrate4cohort';")

upload_data_completion <- dbGetQuery(con,completion_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

upload_data_graduation <- dbGetQuery(con,graduation_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_completion,file=paste0("./Output/education/southdakota_",year,"_highschoolcompletion.csv"),row.names=FALSE)

write.csv(upload_data_graduation,file=paste0("./Output/education/southdakota_",year,"_graduationrate4cohort.csv"),row.names=FALSE)
```


**HIGH SCHOOL COMPLETION RATE BY RACE**
**FOUR-YEAR COHORT GRADUATION RATE BY RACE**
#The same code can be used to run both indicators
```{r}
#MANUALLY INPUT THE RACE DATA USING THE REPORT CARD
race <- c('American Indian/Alaska Native','Asian','Black or African American','Hispanic or Latino','White','All Races','Two or more races')
graduation <- c(.4635,.8452,.8239,.7579,.9072,.8405,.7930)
completion <- c(.6835,.9677,.8825,.7992,.9565,.9116,.8827)

df_graduation_race <- data.frame(race,graduation) %>%
  rename(data=graduation) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(locationid=43) %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='graduationrate4cohortbyrace')

df_completion_race <- data.frame(race,completion) %>%
  rename(data=completion) %>%
  mutate(location='South Dakota') %>%
  mutate(locationtype='State') %>%
  mutate(locationid=43) %>%
  mutate(state='South Dakota') %>%
  mutate(timeframe=fullyear) %>%
  mutate(dataformat='Percent') %>%
  mutate(varname='highschoolcompletionbyrace')

```

```{r}
#CHECK DATASET NAMED df_graduation_race TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',df_graduation_race,append=TRUE,row.names=FALSE)


#CHECK DATASET NAMED df_completion_race TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#add to database
dbWriteTable(con,'southdakota',df_completion_race,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center
#code will pull data for both HS completion and graduation

completion_sql2 <- paste0("SELECT locationid, location, timeframe, dataformat, data, race FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='highschoolcompletionbyrace';")

graduation_sql2 <- paste0("SELECT locationid, location, timeframe, dataformat, data, race FROM southdakota WHERE timeframe='",fullyear,"' AND  varname='graduationrate4cohortbyrace';")

upload_data_completion2 <- dbGetQuery(con,completion_sql2) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)

upload_data_graduation2 <- dbGetQuery(con,graduation_sql2) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_completion2,file=paste0("./Output/education/southdakota_",year,"_highschoolcompletionbyrace.csv"),row.names=FALSE)

write.csv(upload_data_graduation2,file=paste0("./Output/education/southdakota_",year,"_graduationrate4cohortbyrace.csv"),row.names=FALSE)

```


