---
title: "South Dakota FRL by County"
author: "Xanna Burg"
date: "June 2020"
output: html_document
---

## Indicator 1: Eligible recipients of free or reduced-price lunch (county only)
## This code also uses the same files to calculate FRL for tribal areas, where applicable in South Dakota 

**Created by:** Xanna Burg
**Date:** June 2020
**Updated by:**

**Data Source:** South Dakota Department of Education, special request from Child and Adult Nutrition Services
**Purpose:** Input the county-level data

**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: percentage), DataFormat (character: "number" or "percent" or "rate"), LocationId (numeric: assigned for KIDS COUNT system)

```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```

```{r}
####UPDATE to reflect the current year data working with
year <- '2023'
fullyear <- '2023/24' 
statefile <- 'southdakota'
statename <- 'South Dakota'

#input location ID file for SD
locationids <- read.csv("./Input/SD KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)

#input file with all SD counties
countylist <- read.csv("./Input/SD County List.csv")
countylist$county <- as.character(countylist$county)
```


## ################# ##
## COUNTY-LEVEL DATA ##
## ################# ##

## STEP 1: IMPORT AND CLEAN DATA
```{r}
frl_data <- read_excel(paste0("./Input/education/southdakota_",year,"_frl.xls"),sheet='Sort by County- PUBLIC ONLY') %>%
  
  #fix data so only the totals are included for each county
  replace_with_na(replace=list(County='County Subtotal')) %>%
  fill(County) %>%
  rename(school='School/Agency Name',
         studentenrollment='Student Enrollment') %>%
  subset(is.na(school)) %>%
  
  #fix the capitalization in the counties
  mutate(location=tolower(County)) %>%
  mutate(location=tools::toTitleCase(location)) %>%
  mutate(location=replace(location, 
                          location=='Mccook', 
                          'McCook')) %>%
    mutate(location=replace(location, 
                          location=='Mcpherson', 
                          'McPherson')) %>%
  mutate(location=replace(location, 
                          location=='Oglala Lakota County', 
                          'Oglala Lakota')) %>%
  
  
  #keep only needed variables
  rename(data='Free & Reduced-Price Eligibility Percentage') %>%
  subset(is.na(`Claim Date`)) %>%
  select(c(location,data,studentenrollment)) %>%
  
  #remove state totals
  subset(location != 'State Totals') %>%
  
  
  #assign NA value to those with **
  mutate(studentenrollment=as.numeric(paste(studentenrollment))) %>%
  mutate(data=replace(data,data=='**' & studentenrollment>20,'>=90%')) %>%
  mutate(data=replace(data,data=='**' & studentenrollment<=20,NA)) %>%
  select(-c(studentenrollment)) %>%
  
  #add in missing counties
  full_join(countylist,by=c('location'='county')) %>%
  
  
  #add in Kids Count variables
  mutate(timeframe=fullyear) %>%
  mutate(state='South Dakota') %>%
  mutate(dataformat='Percent') %>%
  mutate(locationtype='County') %>%
  mutate(varname='freereducedlunch') %>%
  
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(frl_data$locationid))>=1) {
  print(frl_data$location[is.na(frl_data$locationid)])
} else if (sum(is.na(frl_data$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- frl_data %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
View(frl_data)
```

## STEP 2: ADD TO DATABASE
```{r}
#CHECK DATASET NAMED frl_data TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,statefile,frl_data,append=TRUE,row.names=FALSE)
```

## STEP 3: OUTPUT FILE FOR DATA CENTER
```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", statefile," WHERE timeframe='",fullyear,"' AND varname='freereducedlunch' AND locationtype='County';")

upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/education/",statefile,"_",year,"_freereducedlunch.csv"),row.names=FALSE)
```



## ################################# ##
## PROCESS THE DATA FOR TRIBAL AREAS ##
## ################################# ##

## STEP 1: IMPORT AND CLEAN DATA
```{r}
#import the tribal schools crosswalk
tribalschools <- read.csv("./Documentation/Indicator Documentation/South Dakota Data Requests/southdakota_education_tribalschoolcrosswalk.csv")
tribalschools$school <- as.character(tribalschools$school)


frl_data_tribal <- read_excel(paste0("./Input/education/southdakota_",year,"_frl.xls"),sheet='Sort By School District') %>%
  rename(school='School/Agency Name',
         data='Free & Reduced-Price Eligibility Percentage',
         studentenrollment='Student Enrollment') %>%
  
  #add in additional information about tribal schools
  full_join(tribalschools,by=c('school'='school')) %>%
  subset(!is.na(location)) %>%
  
  #create a variable for >=90% for those with **
  mutate(studentenrollment=as.numeric(paste(studentenrollment))) %>%
  mutate(data=replace(data,data=='**' & studentenrollment>20,'>=90%')) %>%
  mutate(data=replace(data,data=='**' & studentenrollment<=20,NA)) %>%
  
  #add in Kids Count variables
  select(c(data,location,tribalarea)) %>%
  mutate(timeframe=fullyear) %>%
  mutate(state='South Dakota') %>%
  mutate(dataformat='Percent') %>%
  mutate(locationtype='Tribal Area') %>%
  mutate(varname='freereducedlunch') 


####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Output cases where percent data is greater than 1
temp_percheck <- frl_data_tribal %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 2. Visually inspect output data
View(frl_data_tribal)
```

## STEP 2: ADD TO DATABASE
```{r}
#CHECK DATASET NAMED frl_data_tribal TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,statefile,frl_data_tribal,append=TRUE,row.names=FALSE)
```

## STEP 3: OUTPUT IS DONE IN SEPARATE MARKDOWN ALONG WITH ALL TRIBAL DATA


