---
title: "South Dakota Vital Statistics"
output: html_document
---

## Indicator 1: Total births (1-year totals)
## Indicator 2: Low birthweight babies (5-year totals)
## Indicator 3: Births to single teens (5-year totals)
## Indicator 4: Births to mother receiving prenatal care during first trimester (5-year totals)
## Indicator 5: Infant mortality (5-year totals)
## Indicator 6: Child deaths (5-year totals)
## Indicator 7: Teen violent deaths (5-year totals)

###### Tribal Data Book Data
## Indicator 8: Resident  births by county to American Indian mothers
## Indicator 9: Low birthweight births to American Indian mothers
## Indicator 10: Resident births by county to single American Indian teens under age 20
## Indicator 11: Resident births by county to American Indian mothers that initiated prenatal care in the 1st trimester
## Indicator 12: Resident deaths by county for American Indian population
## Indicator 13: Resident infant deaths by county for American Indian population
## Indicator 14: Resident deaths by county for American Indian population ages 1-14
## Indicator 15: Resident violent deaths by county for American Indian population ages 15-19


**Created by:** Xanna Burg
**Date:** October 2020
**Updated by:**

**Data Source:** South Dakota Department of Health
**Purpose:** Clean and process the vital statistics data requested from DOH into correct format for Kids Count Data Center.

**Data format:** Final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Category (for Medicaid groups, and for recipient/household groups; character) Data (numeric: number, percentage), DataFormat (character: "number" or "percent" or "currency"), LocationId (numeric: assigned for KIDS COUNT system)


**To use this code for a new year:**
* Update the year in the second code chunk for variable 'year'
* Check each dataset visually and through the report logs prior to commiting to the database.



```{r,message=FALSE}
#load required packages
library(tidyverse)
library(RPostgreSQL)
library(readxl)
library(naniar)
library(stringr)
```


```{r}
####UPDATE to reflect the current year data working with
year <- '2020'
year_5 <- '2016-2020'
statename <- 'South Dakota'


#run this code to create an object needed for the database (DO NOT EDIT)
database_state <- 'southdakota'

#input location ID file for SD
locationids <- read.csv("./Input/SD KC Location IDs.csv")
locationids$Location <- as.character(locationids$Location)

#input list of all counties
countylist <- read.csv("./Input/SD County List.csv")

```



## ############################ ##
## TOTAL BIRTHS (1-YEAR TOTALS) ##
## ############################ ##

## Live births in the columns may be different depending on unknowns. Confirm with DOH to know which one doesn't have missing data ##
```{r}
#import and clean the data
births <- read_excel(path=paste0("./Input/health/southdakota_",year,"_vitalstatistics.xlsx"),sheet='Teen Births',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`Single Teen Births`,`% Single Teen Births`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(data=`Live Births`) %>%
  mutate(data=as.numeric(paste(data))) %>%
  #change number to <3
  mutate(data=as.character(paste(data))) %>%
  mutate(data=replace(data,data=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year,
         state=statename,
         varname='totalbirths_1year',
         dataformat='Number') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(births$locationid))>=1) {
  print(births$location[is.na(births$locationid)])
} else if (sum(is.na(births$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- births %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- births %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(births)

```

```{r}
#CHECK DATASET NAMED births TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,births,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year,"' AND varname='totalbirths_1year';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year,"_totalbirths_1year.csv"),row.names=FALSE)
```



## ###################################### ##
## LOW BIRTHWEIGHT BABIES (5-YEAR TOTALS) ##
## ###################################### ##

```{r}
#import and clean the data
lbw <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticswithconfidenceintervals.xlsx"),sheet='LBW',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`...9`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Low weight births`,
         Percent=`% Low birth weight`,
         moe=`95% CI ±`,
         lci=`Lower Limit`,
         uci=`Upper Limit`) %>%
  select(-c(`Live Births`)) %>%
  mutate(Number=as.numeric(paste(Number)),
         Percent=as.numeric(paste(Percent)),
         moe=as.numeric(paste(moe)),
         lci=as.numeric(paste(moe)),
         uci=as.numeric(paste(moe))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='lowbirthweight5yeartotals') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
lbw_number <- lbw %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
lbw_percent <- lbw %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Percent,moe,lci,uci)) %>%
  rename(data=Percent) %>%
  mutate(dataformat='Percent') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
lbw_all <- lbw_number %>%
  bind_rows(lbw_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(lbw_all$locationid))>=1) {
  print(lbw_all$location[is.na(lbw_all$locationid)])
} else if (sum(is.na(lbw_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- lbw_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- lbw_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(lbw_all)

```

```{r}
#CHECK DATASET NAMED lbw_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,lbw_all,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='lowbirthweight5yeartotals' AND dataformat='Percent';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year_5,"_lowbirthweight5yeartotals.csv"),row.names=FALSE)


#also write a query that exports the confidence intervals for copying into template
moe_sql <- paste0("SELECT locationtype, timeframe, location, data, moe FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='lowbirthweight5yeartotals' AND dataformat='Percent';")


upload_moe1 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='County') %>%
  arrange(location)

upload_moe2 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='State') %>%
  arrange(location)

upload_moe <- upload_moe1 %>%
  bind_rows(upload_moe2)



#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_moe,file=paste0("./Output/health/moe/",database_state,"_",year_5,"_lowbirthweight5yeartotals_moe.csv"),row.names=FALSE)

```


## ###################################### ##
## BIRTHS TO SINGLE TEENS (5-YEAR TOTALS) ##
## ###################################### ##

```{r}
#import and clean the data
teens <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticswithconfidenceintervals.xlsx"),sheet='Teen births',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`...9`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Single Teen Births`,
         Percent=`% Single Teen Births`,
         moe=`95% CI ±`,
         lci=`Lower Limit`,
         uci=`Upper Limit`) %>%
  select(-c(`Live Births`)) %>%
  mutate(Number=as.numeric(paste(Number)),
         Percent=as.numeric(paste(Percent)),
         moe=as.numeric(paste(moe)),
         lci=as.numeric(paste(moe)),
         uci=as.numeric(paste(moe))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='birthstosingleteens5yeartotals') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
teens_number <- teens %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
teens_percent <- teens %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Percent,moe,lci,uci)) %>%
  rename(data=Percent) %>%
  mutate(dataformat='Percent') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
teens_all <- teens_number %>%
  bind_rows(teens_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(teens_all$locationid))>=1) {
  print(teens_all$location[is.na(teens_all$locationid)])
} else if (sum(is.na(teens_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- teens_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- teens_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(teens_all)

```

```{r}
#CHECK DATASET NAMED teens_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,teens_all,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='birthstosingleteens5yeartotals' AND dataformat='Percent';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year_5,"_birthstosingleteens5yeartotals.csv"),row.names=FALSE)


#also write a query that exports the confidence intervals for copying into template
moe_sql <- paste0("SELECT locationtype, timeframe, location, data, moe FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='birthstosingleteens5yeartotals' AND dataformat='Percent';")


upload_moe1 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='County') %>%
  arrange(location)

upload_moe2 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='State') %>%
  arrange(location)

upload_moe <- upload_moe1 %>%
  bind_rows(upload_moe2)



#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_moe,file=paste0("./Output/health/moe/",database_state,"_",year_5,"_birthstosingleteens5yeartotals_moe.csv"),row.names=FALSE)

```



## ########################################## ##
## BIRTHS TO MOTHERS RECEIVING PRENATAL CARE  ##
## DURING THE FIRST TRIMETSER (5-YEAR TOTALS) ##
## ########################################## ##

```{r}
#import and clean the data
pnc <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticswithconfidenceintervals.xlsx"),sheet='Prenatal Care',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`...9`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`1st Trimester Care`,
         Percent=`% 1st Trimester Care`,
         moe=`95% CI ±`,
         lci=`Lower Limit`,
         uci=`Upper Limit`) %>%
  select(-c(`Live Births`)) %>%
  mutate(Number=as.numeric(paste(Number)),
         Percent=as.numeric(paste(Percent)),
         moe=as.numeric(paste(moe)),
         lci=as.numeric(paste(moe)),
         uci=as.numeric(paste(moe))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='firsttrimesterprenatal5yeartotals') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
pnc_number <- pnc %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
pnc_percent <- pnc %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Percent,moe,lci,uci)) %>%
  rename(data=Percent) %>%
  mutate(dataformat='Percent') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
pnc_all <- pnc_number %>%
  bind_rows(pnc_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(pnc_all$locationid))>=1) {
  print(pnc_all$location[is.na(pnc_all$locationid)])
} else if (sum(is.na(pnc_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- pnc_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- pnc_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(pnc_all)

```

```{r}
#CHECK DATASET NAMED pnc_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,pnc_all,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='firsttrimesterprenatal5yeartotals' AND dataformat='Percent';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year_5,"_firsttrimesterprenatal5yeartotals.csv"),row.names=FALSE)


#also write a query that exports the confidence intervals for copying into template
moe_sql <- paste0("SELECT locationtype, timeframe, location, data, moe FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='firsttrimesterprenatal5yeartotals' AND dataformat='Percent';")


upload_moe1 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='County') %>%
  arrange(location)

upload_moe2 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='State') %>%
  arrange(location)

upload_moe <- upload_moe1 %>%
  bind_rows(upload_moe2)



#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_moe,file=paste0("./Output/health/moe/",database_state,"_",year_5,"_firsttrimesterprenatal5yeartotals_moe.csv"),row.names=FALSE)

```




## ################################ ##
## INFANT MORTALITY (5-YEAR TOTALS) ##
## ################################ ##

```{r}
#import and clean the data
infantdeaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticswithconfidenceintervals.xlsx"),sheet='Infant Death',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`...9`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Infant Deaths`,
         Rate=`Infant Mortality Rate`,
         moe=`95% CI ±`,
         lci=`Lower Limit`,
         uci=`Upper Limit`) %>%
  select(-c(`Live Births`)) %>%
  mutate(Number=as.numeric(paste(Number)),
         Rate=as.numeric(paste(Rate)),
         moe=as.numeric(paste(moe)),
         lci=as.numeric(paste(moe)),
         uci=as.numeric(paste(moe))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='infantmortality5yeartotals') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
infantdeaths_number <- infantdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
infantdeaths_rate <- infantdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Rate,moe,lci,uci)) %>%
  rename(data=Rate) %>%
  mutate(dataformat='Rate') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
infantdeaths_all <- infantdeaths_number %>%
  bind_rows(infantdeaths_rate)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(infantdeaths_all$locationid))>=1) {
  print(infantdeaths_all$location[is.na(infantdeaths_all$locationid)])
} else if (sum(is.na(infantdeaths_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- infantdeaths_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- infantdeaths_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(infantdeaths_all)

```

```{r}
#CHECK DATASET NAMED infantdeaths_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,infantdeaths_all,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='infantmortality5yeartotals' AND dataformat='Rate';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year_5,"_infantmortality5yeartotals.csv"),row.names=FALSE)


#also write a query that exports the confidence intervals for copying into template
moe_sql <- paste0("SELECT locationtype, timeframe, location, data, moe FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='infantmortality5yeartotals' AND dataformat='Rate';")


upload_moe1 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='County') %>%
  arrange(location)

upload_moe2 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='State') %>%
  arrange(location)

upload_moe <- upload_moe1 %>%
  bind_rows(upload_moe2)



#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_moe,file=paste0("./Output/health/moe/",database_state,"_",year_5,"_infantmortality5yeartotals_moe.csv"),row.names=FALSE)

```



## ############################ ##
## CHILD DEATHS (5-YEAR TOTALS) ##
## ############################ ##

```{r}
#import and clean the data
childdeaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticswithconfidenceintervals.xlsx"),sheet='Child Death',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`...9`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`POP 1-14`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Deaths`,
         Rate=`Rate`,
         moe=`95% CI ±`,
         lci=`Lower Limit`,
         uci=`Upper Limit`) %>%
  select(-c(`POP 1-14`)) %>%
  mutate(Number=as.numeric(paste(Number)),
         Rate=as.numeric(paste(Rate)),
         moe=as.numeric(paste(moe)),
         lci=as.numeric(paste(moe)),
         uci=as.numeric(paste(moe))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='childdeaths5yeartotals') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
childdeaths_number <- childdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
childdeaths_rate <- childdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Rate,moe,lci,uci)) %>%
  rename(data=Rate) %>%
  mutate(dataformat='Rate') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
childdeaths_all <- childdeaths_number %>%
  bind_rows(childdeaths_rate)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childdeaths_all$locationid))>=1) {
  print(childdeaths_all$location[is.na(childdeaths_all$locationid)])
} else if (sum(is.na(childdeaths_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- childdeaths_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- childdeaths_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(childdeaths_all)

```

```{r}
#CHECK DATASET NAMED childdeaths_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,childdeaths_all,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='childdeaths5yeartotals' AND dataformat='Rate';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year_5,"_childdeaths5yeartotals.csv"),row.names=FALSE)


#also write a query that exports the confidence intervals for copying into template
moe_sql <- paste0("SELECT locationtype, timeframe, location, data, moe FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='childdeaths5yeartotals' AND dataformat='Rate';")


upload_moe1 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='County') %>%
  arrange(location)

upload_moe2 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='State') %>%
  arrange(location)

upload_moe <- upload_moe1 %>%
  bind_rows(upload_moe2)



#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_moe,file=paste0("./Output/health/moe/",database_state,"_",year_5,"_childdeaths5yeartotals_moe.csv"),row.names=FALSE)

```



## ################################### ##
## TEEN VIOLENT DEATHS (5-YEAR TOTALS) ##
## ################################### ##

```{r}
#import and clean the data
teendeaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticswithconfidenceintervals.xlsx"),sheet='Teen violent death',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`,`...9`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Pop 15-19`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Deaths`,
         Rate=`Rate`,
         moe=`95% CI ±`,
         lci=`Lower Limit`,
         uci=`Upper Limit`) %>%
  select(-c(`Pop 15-19`)) %>%
  mutate(Number=as.numeric(paste(Number)),
         Rate=as.numeric(paste(Rate)),
         moe=as.numeric(paste(moe)),
         lci=as.numeric(paste(moe)),
         uci=as.numeric(paste(moe))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='teenviolentdeaths5yeartotals') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
teendeaths_number <- teendeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
teendeaths_rate <- teendeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Rate,moe,lci,uci)) %>%
  rename(data=Rate) %>%
  mutate(dataformat='Rate') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
teendeaths_all <- teendeaths_number %>%
  bind_rows(teendeaths_rate)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(teendeaths_all$locationid))>=1) {
  print(teendeaths_all$location[is.na(teendeaths_all$locationid)])
} else if (sum(is.na(teendeaths_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- teendeaths_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- teendeaths_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(teendeaths_all)

```

```{r}
#CHECK DATASET NAMED teendeaths_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,teendeaths_all,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

upload_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='teenviolentdeaths5yeartotals' AND dataformat='Rate';")


upload_datacenter <- dbGetQuery(con,upload_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_datacenter,file=paste0("./Output/health/",database_state,"_",year_5,"_teenviolentdeaths5yeartotals.csv"),row.names=FALSE)


#also write a query that exports the confidence intervals for copying into template
moe_sql <- paste0("SELECT locationtype, timeframe, location, data, moe FROM ", database_state," WHERE timeframe='",year_5,"' AND varname='teenviolentdeaths5yeartotals' AND dataformat='Rate';")


upload_moe1 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='County') %>%
  arrange(location)

upload_moe2 <- dbGetQuery(con,moe_sql) %>%
  subset(locationtype=='State') %>%
  arrange(location)

upload_moe <- upload_moe1 %>%
  bind_rows(upload_moe2)



#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_moe,file=paste0("./Output/health/moe/",database_state,"_",year_5,"_teenviolentdeaths5yeartotals_moe.csv"),row.names=FALSE)

```






## DATA FOR SOUTH DAKOTA TRIBAL DATA BOOKS



## #################################################### ##
## RESIDENT BIRTHS BY COUNTY TO AMERICAN INDIAN MOTHERS ##
## #################################################### ##

```{r}
#---DO NOT EDIT DATES---#
year_minus1 <- as.character(as.numeric(year)-1)
year_minus2 <- as.character(as.numeric(year)-2)
year_minus3 <- as.character(as.numeric(year)-3)
year_minus4 <- as.character(as.numeric(year)-4)

#first import total births across all races to use for calculating percentages
totalbirths_sql <- paste0("SELECT location, timeframe, data FROM ", database_state," WHERE varname='totalbirths_1year'")

totalbirths <- dbGetQuery(con,totalbirths_sql) %>%
  mutate(data=as.numeric(paste(data))) %>%
  #select the most recent five years
  subset(timeframe==year | timeframe==year_minus1 | timeframe==year_minus2 | 
           timeframe==year_minus3 | timeframe==year_minus4) %>%
  group_by(location) %>%
  summarise(totalbirths=sum(data),.groups='keep') %>%
  ungroup

#import and clean the data
na_births <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI births',skip=1) %>%
  rename(location=`Resident County`,
         Number=Count) %>%
  #remove extra text and Unknown county
  subset(!is.na(Number)) %>%
  subset(location!='Not Stated') %>%
  
  #if there are zero in a county, assign in NA
  mutate(Number=replace(Number,Number=='0','NA')) %>%
  
  #merge in total births to calculate percentage
  mutate(Number=as.numeric(paste(Number))) %>%
  left_join(totalbirths, by=c('location'='location')) %>%
  mutate(Percent=Number/totalbirths) %>%
  select(-c(totalbirths)) %>%

  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='totalbirths_5year_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  
  #pivot longer
  mutate(Percent=as.character(paste(Percent))) %>%
  pivot_longer(cols=c(Number,Percent),names_to='dataformat',values_to='data')




####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(na_births$locationid))>=1) {
  print(na_births$location[is.na(na_births$locationid)])
} else if (sum(is.na(na_births$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- na_births %>%  mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- na_births %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(na_births)

```

```{r}
#CHECK DATASET NAMED na_births TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,na_births,append=TRUE,row.names=FALSE)
```


## ################################################# ##
## LOW BIRTHWEIGHT BIRTHS TO AMERICAN INDIAN MOTHERS ##
## ################################################# ##

```{r}
#import and clean the data
lbw <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI LBW',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Low weight births`,
         Percent=`% Low birth weight`) %>%
  select(-c(`Live Births`)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign NA to missing  counties
  mutate(Number=replace(Number,is.na(Number),NA)) %>%
  mutate(Percent=replace(Percent,is.na(Percent),NA)) %>%
  
  mutate(Number=as.numeric(paste(Number)),
         Percent=as.numeric(paste(Percent))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='lowbirthweight5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
lbw_number <- lbw %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
lbw_percent <- lbw %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Percent)) %>%
  rename(data=Percent) %>%
  mutate(dataformat='Percent') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
lbw_all <- lbw_number %>%
  bind_rows(lbw_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(lbw_all$locationid))>=1) {
  print(lbw_all$location[is.na(lbw_all$locationid)])
} else if (sum(is.na(lbw_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- lbw_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- lbw_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(lbw_all)

```

```{r}
#CHECK DATASET NAMED lbw_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,lbw_all,append=TRUE,row.names=FALSE)
```



## ################################### ##
## RESIDENT BIRTHS BY COUNTY TO SINGLE ## 
## AMERICAN INDIAN TEENS UNDER AGE 20  ##
## ################################### ##

```{r}
#import and clean the data
teens <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI Teen Births',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Single Teen Births`,
         Percent=`% Single Teen Births`) %>%
  select(-c(`Live Births`)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign zero to missing  counties
  mutate(Number=replace(Number,is.na(Number),NA)) %>%
  mutate(Percent=replace(Percent,is.na(Percent),NA)) %>%
  
  mutate(Number=as.numeric(paste(Number)),
         Percent=as.numeric(paste(Percent))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='birthstosingleteens5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
teens_number <- teens %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
teens_percent <- teens %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Percent)) %>%
  rename(data=Percent) %>%
  mutate(dataformat='Percent') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
teens_all <- teens_number %>%
  bind_rows(teens_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(teens_all$locationid))>=1) {
  print(teens_all$location[is.na(teens_all$locationid)])
} else if (sum(is.na(teens_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- teens_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- teens_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(teens_all)

```

```{r}
#CHECK DATASET NAMED teens_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,teens_all,append=TRUE,row.names=FALSE)
```


## ##################################### ##
## RESIDENT BIRTHS BY COUNTY TO AMERICAN ## 
## INDIAN MOTHERS THAT INITIATED         ##
## PRENATAL CARE IN THE FIRST TRIMESTER  ##
## ##################################### ##

```{r}
#import and clean the data
pnc <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI Prenatal Care',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`1st Trimester Care`,
         Percent=`% 1st Trimester Care`) %>%
  select(-c(`Live Births`)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign zero to missing  counties
  mutate(Number=replace(Number,is.na(Number),NA)) %>%
  mutate(Percent=replace(Percent,is.na(Percent),NA)) %>%
  
  mutate(Number=as.numeric(paste(Number)),
         Percent=as.numeric(paste(Percent))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='firsttrimesterprenatal5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
pnc_number <- pnc %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
pnc_percent <- pnc %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Percent)) %>%
  rename(data=Percent) %>%
  mutate(dataformat='Percent') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
pnc_all <- pnc_number %>%
  bind_rows(pnc_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(pnc_all$locationid))>=1) {
  print(pnc_all$location[is.na(pnc_all$locationid)])
} else if (sum(is.na(pnc_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- pnc_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- pnc_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(pnc_all)

```

```{r}
#CHECK DATASET NAMED pnc_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,pnc_all,append=TRUE,row.names=FALSE)
```



## ####################################################### ##
## RESIDENT DEATH BY COUNTY FOR AMERICAN INDIAN POPULATION ##
## ####################################################### ##

```{r}
#import and clean the data
deaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI deaths',skip=1) %>%
  rename(location=`Resident County`,
         data=Count) %>%
  #remove extra text and Unknown county
  subset(!is.na(data)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign zero to missing  counties
  mutate(data=replace(data,is.na(data),NA)) %>%
  
  mutate(data=as.numeric(paste(data))) %>%
  mutate(data=replace(data,data==0,NA)) %>%
  #change number to <3
  mutate(data=as.character(paste(data))) %>%
  mutate(data=replace(data,data=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='residentdeaths5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) %>%
  mutate(dataformat='Number')




####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(deaths$locationid))>=1) {
  print(deaths$location[is.na(deaths$locationid)])
} else if (sum(is.na(deaths$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- deaths %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- deaths %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(deaths)

```

```{r}
#CHECK DATASET NAMED deaths TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,deaths,append=TRUE,row.names=FALSE)
```



## ################################################################ ##
## RESIDENT INFANT DEATHS BY COUNTY FOR AMERICAN INDIAN POPULATION  ##
## ################################################################ ##

```{r}
#import and clean the data
infantdeaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI Infant Deaths',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Live Births`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Infant Deaths`,
         Rate=`Infant Mortality Rate`) %>%
  select(-c(`Live Births`)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign zero to missing  counties
  mutate(Number=replace(Number,is.na(Number),NA)) %>%
  mutate(Rate=replace(Rate,is.na(Rate),NA)) %>%
  
  mutate(Number=as.numeric(paste(Number)),
         Rate=as.numeric(paste(Rate))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='infantmortality5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
infantdeaths_number <- infantdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
infantdeaths_percent <- infantdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Rate)) %>%
  rename(data=Rate) %>%
  mutate(dataformat='Rate') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
infantdeaths_all <- infantdeaths_number %>%
  bind_rows(infantdeaths_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(infantdeaths_all$locationid))>=1) {
  print(infantdeaths_all$location[is.na(infantdeaths_all$locationid)])
} else if (sum(is.na(infantdeaths_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- infantdeaths_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- infantdeaths_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(infantdeaths_all)

```

```{r}
#CHECK DATASET NAMED infantdeaths_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,infantdeaths_all,append=TRUE,row.names=FALSE)
```



## ################################################################### ##
## RESIDENT DEATHS BY COUNTY FOR AMERICAN INDIAN POPULATION AGES 1-14  ##
## ################################################################### ##

```{r}
#import and clean the data
childdeaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI Child Deaths',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Pop 1-14 AI`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Deaths`,
         Rate=`Rate`) %>%
  select(-c(`Pop 1-14 AI`)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign zero to missing  counties
  mutate(Number=replace(Number,is.na(Number),NA)) %>%
  mutate(Rate=replace(Rate,is.na(Rate),NA)) %>%
  
  mutate(Number=as.numeric(paste(Number)),
         Rate=as.numeric(paste(Rate))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='childdeaths5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
childdeaths_number <- childdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
childdeaths_percent <- childdeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Rate)) %>%
  rename(data=Rate) %>%
  mutate(dataformat='Rate') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
childdeaths_all <- childdeaths_number %>%
  bind_rows(childdeaths_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(childdeaths_all$locationid))>=1) {
  print(childdeaths_all$location[is.na(childdeaths_all$locationid)])
} else if (sum(is.na(childdeaths_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- childdeaths_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- childdeaths_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(childdeaths_all)

```

```{r}
#CHECK DATASET NAMED childdeaths_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,childdeaths_all,append=TRUE,row.names=FALSE)
```



## ##################################### ##
## RESIDENT VIOLENT DEATHS BY COUNTY FOR ##
## AMERICAN INDIAN POPULATION AGES 15-19 ##
## ##################################### ##

```{r}
#import and clean the data
teendeaths <- read_excel(path=paste0("./Input/health/southdakota_",year_5,"_vitalstatisticsamericanindian.xlsx"),sheet='AI Teen Deaths',skip=1) %>%
  rename(location=`...2`) %>%
  select(-c(`...1`)) %>%
  #remove extra text and Unknown county
  subset(!is.na(`Pop 15-19 AI`)) %>%
  mutate(location=replace(location,is.na(location),'South Dakota')) %>%
  subset(location!='Unknown') %>%
  
  rename(Number=`Deaths`,
         Rate=`Rate`) %>%
  select(-c(`Pop 15-19 AI`)) %>%
  
  #merge in county list to include missing counties
  full_join(countylist,by=c('location'='county')) %>%
  #assign zero to missing  counties
  mutate(Number=replace(Number,is.na(Number),NA)) %>%
  mutate(Rate=replace(Rate,is.na(Rate),NA)) %>%
  
  mutate(Number=as.numeric(paste(Number)),
         Rate=as.numeric(paste(Rate))) %>%
  #change number to <3
  mutate(Number=as.character(paste(Number))) %>%
  mutate(Number=replace(Number,Number=='NA','<3')) %>%
  
  #add in KC variables
  mutate(locationtype=ifelse(location=='South Dakota','State','County')) %>%
  mutate(timeframe=year_5,
         state=statename,
         varname='teenviolentdeaths5yeartotals_na') %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)

#create dataset that only includes number
teendeaths_number <- teendeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Number)) %>%
  rename(data=Number) %>%
  mutate(dataformat='Number')

#create dataset that only includes percent
teendeaths_percent <- teendeaths %>%
  select(c(location,locationtype,locationid,timeframe,state,varname,Rate)) %>%
  rename(data=Rate) %>%
  mutate(dataformat='Rate') %>%
  
  mutate(data=as.character(paste(data)))
  

#create dataset that includes all data
teendeaths_all <- teendeaths_number %>%
  bind_rows(teendeaths_percent)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(teendeaths_all$locationid))>=1) {
  print(teendeaths_all$location[is.na(teendeaths_all$locationid)])
} else if (sum(is.na(teendeaths_all$locationid))==0) {
  'all locations match'
}

#2. Check that no values are less than 3
temp_suppresscheck <- teendeaths_all %>% mutate(data=as.numeric(paste(data))) %>% subset(dataformat=='Number') %>% subset(data<3)
temp_rows <- as.numeric(nrow(temp_suppresscheck))
if (temp_rows==0) {
  'data suppression followed at individual location level'
} else if (temp_rows>=1) {
  View(temp_suppresscheck)
}

#3. Check that there is no suppression or >1 suppression
########## MANUAL STEP #############
#need to check that county data cannot be identified using totals
temp_testsuppression <- teendeaths_all %>%
  subset(locationtype=='County' & dataformat=='Number') %>%
  group_by(location) %>%
  summarise_all(~sum(is.na(.))) %>%
  transmute(location,sumNA=rowSums(.[-1])) %>%
  subset(sumNA==1)

temp_rows <- as.numeric(nrow(temp_testsuppression))
if (temp_rows==0 | temp_rows>1) {
  'no additional data suppression needed'
}


# 4. Visually inspect output data
View(teendeaths_all)

```

```{r}
#CHECK DATASET NAMED teendeaths_all TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,teendeaths_all,append=TRUE,row.names=FALSE)
```


