---
title: "Child population by ethnicity"
author: "Xanna Burg"
date: "4/20/2020"
output: html_document
---


## Indicator: Child population by ethnicity

**This code populates historic estimates of population by ethnicity for age groups. The API call does not allow for multiple predicates to filter by both age and ethnicity The dataset for 2000-2009 can by downloaded from CDC Wonder**

https://wonder.cdc.gov/controller/datarequest/
Group results by: County, Yearly July 1st Estimates, Ethnicity
Select State
Ages 0 to 19
All races, Pick both Ethnicities, then years 2000-2009
Export results to .txt then save as CSV


**Created by:** Xanna Burg
**Date:** April 2020
**Updated by:**

**Data Source:** U.S. Census Bureau, Annual Population Estimates. Specific table: pep/charagegroups 
**Purpose:** Connect to census data table, clean data, add to master database, output csv to upload to Data Center in correct format. Update Montana, North Dakota, and South Dakota KIDS COUNT Data Center indicator with most recent year of vintage data.
**Description:** this program calculates **two indicators: Child population by race and Child population by age group and race**
* Geographies: state total, county, state planning regions (where applicable, which are groups of counties)
* Number - count of population
* Note - this data uses un-bridged race categories


**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system), Race (character), Age group (character)



```{r,message=FALSE}
#load required packages
library(tidyverse)
library(tidycensus)
library(censusapi)
library(stringr)
```

```{r}
#county and state names that match fips codes (from tidycensus package)
fips <- fips_codes
fips$county_code = str_remove(fips$county_code, "^0+")

#import tribal fips code (from file on local drive)
fips_tribal <- read.csv("../Input/AIANNHCE Codes.csv")

```


```{r}
#CHANGE THE OBJECTS 'statename' and 'year' TO MATCH THE STATE AND YEAR
#state should match exactly as is in FIPS code in order for code to correctly run
statename <- "Montana"
year <- "2000"


#run this code to create an object needed for the database (DO NOT EDIT)
if (statename=='Montana') {
  database_state <- 'montana'
} else if (statename=='North Dakota') {
  database_state <- 'northdakota'
} else if (statename=='South Dakota') {
  database_state <- 'southdakota'
}

#import the location ids matching the correct state (for MT, ND, and SD; DO NOT EDIT)
locationids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Location IDs.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Location IDs.csv")
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Location IDs.csv")
}
locationids$Location <- as.character(locationids$Location) #assign as character instead of factor for merging

#import the region match file matching the correct state (for MT, ND, and SD; DO NOT EDIT)
regionids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Region List.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Region List.csv") 
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Region List.csv")
}
regionids$county <- as.character(regionids$county)
regionids$region <- as.character(regionids$region)

#RUN THIS CODE, BUT NOT REQUIRED TO CHANGE ANYTHING
#the api will subset data to each state based on the state FIPS code: MT=30, North Dakota=38, South Dakota=46
statecode <- as.numeric(unique(fips$state_code[fips$state_name==statename]))


pop_data <- read.csv(paste0('../Input/demographics/',database_state,'_',year,'_childpopulationbyethnicity.csv'))
```

```{r}
#COUNTY DATA
pop_county <- pop_data %>%
  
  #subset for total rows
  subset(Ethnicity != '') %>%
  
  #create the timeframe using the Date codes
  rename(timeframe='Yearly.July.1st.Estimates',
         data=Population,
         race=Ethnicity) %>%
  mutate(location=gsub(" County.*","",County)) %>%
  
  select(c(location,timeframe,data,race)) %>%
  
  mutate(race=case_when(
    race=='Hispanic or Latino' ~ 'Hispanic',
    race=='Not Hispanic or Latino' ~ 'Non-Hispanic')) %>%
  
  #create a wide dataset in order to calculate totals and percent
  pivot_wider(values_from=data,names_from=race) %>%
  mutate(totalpop=Hispanic+`Non-Hispanic`) %>%
  mutate(percent_hispanic=Hispanic/totalpop) %>%
  mutate(percent_nonhispanic=`Non-Hispanic`/totalpop) %>%
  
  #convert back to long format
  pivot_longer(cols=c(Hispanic,`Non-Hispanic`),names_to='race',values_to='Number') %>%
  pivot_longer(cols=c(percent_hispanic,percent_nonhispanic),names_to='percent_race',values_to='Percent') %>%
  mutate(percent_race=case_when(
    percent_race=='percent_hispanic' ~ 'Hispanic',
    percent_race=='percent_nonhispanic' ~ 'Non-Hispanic')) %>%
  subset(race==percent_race) %>%
  select(-c(percent_race,totalpop)) %>%
  
  pivot_longer(cols=c(Number,Percent),names_to='dataformat',values_to='data') %>%
  
  mutate(locationtype='County')
  
  

###########
#STATE DATA
pop_state <- pop_data %>%
  
  #subset for total rows
  subset(Ethnicity != '') %>%
  
  #create the timeframe using the Date codes
  rename(timeframe='Yearly.July.1st.Estimates',
         data=Population,
         race=Ethnicity) %>%
  mutate(location=gsub(" County.*","",County)) %>%
  
  select(c(location,timeframe,data,race)) %>%
  
  mutate(race=case_when(
    race=='Hispanic or Latino' ~ 'Hispanic',
    race=='Not Hispanic or Latino' ~ 'Non-Hispanic')) %>%
  
  #create a wide dataset in order to calculate totals and percent
  pivot_wider(values_from=data,names_from=race) %>%
  mutate(totalpop=Hispanic+`Non-Hispanic`) %>%
  
  #calculate state sums
  mutate(location=statename) %>%
  group_by(location,timeframe) %>%
  summarise(Hispanic=sum(Hispanic),
            `Non-Hispanic`=sum(`Non-Hispanic`),
            totalpop=sum(totalpop)) %>%
  ungroup %>%
  
  mutate(percent_hispanic=Hispanic/totalpop) %>%
  mutate(percent_nonhispanic=`Non-Hispanic`/totalpop) %>%
  
  #convert back to long format
  pivot_longer(cols=c(Hispanic,`Non-Hispanic`),names_to='race',values_to='Number') %>%
  pivot_longer(cols=c(percent_hispanic,percent_nonhispanic),names_to='percent_race',values_to='Percent') %>%
  mutate(percent_race=case_when(
    percent_race=='percent_hispanic' ~ 'Hispanic',
    percent_race=='percent_nonhispanic' ~ 'Non-Hispanic')) %>%
  subset(race==percent_race) %>%
  select(-c(percent_race,totalpop)) %>%
  
  pivot_longer(cols=c(Number,Percent),names_to='dataformat',values_to='data') %>%
  
  mutate(locationtype='State')


############
#REGION DATA
pop_region <- pop_data %>%
  
  #subset for total rows
  subset(Ethnicity != '') %>%
  
  #create the timeframe using the Date codes
  rename(timeframe='Yearly.July.1st.Estimates',
         data=Population,
         race=Ethnicity) %>%
  mutate(location=gsub(" County.*","",County)) %>%
  
  select(c(location,timeframe,data,race)) %>%
  
  mutate(race=case_when(
    race=='Hispanic or Latino' ~ 'Hispanic',
    race=='Not Hispanic or Latino' ~ 'Non-Hispanic')) %>%
  
  #create a wide dataset in order to calculate totals and percent
  pivot_wider(values_from=data,names_from=race) %>%
  mutate(totalpop=Hispanic+`Non-Hispanic`) %>%
  
  #calculate state sums
  left_join(regionids,(by=c('location'='county'))) %>%
  subset(region!='') %>%
  select(-c(location)) %>%
  rename(location=region) %>%
  group_by(location,timeframe) %>%
  summarise(Hispanic=sum(Hispanic),
            `Non-Hispanic`=sum(`Non-Hispanic`),
            totalpop=sum(totalpop)) %>%
  ungroup %>%
  
  mutate(percent_hispanic=Hispanic/totalpop) %>%
  mutate(percent_nonhispanic=`Non-Hispanic`/totalpop) %>%
  
  #convert back to long format
  pivot_longer(cols=c(Hispanic,`Non-Hispanic`),names_to='race',values_to='Number') %>%
  pivot_longer(cols=c(percent_hispanic,percent_nonhispanic),names_to='percent_race',values_to='Percent') %>%
  mutate(percent_race=case_when(
    percent_race=='percent_hispanic' ~ 'Hispanic',
    percent_race=='percent_nonhispanic' ~ 'Non-Hispanic')) %>%
  subset(race==percent_race) %>%
  select(-c(percent_race,totalpop)) %>%
  
  pivot_longer(cols=c(Number,Percent),names_to='dataformat',values_to='data') %>%
  
  mutate(locationtype='Planning Region')

```

```{r}
#UNION COUNTY, STATE, STATE PLANNING REGION
pop <- pop_county %>%
  bind_rows(pop_state) %>%
  bind_rows(pop_region) %>%
  
  mutate(state=statename) %>%
  mutate(vintageyear=year) %>%
  mutate(varname='childpopulationbyethnicity') %>%
  
  #merge in location ids
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(pop$locationid))>=1) {
  print(pop$location[is.na(pop$locationid)])
} else if (sum(is.na(pop$locationid))==0) {
  'all locations match'
}


# 2. Output cases where percent data is greater than 1
temp_percheck <- pop %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
view(pop)

```

```{r}
#CHECK DATASET NAMED pop TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,pop,append=TRUE,row.names=FALSE)
```
