---
title: "Child population by ethnicity"
author: "Xanna Burg"
output: html_document
---


## Indicator: Child population by ethnicity


**Created by:** Xanna Burg
**Date:** April 2020
**Updated by:**

**Data Source:** U.S. Census Bureau, Annual Population Estimates. In 2020, the data was not available by API, this code was updated to pull data from a file downloaded from data.census.gov.
**Purpose:** Connect to census data table, clean data, add to master database, output csv to upload to Data Center in correct format. Update Montana, North Dakota, and South Dakota KIDS COUNT Data Center indicator with most recent year of vintage data.
**Description:** this program calculates **two indicators: Child population by race and Child population by age group and race**
* Geographies: state total, county, state planning regions (where applicable, which are groups of counties)
* Number - count of population
* Note - this data uses un-bridged race categories


**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system), Race (character), Age group (character)


**Data file downloaded from:** https://www.census.gov/data/datasets/time-series/demo/popest/2020s-counties-detail.html 


```{r,message=FALSE}
#load required packages
library(tidyverse)
library(tidycensus)
library(censusapi)
library(stringr)
```

```{r}
#metadata for available variables
#pop_vars <- listCensusMetadata(name="pep/charagegroups",
                              #vintage=year,
                               #type="variables")
#view(pop_vars)

#metadata for available geographies
#pop_geos <- listCensusMetadata(name="pep/charagegroups",
                                #vintage=year,
                               #type="geography")
#view(pop_geos)

#county and state names that match fips codes (from tidycensus package)
fips <- fips_codes %>%
  mutate(state_code=as.numeric(paste(state_code))) %>%
  mutate(county_code=as.numeric(paste(county_code)))
#fips$county_code = str_remove(fips$county_code, "^0+")

#import tribal fips code (from file on local drive)
fips_tribal <- read.csv("../Input/AIANNHCE Codes.csv")

```

## UPDATE THIS CODE CHUNK
```{r}
#CHANGE THE OBJECTS 'statename' and 'year' TO MATCH THE STATE AND YEAR
#state should match exactly as is in FIPS code in order for code to correctly run
statename <- "Montana"
year <- "2020"



#run this code to create an object needed for the database (DO NOT EDIT)
if (statename=='Montana') {
  database_state <- 'montana'
} else if (statename=='North Dakota') {
  database_state <- 'northdakota'
} else if (statename=='South Dakota') {
  database_state <- 'southdakota'
}

#import the location ids matching the correct state (for MT, ND, and SD; DO NOT EDIT)
locationids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Location IDs.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Location IDs.csv")
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Location IDs.csv")
}
locationids$Location <- as.character(locationids$Location) #assign as character instead of factor for merging

#import the region match file matching the correct state (for MT, ND, and SD; DO NOT EDIT)
regionids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Region List.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Region List.csv") 
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Region List.csv")
}
regionids$county <- as.character(regionids$county)
regionids$region <- as.character(regionids$region)

#RUN THIS CODE, BUT NOT REQUIRED TO CHANGE ANYTHING
#the api will subset data to each state based on the state FIPS code: MT=30, North Dakota=38, South Dakota=46
statecode <- as.numeric(unique(fips$state_code[fips$state_name==statename]))
```

## IMPORT THE CENSUS DATA
```{r}
#pep/charagegroups is: Population Estimates Program
#see lookup table for categorical vars: 
#https://www.census.gov/data/developers/data-sets/popest-popproj/popest/popest-vars.Vintage_2018.html

#######################
#create datasets that can be used for both total under age 20 and by age group

pop_county <- read.csv('../Input/demographics/CC-EST2020-ALLDATA.csv') %>%
  
  subset(STATE==statecode) %>%
  
  #subset for correct age groups
  subset(AGEGRP=='1' | AGEGRP=='2' | AGEGRP=='3' | AGEGRP=='4') %>%
  
  #create the timeframe using the Date codes
  mutate(timeframe=case_when(
    YEAR=='3' ~ '2010',
    YEAR=='4' ~ '2011',
    YEAR=='5' ~ '2012',
    YEAR=='6' ~ '2013',
    YEAR=='7' ~ '2014',
    YEAR=='8' ~ '2015',
    YEAR=='9' ~ '2016',
    YEAR=='10' ~ '2017',
    YEAR=='11' ~ '2018',
    YEAR=='12' ~ '2019',
    YEAR=='13' ~ '2020')) %>%
  subset(!is.na(timeframe)) %>%
  
  #assign population counts to numeric data
  mutate(NH_MALE=as.numeric(NH_MALE),
         NH_FEMALE=as.numeric(NH_FEMALE),
         H_MALE=as.numeric(H_MALE),
         H_FEMALE=as.numeric(H_FEMALE)) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('COUNTY'='county_code','STATE'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county)) %>%
    
    mutate(`Non-Hispanic`=NH_MALE+NH_FEMALE,
           `Hispanic`=H_MALE+H_FEMALE) %>%
  
  select(c(timeframe,county,AGEGRP,`Non-Hispanic`,`Hispanic`)) %>%
  
  group_by(timeframe,county) %>%
  summarise(`Non-Hispanic`=sum(`Non-Hispanic`),
            `Hispanic`=sum(`Hispanic`)) %>%
  ungroup %>%
  pivot_longer(cols=c(`Non-Hispanic`,`Hispanic`),names_to='race',values_to='data') %>%
  
  rename(location=county) %>%
  mutate(locationtype='County',
         state=statename,
         dataformat='Number',
         vintageyear=year,
         varname='childpopulationbyethnicity') %>%
  
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) 
    
    
  
  #rename race categories
  #mutate(race=case_when(
    #RACE=='0' ~ 'All races',
    #RACE=='1' ~ 'White',
    #RACE=='2' ~ 'Black or African American',
    #RACE=='3' ~ 'American Indian and Alaska Native',
    #RACE=='4' ~ 'Asian',
    #RACE=='5' ~ 'Native Hawaiian and Other Pacific Islander',
    #RACE=='6' ~ 'Two or more races')) %>%
  
  
  #select(c(county,POP,AGEGROUP,timeframe,state_name,race)) 




#STATE DATA
pop_state <- pop_county %>%
  mutate(location=statename) %>%
  group_by(timeframe,location,race,state,dataformat,vintageyear,varname) %>%
  summarise(data=sum(data)) %>%
  ungroup %>%
  mutate(locationtype='State')

#pop_state <- getCensus(name="pep/charagegroups",
 #         vintage=year,
  #        key=Sys.getenv("CENSUS_API_KEY"),
   #       vars=c("POP","DATE_CODE","DATE_DESC","AGEGROUP"),
    #      region=paste0("state:",statecode),
     #     RACE=0,
      #    RACE=1,
       #   RACE=2,
        #  RACE=3,
         # RACE=4,
          #RACE=5,
          #RACE=6,
          #HISP=0) %>%
  
  #subset(AGEGROUP=='1' | AGEGROUP=='2' | AGEGROUP=='3' | AGEGROUP=='4') %>%
  
  #create the timeframe using the Date codes
  #mutate(timeframe=case_when(
    #DATE_CODE=='3' ~ '2010',
    #DATE_CODE=='4' ~ '2011',
    #DATE_CODE=='5' ~ '2012',
    #DATE_CODE=='6' ~ '2013',
    #DATE_CODE=='7' ~ '2014',
    #DATE_CODE=='8' ~ '2015',
    #DATE_CODE=='9' ~ '2016',
    #DATE_CODE=='10' ~ '2017',
    #DATE_CODE=='11' ~ '2018',
    #DATE_CODE=='12' ~ '2019')) %>%
  #subset(!is.na(timeframe)) %>%
  
  #assign population counts to numeric data
  #mutate(POP=as.numeric(POP)) %>%
  
  
  #rename race categories
  #mutate(race=case_when(
    #RACE=='0' ~ 'All races',
    #RACE=='1' ~ 'White',
    #RACE=='2' ~ 'Black or African American',
    #RACE=='3' ~ 'American Indian and Alaska Native',
    #RACE=='4' ~ 'Asian',
    #RACE=='5' ~ 'Native Hawaiian and Other Pacific Islander',
    #RACE=='6' ~ 'Two or more races')) %>%
  
  #mutate(location=statename) %>%
  
  #select(c(location,POP,AGEGROUP,timeframe,race))
  
  
  
  
#REGION DATA
pop_region <- pop_county %>%
  left_join(regionids, by=c('location'='county')) %>%
  group_by(timeframe,region,race,state,dataformat,vintageyear,varname) %>%
  summarise(data=sum(data)) %>%
  ungroup %>%
  mutate(locationtype='Planning Region') %>%
  rename(location=region) %>%
  subset(location != '')
  

#pop_region <- pop_county %>%
 # mutate(county=replace(county, 
   #                       statename=='Montana' & county=='Lewis and Clark', 
    #                      'Lewis & Clark')) %>%
  #mutate(county=replace(county, 
     #                     statename=='South Dakota' & county=='Shannon', 
      #                    'Oglala Lakota')) %>%
  #left_join(regionids, by=c('county'='county')) %>%
  #group_by(region,AGEGROUP,race,timeframe) %>%
  #summarise(POP=sum(POP)) %>%
  #ungroup %>%
  #subset(region != '')


#combine all
pop <- pop_county %>%
  bind_rows(pop_state) %>%
  bind_rows(pop_region)
```

**Percentages **
```{r}
totalpop_county <- read.csv('../Input/demographics/CC-EST2020-ALLDATA.csv') %>%
  
  subset(STATE==statecode) %>%
  
  #subset for correct age groups
  subset(AGEGRP=='1' | AGEGRP=='2' | AGEGRP=='3' | AGEGRP=='4') %>%
  
  #create the timeframe using the Date codes
  mutate(timeframe=case_when(
    YEAR=='3' ~ '2010',
    YEAR=='4' ~ '2011',
    YEAR=='5' ~ '2012',
    YEAR=='6' ~ '2013',
    YEAR=='7' ~ '2014',
    YEAR=='8' ~ '2015',
    YEAR=='9' ~ '2016',
    YEAR=='10' ~ '2017',
    YEAR=='11' ~ '2018',
    YEAR=='12' ~ '2019',
    YEAR=='13' ~ '2020')) %>%
  subset(!is.na(timeframe)) %>%
  
  #assign population counts to numeric data
  mutate(TOT_POP=as.numeric(paste(TOT_POP))) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('COUNTY'='county_code','STATE'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county)) %>%
    
  group_by(timeframe,county) %>%
  summarise(totalpop=sum(TOT_POP)) %>%
  ungroup %>%

  rename(location=county)


totalpop_state <- totalpop_county %>%
  mutate(location=statename) %>%
  group_by(location,timeframe) %>%
  summarise(totalpop=sum(totalpop)) %>%
  ungroup 

totalpop_region <- totalpop_county %>%
  left_join(regionids,by=c('location'='county')) %>%
  group_by(region,timeframe) %>%
  summarise(totalpop=sum(totalpop)) %>%
  ungroup %>%
  rename(location=region)

totalpop <- totalpop_county %>%
  bind_rows(totalpop_state) %>%
  bind_rows(totalpop_region) %>%

 mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) 


pop_percentages <- pop %>%
  left_join(totalpop,by=c('timeframe'='timeframe','location'='location')) %>%
  mutate(data_percent=data/totalpop) %>%
  select(-c(data,totalpop)) %>%
  rename(data=data_percent) %>%
  mutate(dataformat='Percent')



pop_final <- pop %>%
  bind_rows(pop_percentages) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId)



####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(pop_final$locationid))>=1) {
  print(pop_final$location[is.na(pop_final$locationid)])
} else if (sum(is.na(pop_final$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(pop_final)

```

## COMMIT TO DATABASE (MONTANA AND SOUTH DAKOTA WITHOUT PLANNING REGIONS)
```{r}
#CHECK DATASET NAMED pop TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
pop2 <- pop_final %>% subset(locationtype != 'Planning Region')
dbWriteTable(con,database_state,pop2,append=TRUE,row.names=FALSE)
```

## COMMIT TO DATABASE (NORTH DAKOTA WITH PLANNING REGIONS)
```{r}
#CHECK DATASET NAMED pop TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,pop_final,append=TRUE,row.names=FALSE)
```

## OUTPUT DATASET TO UPLOAD TO DATA CENTER
```{r}
#write query from database to get needed format for KC data center

pop_sql <- paste0("SELECT locationid, location, timeframe, dataformat, race, data FROM ", database_state," WHERE vintageyear='",year,"' AND varname='childpopulationbyethnicity';")


#edit this to include estimates available prior to current vintage year
pop_sql_prioryears <- paste0("SELECT locationid, location, timeframe, dataformat, race, data FROM ", database_state," WHERE (vintageyear='2000' OR vintageyear='1990') AND varname='childpopulationbyethnicity';")

upload_data_pop1 <- dbGetQuery(con,pop_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)

upload_data_pop2 <- dbGetQuery(con,pop_sql_prioryears) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Race=race)




#combine current year and prior years to replace all data on data center
upload_data_pop <- upload_data_pop1 %>%
  bind_rows(upload_data_pop2) 


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_pop,file=paste0("../Output/demographics/",database_state,"_",year,"_childpopulationbyethnicity.csv"),row.names=FALSE)
```

