---
title: "Small Area Health Insurance Estimates"
author: "Xanna Burg"
date: "3/27/2020"
output: html_document
---

## Indicator 1: Uninsured children ages 0 to 18 (annual estimates)
## Indicator 2: Uninsured children ages 0 to 18 by income (annual estimates)

**Created by:** Xanna Burg
**Date:** April 2020
**Updated by:**

**Data Source:** U.S. Census Bureau, Small Area Health Insurance Estimates
https://www.census.gov/programs-surveys/sahie.html 
**Purpose:** Connect to Census SAHIE data, clean data, and output dataset to upload to KIDS COUNT Data Center.

**Data format:** Final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Category (for income groups; character) Data (numeric: number, percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)

A secondary output is an Excel file to copy to the Margin of Error template in order to report margin of error for the estimates.

**To use this code for a new year:**
* Update the year in the third code chunk for variable 'year'
* Update the state name (exactly as appears in FIPS) of interest
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}
#install required packages the first time you use this code
#install.packages('tidyverse')
#install.packages('tidycensus')
#install.packages('censusapi')
#install.packages('stringr')
#install.packages('gt')

#load required packages
library(tidyverse)
library(tidycensus)
library(censusapi)
library(stringr)
library(gt)
```


```{r}
#metadata for available variables
sahie_vars <- listCensusMetadata(name="timeseries/healthins/sahie",
                               type="variables")
#view(sahie_vars)

#metadata for available geographies
sahie_geos <- listCensusMetadata(name="timeseries/healthins/sahie",
                               type="geography")
#view(sahie_geos)

#county and state names that match fips codes (from tidycensus package)
fips <- fips_codes

```


```{r}
#CHANGE THE OBJECTS 'statename' and 'year' TO MATCH THE STATE AND YEAR
#state should match exactly as is in FIPS code in order for code to correctly run
statename <- "South Dakota"
year <- "2018"


#run this code to create an object needed for the database (DO NOT EDIT)
if (statename=='Montana') {
  database_state <- 'montana'
} else if (statename=='North Dakota') {
  database_state <- 'northdakota'
} else if (statename=='South Dakota') {
  database_state <- 'southdakota'
}

#import the location ids matching the correct state (for MT, ND, and SD; DO NOT EDIT)
locationids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Location IDs.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Location IDs.csv")
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Location IDs.csv")
}
locationids$Location <- as.character(locationids$Location) #assign as character instead of factor for merging

#import the region match file matching the correct state (for MT, ND, and SD; DO NOT EDIT)
regionids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Region List.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Region List.csv") 
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Region List.csv")
}
regionids$county <- as.character(regionids$county)
regionids$region <- as.character(regionids$region)


#UPDATE THE BELOW FILE PATH IF USING THIS CODE OUTSIDE OF MT, ND, OR SD
#the csv file should have 2 columns: "LocationId" and "Location"
#locationids <- read.csv("your_file_path_here.csv")


#RUN THIS CODE, BUT NOT REQUIRED TO CHANGE ANYTHING
#the api will subset data to each state based on the state FIPS code: MT=30, North Dakota=38, South Dakota=46
statecode <- as.numeric(unique(fips$state_code[fips$state_name==statename]))


```


**NUMBER AND PERCENT OF CHILDREN UNDER AGE 19 WITHOUT HEALTH INSURANCE**

```{r}
##UNDER THE getCensus() FUNCTION, ADD YOUR OWN CENSUS API KEY IN ORDER TO USE THIS CODE. OR FOLLOW THE INSTRUCTIONS ON THE PACKAGE DOCUMENTATION TO SET YOUR API KEY IN YOUR R ENVIRONMENT
#https://cran.r-project.org/web/packages/censusapi/vignettes/getting-started.html


#timeseries/healthins/sahie table is: Time Series Small Area Health Insurance Estimates
#see lookup table for categorical vars https://www2.census.gov/programs-surveys/sahie/technical-documentation/file-layouts/sahie-file-layout-2008-2017.pdf


#COUNTY DATA
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_county <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4", 
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #assign location type to county
  mutate(locationtype='County') %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%

  #remove unneeded columns
  select(-c(starts_with("NUI")),starts_with(("NIC"))) %>%
  
  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,percentuninsured_under18,numberuninsured_under18,county,state_name,locationtype)) %>%
  rename(timeframe=YEAR,
         location=county,
         state=state_name) %>%
  
  #assign variable name needed for database
  mutate(varname='uninsuredchildrenages0to18sahie') %>%
  
  #stack number and percent - wide data to long data
  pivot_longer(cols=c('numberuninsured_under18','percentuninsured_under18'),names_to='dataformat',values_to='data') %>%
  mutate(dataformat=case_when(
    dataformat=='numberuninsured_under18' ~ 'Number',
    dataformat=='percentuninsured_under18' ~ 'Percent')) 

```

```{r}
#STATE DATA
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_state <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT"),
          region=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #assign location type to state
  mutate(locationtype='State') %>%
  
  #add in state name
  mutate(location=statename) %>%
  mutate(state_name=statename) %>%

  #remove unneeded columns
  select(-c(starts_with("NUI")),starts_with(("NIC"))) %>%
  

  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,percentuninsured_under18,numberuninsured_under18,location,state_name,locationtype)) %>%
  rename(timeframe=YEAR,
         state=state_name) %>%
  
  #assign variable name needed for database
  mutate(varname='uninsuredchildrenages0to18sahie') %>%
  
  #stack number and percent - wide data to long data
  pivot_longer(cols=c('numberuninsured_under18','percentuninsured_under18'),names_to='dataformat',values_to='data') %>%
  mutate(dataformat=case_when(
    dataformat=='numberuninsured_under18' ~ 'Number',
    dataformat=='percentuninsured_under18' ~ 'Percent')) 
```

```{r}
#STATE PLANNING REGION
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_region <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%
  
  #create state planning region column
  mutate(county=replace(county, 
                          statename=='Montana' & county=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(county=replace(county, 
                          statename=='South Dakota' & county=='Shannon', 
                          'Oglala Lakota')) %>%
  
  left_join(regionids,by=c('county'='county'))  %>%
  group_by(.,region,YEAR) %>%
  summarise(numberuninsured_under18=sum(NUI_PT),
                        percentuninsured_under18=
                  (sum(NUI_PT))/((sum(NUI_PT)+sum(NIC_PT)))) %>%
  subset(region!="") %>%
  mutate(locationtype='State Planning Region') %>%

  #assign location type to region
  mutate(locationtype='Planning Region') %>%
  mutate(state_name=statename) %>%
  
  #remove unneeded columns
  select(-c(starts_with("NUI")),starts_with(("NIC"))) %>%
  

  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,percentuninsured_under18,numberuninsured_under18,region,state_name,locationtype)) %>%
  rename(timeframe=YEAR,
         location=region,
         state=state_name) %>%
  
  #assign variable name needed for database
  mutate(varname='uninsuredchildrenages0to18sahie') %>%
  
  #stack number and percent - wide data to long data
  pivot_longer(cols=c('numberuninsured_under18','percentuninsured_under18'),names_to='dataformat',values_to='data') %>%
  mutate(dataformat=case_when(
    dataformat=='numberuninsured_under18' ~ 'Number',
    dataformat=='percentuninsured_under18' ~ 'Percent')) 

```

```{r}
#UNION COUNTY, STATE, AND STATE PLANNING REGION
sahie <- sahie_county %>%
  bind_rows(sahie_state) %>%
  bind_rows(sahie_region) %>%

  #merge in location ids
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(sahie$locationid))>=1) {
  print(sahie$location[is.na(sahie$locationid)])
} else if (sum(is.na(sahie$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- sahie %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
view(sahie)
```

                  
```{r}
#CHECK DATASET NAMED sahie TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,sahie,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

sahie_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",year,"' AND varname='uninsuredchildrenages0to18sahie';")
#remove the year from the sql WHERE clause if running all available years

upload_data_sahie <- dbGetQuery(con,sahie_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_sahie,file=paste0("../Output/health/",database_state,"_",year,"_uninsuredchildrenages0to18sahie.csv"),row.names=FALSE)
```


```{r}
#IF NOT USING A DATABASE, REMOVE THE # TO UN-COMMENT AND THEN RUN THIS CODE CHUNK TO FORMAT A .CSV FOR UPLOADING TO THE KIDS COUNT DATA CENTER

#upload_kidscount_sahie <- sahie %>%
  #select(c(location,locationid,timeframe,dataformat,data)) %>%
  #rename(Location=location,
         #LocationId=location,
         #TimeFrame=timeframe,
         #DataFormat=dataformat,
         #Data=data)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
#write.csv(upload_kidscount_sahie,file=("Add_your_file_path/file_name_here.csv"),row.names=FALSE)

```

            
            
**NUMBER AND PERCENT OF CHILDREN UNDER AGE 19 WITHOUT HEALTH INSURANCE**
**MARGIN OF ERROR ESTIMATES DATA FILE**
```{r}
#CREATE A DATA FILE THAT CAN BE PUBLISHED WITH MARGIN OF ERROR ESTIMATES

#COUNTY DATA WITH MARGIN OF ERROR
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_county_moe <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","NUI_MOE","PCTUI_MOE"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #calculate the margin of error estimates
  mutate(numberuninsured_under18_moe=NUI_MOE) %>%
  mutate(percentuninsured_under18_moe=PCTUI_MOE/100) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%

  #select only columns of interest
  select(c(YEAR,county,numberuninsured_under18,numberuninsured_under18_moe,
           percentuninsured_under18,percentuninsured_under18_moe)) %>%
  
  #rename to KIDS COUNT requirements 
  rename(timeframe=YEAR,
         location=county) 
  
```

```{r}
#STATE DATA
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_state_moe <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","NUI_MOE","PCTUI_MOE"),
          region=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #calculate margin of error estimates
  mutate(numberuninsured_under18_moe=NUI_MOE) %>%
  mutate(percentuninsured_under18_moe=PCTUI_MOE/100) %>%
  
  #add in state name
  mutate(location=statename) %>%

  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,location,numberuninsured_under18,numberuninsured_under18_moe,
           percentuninsured_under18,percentuninsured_under18_moe)) %>%
  rename(timeframe=YEAR) 
```

```{r}
#state planning region estimates are derived estimates

#STATE PLANNING REGION
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_region_moe <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","NUI_MOE","PCTUI_MOE","NIC_MOE"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%
  
  #create state planning region column
  mutate(county=replace(county, 
                          statename=='Montana' & county=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(county=replace(county, 
                          statename=='South Dakota' & county=='Shannon', 
                          'Oglala Lakota')) %>%
  
  left_join(regionids,by=c('county'='county'))  %>%
  group_by(.,region,YEAR) %>%
  summarise(numberuninsured_under18=sum(NUI_PT),
            percentuninsured_under18=(sum(NUI_PT))/((sum(NUI_PT)+sum(NIC_PT))),
            numberuninsured_under18_moe=sqrt(sum((NUI_MOE^2))),
            calculationa=(1/sum(sum(NUI_PT)+sum(NIC_PT))),
            moe_x=sqrt((sum(NUI_MOE^2)))^2,
            calculationb=((sum(NUI_PT))/((sum(NUI_PT)+sum(NIC_PT))))^2,
            moe_y=(sqrt(sum(NUI_MOE^2)+sum(NIC_MOE^2)))^2,
            percentuninsured_under18_moe=
              calculationa*sqrt(moe_x-(calculationb*moe_y))) %>%
  subset(region!="") %>%


  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,region,numberuninsured_under18,numberuninsured_under18_moe,
           percentuninsured_under18,percentuninsured_under18_moe)) %>%
  rename(timeframe=YEAR,
         location=region)
```

```{r}
#COMBINE COUNTY, STATE, AND STATE PLANNING REGION INTO OUTPUT DATA FILE
sahie_moe <- sahie_county_moe %>%
  arrange(location) %>%
  bind_rows(sahie_state_moe) %>%
  bind_rows(sahie_region_moe) %>%
  rename(Location=location,
         Year=timeframe,
         '# Estimate'=numberuninsured_under18,
         '# Margin of Error (+/-)'=numberuninsured_under18_moe,
         '% Estimate'=percentuninsured_under18,
         '% Margin of Error (+/-)'=percentuninsured_under18_moe) 


#format a nice table in html for moe
#moe_table <- sahie_moe %>%
#  gt() %>%
#  tab_header(title=md(paste0("**Youth ages 0 to 18 who are uninsured in ",year," (annual estimates)**"))) %>%
#  tab_spanner(label="Number Uninsured",
#              columns=vars('# Estimate','# Margin of Error (+/-)')) %>%
#  tab_spanner(label="Percent Uninsured",
#              columns=vars('% Estimate','% Margin of Error (+/-)')) %>%
#  fmt_percent(columns=vars('% Estimate','% Margin of Error (+/-)'),decimals=1) %>%
#  fmt_number(columns=vars('# Estimate','# Margin of Error (+/-)'),decimals=0)

#view the table
#moe_table



#output to csv
write.csv(sahie_moe,paste0("../Output/health/moe/",database_state,"_uninsuredchildrenages0to18sahie_moe_",year,".csv"),row.names=FALSE)


```















**NUMBER AND PERCENT OF CHILDREN UNDER AGE 19 WITHOUT HEALTH INSURANCE BY INCOME LEVEL**

```{r}
##UNDER THE getCensus() FUNCTION, ADD YOUR OWN CENSUS API KEY IN ORDER TO USE THIS CODE. OR FOLLOW THE INSTRUCTIONS ON THE PACKAGE DOCUMENTATION TO SET YOUR API KEY IN YOUR R ENVIRONMENT
#https://cran.r-project.org/web/packages/censusapi/vignettes/getting-started.html


#timeseries/healthins/sahie table is: Time Series Small Area Health Insurance Estimates
#see lookup table for categorical vars https://www2.census.gov/programs-surveys/sahie/technical-documentation/file-layouts/sahie-file-layout-2008-2017.pdf


#COUNTY DATA
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
#MATCHING ALL AVAILALBE INCOME CATEGORIES
sahie_county_income <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","IPRCAT","IPR_DESC"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 

   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #assign location type to county
  mutate(locationtype='County') %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%

  #remove unneeded columns
  select(-c(starts_with("NUI")),starts_with(("NIC"))) %>%
  
  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,IPR_DESC,percentuninsured_under18,numberuninsured_under18,county,state_name,locationtype)) %>%
  rename(timeframe=YEAR,
         location=county,
         state=state_name,
         category=IPR_DESC) %>%
  
  #assign variable name needed for database
  mutate(varname='uninsuredchildrenages0to18byincomesahie') %>%
  
  #stack number and percent - wide data to long data
  pivot_longer(cols=c('numberuninsured_under18','percentuninsured_under18'),names_to='dataformat',values_to='data') %>%
  mutate(dataformat=case_when(
    dataformat=='numberuninsured_under18' ~ 'Number',
    dataformat=='percentuninsured_under18' ~ 'Percent')) 

```

```{r}
#STATE DATA
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
#MATCHING ALL AVAILALBE INCOME CATEGORIES
sahie_state_income <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","IPRCAT","IPR_DESC"),
          region=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #assign location type to state
  mutate(locationtype='State') %>%
  
  #add in state name
  mutate(location=statename) %>%
  mutate(state_name=statename) %>%

  #remove unneeded columns
  select(-c(starts_with("NUI")),starts_with(("NIC"))) %>%
  

  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,IPR_DESC,percentuninsured_under18,numberuninsured_under18,location,state_name,locationtype)) %>%
  rename(timeframe=YEAR,
         state=state_name,
         category=IPR_DESC) %>%
  
  #assign variable name needed for database
  mutate(varname='uninsuredchildrenages0to18byincomesahie') %>%
  
  #stack number and percent - wide data to long data
  pivot_longer(cols=c('numberuninsured_under18','percentuninsured_under18'),names_to='dataformat',values_to='data') %>%
  mutate(dataformat=case_when(
    dataformat=='numberuninsured_under18' ~ 'Number',
    dataformat=='percentuninsured_under18' ~ 'Percent')) 
```

```{r}
#STATE PLANNING REGION
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
#MATCHING ALL AVAILABLE INCOME CATEGORIES
sahie_region_income <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","IPRCAT","IPR_DESC"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%
  
  #create state planning region column
  mutate(county=replace(county, 
                          statename=='Montana' & county=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(county=replace(county, 
                          statename=='South Dakota' & county=='Shannon', 
                          'Oglala Lakota')) %>%
  
  left_join(regionids,by=c('county'='county'))  %>%
  group_by(.,region,YEAR,IPR_DESC) %>%
  summarise(numberuninsured_under18=sum(NUI_PT),
                        percentuninsured_under18=
                  (sum(NUI_PT))/((sum(NUI_PT)+sum(NIC_PT)))) %>%
  subset(region!="") %>%
  mutate(locationtype='State Planning Region') %>%

  #assign location type to region
  mutate(locationtype='Planning Region') %>%
  mutate(state_name=statename) %>%
  
  #remove unneeded columns
  select(-c(starts_with("NUI")),starts_with(("NIC"))) %>%
  

  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,IPR_DESC,percentuninsured_under18,numberuninsured_under18,region,state_name,locationtype)) %>%
  rename(timeframe=YEAR,
         location=region,
         state=state_name,
         category=IPR_DESC) %>%
  
  #assign variable name needed for database
  mutate(varname='uninsuredchildrenages0to18byincomesahie') %>%
  
  #stack number and percent - wide data to long data
  pivot_longer(cols=c('numberuninsured_under18','percentuninsured_under18'),names_to='dataformat',values_to='data') %>%
  mutate(dataformat=case_when(
    dataformat=='numberuninsured_under18' ~ 'Number',
    dataformat=='percentuninsured_under18' ~ 'Percent')) 

```

```{r}
#UNION COUNTY, STATE, AND STATE PLANNING REGION
sahie_income <- sahie_county_income %>%
  bind_rows(sahie_state_income) %>%
  bind_rows(sahie_region_income) %>%

  #merge in location ids
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(sahie_income$locationid))>=1) {
  print(sahie$location[is.na(sahie_income$locationid)])
} else if (sum(is.na(sahie_income$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- sahie_income %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
view(sahie_income)
```

```{r}
#CHECK DATASET NAMED sahie_income TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,sahie_income,append=TRUE,row.names=FALSE)
```

```{r}
#write query from database to get needed format for KC data center

sahie_income_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data, category FROM ", database_state," WHERE timeframe='",year,"' AND varname='uninsuredchildrenages0to18byincomesahie';")
#remove the year from the sql WHERE clause if running all available years

upload_data_sahie_income <- dbGetQuery(con,sahie_income_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data,
         Category=category)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_sahie_income,file=paste0("../Output/health/",database_state,"_",year,"_uninsuredchildrenages0to18byincomesahie.csv"),row.names=FALSE)
```





**NUMBER AND PERCENT OF CHILDREN UNDER AGE 19 WITHOUT HEALTH INSURANCE BY INCOME**
**MARGIN OF ERROR ESTIMATES DATA FILE**
```{r}
#CREATE A DATA FILE THAT CAN BE PUBLISHED WITH MARGIN OF ERROR ESTIMATES

#COUNTY DATA WITH MARGIN OF ERROR
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
#INCLUDE ALL INCOME GROUPS
sahie_county_income_moe <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","NUI_MOE","PCTUI_MOE",
                 "IPRCAT","IPR_DESC"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #calculate the margin of error estimates
  mutate(numberuninsured_under18_moe=NUI_MOE) %>%
  mutate(percentuninsured_under18_moe=PCTUI_MOE/100) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%

  #select only columns of interest
  select(c(YEAR,IPR_DESC,county,numberuninsured_under18,numberuninsured_under18_moe,
           percentuninsured_under18,percentuninsured_under18_moe)) %>%
  
  #rename to KIDS COUNT requirements 
  rename(timeframe=YEAR,
         location=county,
         category=IPR_DESC) 
  
```

```{r}
#STATE DATA
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
#ALL INCOME CATEGORIES
sahie_state_income_moe <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","NUI_MOE","PCTUI_MOE",
                 "IPRCAT","IPR_DESC"),
          region=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  mutate(percentuninsured_under18=PCTUI_PT/100) %>% 
  mutate(numberuninsured_under18=NUI_PT) %>%  
  
  #calculate margin of error estimates
  mutate(numberuninsured_under18_moe=NUI_MOE) %>%
  mutate(percentuninsured_under18_moe=PCTUI_MOE/100) %>%
  
  #add in state name
  mutate(location=statename) %>%

  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,IPR_DESC,location,numberuninsured_under18,numberuninsured_under18_moe,
           percentuninsured_under18,percentuninsured_under18_moe)) %>%
  rename(timeframe=YEAR,
         category=IPR_DESC) 
```

```{r}
#state planning region estimates are derived estimates

#STATE PLANNING REGION
#MATCHING AGE CATEGORY 4 FOR 19 YEARS AND YOUNGER
sahie_region_income_moe <- getCensus(name="timeseries/healthins/sahie",
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("PCTUI_PT","NUI_PT","NIC_PT","NUI_MOE","PCTUI_MOE","NIC_MOE",
                 "IPRCAT","IPR_DESC"),
          region="county:*",
          regionin=paste0("state:",statecode),
          AGECAT="4",
          YEAR=year) %>% # remove YEAR=year to include all available years
 
   #clean the dataset that's been imported
  
  mutate_at(vars(contains("NIC")), as.numeric) %>%
  mutate_at(vars(contains("NUI")), as.numeric) %>%
  mutate_at(vars(contains("PCTUI")), as.numeric) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%
  
  #create state planning region column
  mutate(county=replace(county, 
                          statename=='Montana' & county=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(county=replace(county, 
                          statename=='South Dakota' & county=='Shannon', 
                          'Oglala Lakota')) %>%
  
  left_join(regionids,by=c('county'='county'))  %>%
  group_by(.,region,YEAR,IPR_DESC) %>%
  summarise(numberuninsured_under18=sum(NUI_PT),
            percentuninsured_under18=(sum(NUI_PT))/((sum(NUI_PT)+sum(NIC_PT))),
            numberuninsured_under18_moe=sqrt(sum((NUI_MOE^2))),
            calculationa=(1/sum(sum(NUI_PT)+sum(NIC_PT))),
            moe_x=sqrt((sum(NUI_MOE^2)))^2,
            calculationb=((sum(NUI_PT))/((sum(NUI_PT)+sum(NIC_PT))))^2,
            moe_y=(sqrt(sum(NUI_MOE^2)+sum(NIC_MOE^2)))^2,
            percentuninsured_under18_moe=
              calculationa*sqrt(moe_x-(calculationb*moe_y))) %>%
  subset(region!="") %>%


  #select only needed variables and rename to KIDS COUNT requirements
  select(c(YEAR,IPR_DESC,region,numberuninsured_under18,numberuninsured_under18_moe,
           percentuninsured_under18,percentuninsured_under18_moe)) %>%
  rename(timeframe=YEAR,
         location=region,
         category=IPR_DESC)
```

```{r}
#COMBINE COUNTY, STATE, AND STATE PLANNING REGION INTO OUTPUT DATA FILE
sahie_income_moe <- sahie_county_income_moe %>%
  arrange(location) %>%
  bind_rows(sahie_state_income_moe) %>%
  bind_rows(sahie_region_income_moe) %>%
  rename(Location=location,
         Year=timeframe,
         '# Estimate'=numberuninsured_under18,
         '# Margin of Error (+/-)'=numberuninsured_under18_moe,
         '% Estimate'=percentuninsured_under18,
         '% Margin of Error (+/-)'=percentuninsured_under18_moe) %>%
 # select(-c(Year))
  
  #structure for easy import into template
  pivot_wider(id_cols=c(Year,category,Location),names_from=category,
              values_from = c('# Estimate','# Margin of Error (+/-)',
                              '% Estimate','% Margin of Error (+/-)'))

sahie_income_moe_ordered <- sahie_income_moe[c("Year","Location","# Estimate_All Incomes", "# Margin of Error (+/-)_All Incomes", "% Estimate_All Incomes", "% Margin of Error (+/-)_All Incomes", "# Estimate_<= 138% of Poverty", "# Margin of Error (+/-)_<= 138% of Poverty", "% Estimate_<= 138% of Poverty", "% Margin of Error (+/-)_<= 138% of Poverty", "# Estimate_<= 200% of Poverty", "# Margin of Error (+/-)_<= 200% of Poverty", "% Estimate_<= 200% of Poverty", "% Margin of Error (+/-)_<= 200% of Poverty", "# Estimate_<= 250% of Poverty","# Margin of Error (+/-)_<= 250% of Poverty","% Estimate_<= 250% of Poverty","% Margin of Error (+/-)_<= 250% of Poverty","# Estimate_<= 400% of Poverty", "# Margin of Error (+/-)_<= 400% of Poverty", "% Estimate_<= 400% of Poverty","% Margin of Error (+/-)_<= 400% of Poverty", "# Estimate_138% to 400% of Poverty","# Margin of Error (+/-)_138% to 400% of Poverty","% Estimate_138% to 400% of Poverty","% Margin of Error (+/-)_138% to 400% of Poverty")]




#output to csv
write.csv(sahie_income_moe_ordered,paste0("../Output/health/moe/",database_state,"_uninsuredchildrenages0to18sahiebyincome_moe_",year,".csv"),row.names=FALSE)


```



