---
title: "Total population"
author: "Xanna Burg"
output: html_document
---


## Indicator: Total population (all ages)
**Created by:** Xanna Burg
**Date:** April 2020
**Updated by:**

**Data Source:** U.S. Census Bureau, Annual Population Estimates. Specific table: pep/charagegroups 
**Purpose:** Connect to census data table, clean data, add to master database, output csv to upload to Data Center in correct format. Update Montana, North Dakota, and South Dakota KIDS COUNT Data Center indicator with most recent year of vintage data.
**Description:** this program calculates **Total population (all ages)**
* Geographies: state total, county, state planning regions (where applicable, which are groups of counties)
* Number - count of population


**Data format:** final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (numeric: year), Data (numeric: number or percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)



```{r,message=FALSE}
#load required packages
library(tidyverse)
library(tidycensus)
library(censusapi)
library(stringr)
```

```{r}
#metadata for available variables
#pop_vars <- listCensusMetadata(name="pep/charagegroups",
 #                             vintage=year,
  #                             type="variables")
#view(pop_vars)

#metadata for available geographies
#pop_geos <- listCensusMetadata(name="pep/charagegroups",
                                #vintage=year,
                               #type="geography")
#view(pop_geos)

#county and state names that match fips codes (from tidycensus package)
fips <- fips_codes %>%
  mutate(county_code=as.numeric(paste(county_code))) %>%
  mutate(state_code=as.numeric(paste(state_code)))
#fips$county_code = str_remove(fips$county_code, "^0+")

#import tribal fips code (from file on local drive)
fips_tribal <- read.csv("../Input/AIANNHCE Codes.csv")

```


```{r}
#CHANGE THE OBJECTS 'statename' and 'year' TO MATCH THE STATE AND YEAR
#state should match exactly as is in FIPS code in order for code to correctly run
statename <- "South Dakota"
year <- "2022"



#run this code to create an object needed for the database (DO NOT EDIT)
if (statename=='Montana') {
  database_state <- 'montana'
} else if (statename=='North Dakota') {
  database_state <- 'northdakota'
} else if (statename=='South Dakota') {
  database_state <- 'southdakota'
}

#import the location ids matching the correct state (for MT, ND, and SD; DO NOT EDIT)
locationids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Location IDs.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Location IDs.csv")
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Location IDs.csv")
}
locationids$Location <- as.character(locationids$Location) #assign as character instead of factor for merging

#import the region match file matching the correct state (for MT, ND, and SD; DO NOT EDIT)
regionids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Region List.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Region List.csv") 
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Region List.csv")
}
regionids$county <- as.character(regionids$county)
regionids$region <- as.character(regionids$region)

#RUN THIS CODE, BUT NOT REQUIRED TO CHANGE ANYTHING
#the api will subset data to each state based on the state FIPS code: MT=30, North Dakota=38, South Dakota=46
statecode <- as.numeric(unique(fips$state_code[fips$state_name==statename]))
```


## STEP 1: IMPORT DATA AND CLEAN 
```{r}
############
#COUNTY DATA
#pop_county <- getCensus(name="pep/charagegroups",
 #         vintage=year,
  #        key=Sys.getenv("CENSUS_API_KEY"),
   #       vars=c("POP","DATE_CODE","DATE_DESC","AGEGROUP"),
    #      region="county:*",
     #     regionin=paste0("state:",statecode)) %>%
  
pop_county <- read.csv(paste0("../Input/demographics/CC-EST",year,"-ALLDATA.csv")) %>%
  
  subset(STATE==statecode) %>%
  
  #include all ages
  subset(AGEGRP=='0') %>%
  
  #clean the data that's been imported
  
  #assign population counts to numeric data
  mutate(POP=as.numeric(TOT_POP)) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  left_join(fips,by=c('COUNTY'='county_code','STATE'='state_code')) %>% 
  mutate(county=gsub("\\s*\\w*$", "", county)) %>%
  
  #create the timeframe using the Date codes- 2020 file
  #mutate(timeframe=case_when(
    #YEAR=='3' ~ '2010',
    #YEAR=='4' ~ '2011',
    #YEAR=='5' ~ '2012',
    #YEAR=='6' ~ '2013',
    #YEAR=='7' ~ '2014',
    #YEAR=='8' ~ '2015',
    #YEAR=='9' ~ '2016',
    #YEAR=='10' ~ '2017',
    #YEAR=='11' ~ '2018',
    #YEAR=='12' ~ '2019',
    #YEAR=='13' ~ '2020')) %>%
  
   #create the timeframe using the Date codes- 2021 file
  mutate(timeframe=case_when(
    YEAR=='2' ~ '2020',
    YEAR=='3' ~ '2021',
    YEAR=='4' ~ '2022')) %>%
  
  subset(!is.na(timeframe)) %>%
  
  #select Kids Count vars
  select(c(county,timeframe,state_name,POP)) %>%
  rename(location=county,
         state=state_name,
         data=POP) %>%
  
  mutate(locationtype='County') %>%
  mutate(vintageyear=year) %>%
  mutate(dataformat='Number') %>%
  mutate(varname='totalpopulation') %>%
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) 

###########
#STATE DATA
pop_state <- pop_county %>%
  mutate(location=statename) %>%
  group_by(location,timeframe,state,vintageyear,dataformat,varname) %>%
  summarise(data=sum(data)) %>%
  ungroup %>%
  mutate(locationtype='State')



#pop_state <- getCensus(name="pep/charagegroups",
 #         vintage=year,
  #        key=Sys.getenv("CENSUS_API_KEY"),
   #       vars=c("POP","DATE_CODE","DATE_DESC","AGEGROUP"),
    #      region=paste0("state:",statecode)) %>%
  
  #include all ages
  #subset(AGEGROUP=='0') %>%

  #clean the data that's been imported
  
  #assign population counts to numeric data
#  mutate(POP=as.numeric(POP)) %>%
  
  #create the timeframe using the Date codes
 # mutate(timeframe=case_when(
  #  DATE_CODE=='1' ~ '2010',
   # DATE_CODE=='4' ~ '2011',
  #  DATE_CODE=='5' ~ '2012',
   # DATE_CODE=='6' ~ '2013',
  #  DATE_CODE=='7' ~ '2014',
   # DATE_CODE=='8' ~ '2015',
  #  DATE_CODE=='9' ~ '2016',
   # DATE_CODE=='10' ~ '2017',
  #  DATE_CODE=='11' ~ '2018',
   # DATE_CODE=='12' ~ '2019')) %>%
  
  #subset(!is.na(timeframe)) %>%
  
  #select Kids Count vars
  #select(c(timeframe,POP)) %>%
  #rename(data=POP) %>%
  
  #mutate(location=statename) %>%
  #mutate(state=statename) %>%
  #mutate(locationtype='State') %>%
  #mutate(vintageyear=year) %>%
  #mutate(dataformat='Number') %>%
  #mutate(varname='totalpopulation')

############
#REGION DATA
pop_region <- pop_county %>%
  left_join(regionids,by=c('location'='county')) %>% 
  select(-c(location,locationtype)) %>%
  rename(location=region) %>%
  group_by(location,timeframe,state,vintageyear,dataformat,varname) %>%
  summarise(data=sum(data)) %>%
  ungroup %>%
  mutate(locationtype='Planning Region') %>%
  #remove unmatched regions (not all counties map to a region in South Dakota)
  filter(location!='') 
  
  
  
#pop_region <- getCensus(name="pep/charagegroups",
 #         vintage=year,
  #        key=Sys.getenv("CENSUS_API_KEY"),
   #       vars=c("POP","DATE_CODE","DATE_DESC","AGEGROUP"),
    #      region="county:*",
     #     regionin=paste0("state:",statecode)) %>%
  
  #include all ages
  #subset(AGEGROUP=='0') %>%

  #clean the data that's been imported
  
  #assign population counts to numeric data
  #mutate(POP=as.numeric(POP)) %>%
  
  #add in county name from fips codes, remove word 'county', add in state name
  #left_join(fips,by=c('county'='county_code','state'='state_code')) %>% 
  #mutate(county=gsub("\\s*\\w*$", "", county.y)) %>%
  
  
  #create state planning region column
  #mutate(county=replace(county, 
   #                       statename=='Montana' & county=='Lewis and Clark', 
    #                      'Lewis & Clark')) %>%
  #mutate(county=replace(county, 
     #                     statename=='South Dakota' & county=='Shannon', 
      #                    'Oglala Lakota')) %>%
  
  #left_join(regionids,by=c('county'='county'))  %>%
  #group_by(region,DATE_CODE) %>%
  #summarise(data=sum(POP)) %>%
  
  #remove unmatched regions (not all counties map to a region in South Dakota)
  #filter(region!='') %>%
  
  
  #create the timeframe using the Date codes
  #mutate(timeframe=case_when(
    #DATE_CODE=='1' ~ '2010',
    #DATE_CODE=='4' ~ '2011',
    #DATE_CODE=='5' ~ '2012',
    #DATE_CODE=='6' ~ '2013',
    #DATE_CODE=='7' ~ '2014',
    #DATE_CODE=='8' ~ '2015',
    #DATE_CODE=='9' ~ '2016',
    #DATE_CODE=='10' ~ '2017',
    #DATE_CODE=='11' ~ '2018',
    #DATE_CODE=='12' ~ '2019')) %>%
  
  #subset(!is.na(timeframe)) %>%
  
  #select Kids Count vars
  #select(c(region,timeframe,data)) %>%
  #rename(location=region) %>%
  
  #mutate(state=statename) %>%
  #mutate(locationtype='Planning Region') %>%
  #mutate(vintageyear=year) %>%
  #mutate(dataformat='Number') %>%
  #mutate(varname='totalpopulation')

###########################################
#UNION COUNTY, STATE, STATE PLANNING REGION
pop <- pop_county %>%
  bind_rows(pop_state) %>%
  bind_rows(pop_region) %>%

  #merge in location ids
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(pop$locationid))>=1) {
  print(pop$location[is.na(pop$locationid)])
} else if (sum(is.na(pop$locationid))==0) {
  'all locations match'
}


# 2. Visually inspect output data
View(pop)

```

## STEP 2a: COMMIT TO SQL DATABASE (MONTANA & SOUTH DAKOTA WITHOUT PLANNING REGIONS)
```{r}
#CHECK DATASET NAMED pop TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
pop2 <- pop %>% subset(locationtype != 'Planning Region')
dbWriteTable(con,database_state,pop2,append=TRUE,row.names=FALSE)
```

## STEP 2b: COMMIT TO SQL DATABASE (NORTH DAKOTA WITH PLANNING REGIONS)
```{r}
#CHECK DATASET NAMED pop TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,pop,append=TRUE,row.names=FALSE)
```


## STEP 3a: OUTPUT TO DATA CENTER WITHOUT TRIBAL AREAS (MONTANA & SOUTH DAKOTA)
```{r}
#write query from database to get needed format for KC data center

pop_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE vintageyear='",year,"' AND varname='totalpopulation';")
#remove the year from the sql WHERE clause if running all available years

#edit this to include estimates available prior to current vintage year

#for intercensal estimates for 2000-2009
pop_sql_prioryears <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE vintageyear='2000' AND varname='totalpopulation' AND locationtype <> 'Planning Region';")

#for intercensal estimates for 2010-2019
pop_sql_prioryears2 <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE vintageyear='2020' AND varname='totalpopulation' AND locationtype <> 'Planning Region';")

upload_data_pop1 <- dbGetQuery(con,pop_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

upload_data_pop2 <- dbGetQuery(con,pop_sql_prioryears) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

upload_data_pop3 <- dbGetQuery(con,pop_sql_prioryears2) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data) %>%
  subset(TimeFrame != '2020')

#combine current year and prior years to replace all data on data center
upload_data_pop <- upload_data_pop1 %>%
  bind_rows(upload_data_pop2) %>%
  bind_rows(upload_data_pop3)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_pop,file=paste0("../Output/demographics/",database_state,"_",year,"_totalpopulation.csv"),row.names=FALSE)
```

## STEP 3b: OUTPUT TO DATA CENTER WITH TRIBAL AREAS (NORTH DAKOTA)
```{r}
#write query from database to get needed format for KC data center

pop_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE vintageyear='",year,"' AND varname='totalpopulation';")
#remove the year from the sql WHERE clause if running all available years

#edit this to include estimates available prior to current vintage year
#for intercensal estimates for 2000-2009
pop_sql_prioryears <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE vintageyear='2000' AND varname='totalpopulation';")

#for intercensal estimates for 2010-2019
pop_sql_prioryears2 <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE vintageyear='2020' AND varname='totalpopulation';")

upload_data_pop1 <- dbGetQuery(con,pop_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

upload_data_pop2 <- dbGetQuery(con,pop_sql_prioryears) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

upload_data_pop3 <- dbGetQuery(con,pop_sql_prioryears2) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data) %>%
  subset(TimeFrame != '2020')

#run the totalpopulationtribalarea for the matching year before adding tribal data to this query
acs_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE varname='totalpopulationtribalarea';")
#remove the year from the sql WHERE clause if running all available years

upload_data_acs <- dbGetQuery(con,acs_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

#combine current year and prior years to replace all data on data center
upload_data_pop <- upload_data_pop1 %>%
  bind_rows(upload_data_pop2) %>%
  bind_rows(upload_data_pop3) %>%
  bind_rows(upload_data_acs)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data_pop,file=paste0("../Output/demographics/",database_state,"_",year,"_totalpopulation.csv"),row.names=FALSE)
```


