---
title: "Uninsured Children - Tribal Areas"
author: "Xanna Burg"
date: "6/24/2020"
output: html_document
---

## Indicator 1: Uninsured children ages 0 to 18 (5-year estimates) - FOR TRIBAL AREAS ONLY

**Created by:** Xanna Burg
**Date:** June 2020
**Updated by:**

**Data Source:** U.S. Census Bureau, American Community Survey 5-year Estimates, Table B27001
**Purpose:** Connect to Census ACS data, clean data, and output dataset to upload to KIDS COUNT Data Center.

**Data format:** Final output csv to upload is in long format with the following variables: Location (character: name of location), TimeFrame (text: years), Data (numeric: number, percentage), DataFormat (character: "number" or "percent"), LocationId (numeric: assigned for KIDS COUNT system)

A secondary output is an Excel file to copy to the Margin of Error template in order to report margin of error for the estimates.

**To use this code for a new year:**
* Update the year and acsyear (5 year interval) in the third code chunk for variables 'year' and 'acsyear'
* Update the state name (exactly as appears in FIPS) of interest
* Check each dataset visually and through the report logs prior to commiting to the database.


```{r,message=FALSE}
#install required packages the first time you use this code
#install.packages('tidyverse')
#install.packages('tidycensus')
#install.packages('censusapi')
#install.packages('stringr')

#load required packages
library(tidyverse)
library(tidycensus)
library(censusapi)
library(stringr)
```

```{r}
#metadata for available variables
#acs_vars <- listCensusMetadata(name="2018/acs/acs5",
                               #type="variables")


#metadata for available geographies
#acs_geos <- listCensusMetadata(name="2018/acs/acs5",
                              # type="geography")


#county and state names that match fips codes (from tidycensus package)
fips <- fips_codes

#import tribal fips code (from file on local drive)
fips_tribal <- read.csv("../Input/AIANNHCE Codes.csv")

```

```{r}
#CHANGE THE OBJECTS 'statename' and 'year' TO MATCH THE STATE AND YEAR
#state should match exactly as is in FIPS code in order for code to correctly run
statename <- "North Dakota"
year <- "2022"
acsyear <- "2018-2022"


#run this code to create an object needed for the database (DO NOT EDIT)
if (statename=='Montana') {
  database_state <- 'montana'
} else if (statename=='North Dakota') {
  database_state <- 'northdakota'
} else if (statename=='South Dakota') {
  database_state <- 'southdakota'
}

#import the location ids matching the correct state (for MT, ND, and SD; DO NOT EDIT)
locationids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Location IDs.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Location IDs.csv")
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Location IDs.csv")
}
locationids$Location <- as.character(locationids$Location) #assign as character instead of factor for merging

#import the region match file matching the correct state (for MT, ND, and SD; DO NOT EDIT)
regionids <- if (statename=='Montana') {
  read.csv("../Input/MT KC Region List.csv")
} else if (statename=='North Dakota') {
  read.csv("../Input/ND KC Region List.csv") 
} else if (statename=='South Dakota') {
  read.csv("../Input/SD KC Region List.csv")
}
regionids$county <- as.character(regionids$county)
regionids$region <- as.character(regionids$region)


#UPDATE THE BELOW FILE PATH IF USING THIS CODE OUTSIDE OF MT, ND, OR SD TO ADD LOCATION IDS
#the csv file should have 2 columns: "LocationId" and "Location"
#locationids <- read.csv("your_file_path_here.csv")

#UPDATE THE BELOW FILE PATH IF USING THIS CODE OUTSIDE OF MT, ND, OR SD TO ADD REGIONS (GROUPS OF COUNTIES)
#the csv file should have 2 columns: "county" and "region"
#regionids <- read.csv("your_file_path_here.csv")


#RUN THIS CODE, BUT NOT REQUIRED TO CHANGE ANYTHING
#the api will subset data to each state based on the state FIPS code: MT=30, North Dakota=38, South Dakota=46
statecode <- as.numeric(unique(fips$state_code[fips$state_name==statename]))
```

## ################################################## ##
## UNINSURED CHILDREN AGES 0 TO 18 (5-YEAR ESTIMATES) ##
## ################################################## ##

## STEP 1: COMPILE DATA FOR TRIBAL AREAS, IF APPLICABLE TO STATE
```{r}
###########
#TRIBAL DATA
acs_tribal <- getCensus(name="acs/acs5",
          vintage=year,
          key=Sys.getenv("CENSUS_API_KEY"),
          vars=c("B27001_005E","B27001_008E","B27001_033E","B27001_036E",
                 "B27001_003E","B27001_006E","B27001_031E","B27001_034E",
                 "B27001_005M","B27001_008M","B27001_033M","B27001_036M",
                 "B27001_003M","B27001_006M","B27001_031M","B27001_034M"),
          region="american indian area/alaska native area/hawaiian home land (or part):*",
          regionin=paste0("state:",statecode)) %>%


   #clean the imported data

  #assign location type to county
  mutate(locationtype='Tribal Area') %>%
  
  #add in location name from fips codes
  mutate(american_indian_area_alaska_native_area_hawaiian_home_land_or_part=as.numeric(american_indian_area_alaska_native_area_hawaiian_home_land_or_part)) %>%
  left_join(fips_tribal,by=c('american_indian_area_alaska_native_area_hawaiian_home_land_or_part'='AIANNHCE')) %>% 
  
  
  #calculate the sums and percent
  #########################
  #for all children ages 0 to 18
  mutate(numerator=(B27001_005E+B27001_008E+B27001_033E+B27001_036E)) %>%
  mutate(numerator_moe=sqrt((B27001_005M^2)+(B27001_008M^2)+(B27001_033M^2)+(B27001_036M^2))) %>%
  mutate(denominator=B27001_003E+B27001_006E+B27001_031E+B27001_034E) %>%
  mutate(denominator_moe=sqrt((B27001_003M^2)+(B27001_006M^2)+(B27001_031M^2)+(B27001_034M^2))) %>%
  
  mutate(Number=numerator) %>%
  mutate(number_moe=numerator_moe) %>%
  mutate(Percent=numerator/denominator) %>%
  mutate(percent_moe=(1/denominator)*sqrt((numerator_moe^2)-((Percent^2)*(denominator_moe^2)))) %>%
  mutate(percent_moe=if_else(is.na(percent_moe),(1/denominator)*sqrt((numerator_moe^2)+((Percent^2)*(denominator_moe^2))),percent_moe)) %>%

 #calculate the relative standard error
  mutate(number_relativese=((number_moe/1.645)/Number)*100) %>%
  mutate(percent_relativese=((percent_moe/1.645)/Percent)*100) %>%
  mutate(keep=if_else(percent_relativese>30 | number_relativese>30,0,1)) %>%
 

  #select only needed variables and name to kids count database
  select(c(AIANNHNAME,Number,number_moe,Percent,percent_moe,keep)) %>%
  
  rename(location=AIANNHNAME) %>%
  mutate(state=statename) %>%
  mutate(locationtype='Tribal Area') %>%
  mutate(timeframe=acsyear) %>%
  mutate(varname='uninsuredchildrenages0to18_acstribal') %>%
  
  #change from factor to character for correctly updating below
  mutate(location=as.character(location)) %>%
  
  #rename the location to match kids count location - for locations that cross state lines, to clarify which portion is included
  #includes locations for all three states
  mutate(location=replace(location, 
                          state=='North Dakota' & location=='Standing Rock Reservation', 
                          'Standing Rock Reservation (North Dakota portion)')) %>%
  
    mutate(location=replace(location, 
                          state=='South Dakota' & location=='Standing Rock Reservation', 
                          'Standing Rock Reservation (South Dakota portion)')) %>%
   
  mutate(location=replace(location, 
                          state=='North Dakota' & 
                            location=='Turtle Mountain Reservation and Off-Reservation Trust Land', 
                          'Turtle Mountain Reservation and Off-Reservation Trust Land (North Dakota portion)')) %>%
  
  mutate(location=replace(location, 
                          state=='North Dakota' & 
                            location=='Lake Traverse Reservation and Off-Reservation Trust Land', 
                          'Lake Traverse Reservation and Off-Reservation Trust Land (North Dakota portion)')) %>%
  
   mutate(location=replace(location, 
                          state=='South Dakota' & location=='Lake Traverse Reservation and Off-Reservation Trust Land', 
                          'Lake Traverse Reservation and Off-Reservation Trust Land (South Dakota portion)')) %>%
  
  
  #for North Dakota Lake Traverse data is too small to be meaningful, remove
  filter(state!='North Dakota' | location!='Lake Traverse Reservation and Off-Reservation Trust Land (North Dakota portion)') %>%
  
  #for Montana Turtle Mountain is too small to have data, remove
  filter(state!='Montana' | location !='Turtle Mountain Reservation and Off-Reservation Trust Land') %>%
  
  #for South Dakota Turtle Mountain is too small to have data, remove
  filter(state!='South Dakota' | location !='Turtle Mountain Reservation and Off-Reservation Trust Land') %>%
  
  #for South Dakota Northern Cheyenne is too small to have data, remove
  filter(state!='South Dakota' | location !='Northern Cheyenne Indian Reservation and Off-Reservation Trust Land') %>%
  
  filter(state!='South Dakota' | location !='Shakopee Mdewakanton Sioux Community and Off-Reservation Trust Land') %>%
  
  
  #wide to long
  pivot_longer(cols=c("Percent","Number"),names_to='dataformat',values_to='data') %>%
  pivot_longer(cols=c("percent_moe","number_moe"),names_to='moe_values',values_to='moe') %>%
  
  mutate(moe_values2=case_when(
    moe_values=='number_moe' ~ 'Number',
    moe_values=='percent_moe' ~ 'Percent')) %>%
  subset(dataformat==moe_values2) %>%
  
  
  #replace unreliable estimates
  mutate(data=replace(data,keep==0,NA)) %>%
  mutate(moe=replace(moe,keep==0,NA)) %>%
  select(-c(keep,moe_values,moe_values2)) %>%

  #merge in location ids
  mutate(location=replace(location, 
                          statename=='Montana' & location=='Lewis and Clark', 
                          'Lewis & Clark')) %>%
  mutate(location=replace(location, 
                          statename=='South Dakota' & location=='Shannon', 
                          'Oglala Lakota')) %>%
  left_join(locationids,by=c('location'='Location')) %>%
  rename(locationid=LocationId) 
  

####################
####################
#DATA QUALITY CHECKS
### Before moving forward, check the output log for errors in addition to checking specifically for these quality checks

# 1. Print name of location that has a mismatched location ID
if (sum(is.na(acs_tribal$locationid))>=1) {
  print(acs_tribal$location[is.na(acs_tribal$locationid)])
} else if (sum(is.na(acs_tribal$locationid))==0) {
  'all locations match'
}

# 2. Output cases where percent data is greater than 1
temp_percheck <- acs_tribal %>%
  subset(dataformat=='Percent' & data>1)
temp_rows <- as.numeric(nrow(temp_percheck))

if (temp_rows==0) {
  'no percents greater than 1'
} else if (temp_rows>=1) {
  print(temp_percheck)
}

# 3. Visually inspect output data
View(acs_tribal)
```

## STEP 2: COMMIT TO DATABASE 
```{r}
#CHECK DATASET NAMED acs_tribal TO VERIFY DATA ISN'T MISSING AND APPEARS CORRECT
#this code adds kids count data to MBPC postgres database
#run postgres code (separate file) first
dbWriteTable(con,database_state,acs_tribal,append=TRUE,row.names=FALSE)
```

## STEP 3a: OUTPUT FILES - INCLUDING TRIBAL AREAS (NORTH DAKOTA)
```{r}
#########################
##OUTPUT DATA CENTER FILE

#with tribal data (North Dakota)
datacenter_sql <- paste0("SELECT locationid, location, timeframe, dataformat, data FROM ", database_state," WHERE timeframe='",acsyear,"' AND varname='uninsuredchildrenages0to18_acstribal';")

upload_data <- dbGetQuery(con,datacenter_sql) %>%
  rename(LocationId=locationid,
         Location=location,
         TimeFrame=timeframe,
         DataFormat=dataformat,
         Data=data)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(upload_data,file=paste0("../Output/health/",database_state,"_",year,"_uninsuredchildrenages0to18_acstribal.csv"),row.names=FALSE)


#############################
##OUTPUT MARGIN OF ERROR FILE
moe_sql <- paste0("SELECT location, timeframe, dataformat, data, moe FROM ", database_state," WHERE timeframe='",acsyear,"' AND varname='uninsuredchildrenages0to18_acstribal' AND locationtype='Tribal Area';")

#alphabetize location
moe_output <- dbGetQuery(con,moe_sql) %>%
  pivot_wider(names_from=dataformat,values_from=c(data,moe)) %>%
  select(timeframe,location,data_Number,moe_Number,data_Percent,moe_Percent) %>%
  arrange(location)


#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(moe_output,file=paste0("../Output/health/moe/",database_state,"_",year,"_uninsuredchildrenages0to18_acstribal_moe.csv"),row.names=FALSE)

```



# STEP 3b: OUTPUT KIDS COUNT DATA CENTER FORMAT (SKIP ADDING TO DATABASE AND MARGINS OF ERROR)
#This is code that can be used by states that do not have a database to store KIDS COUNT data, but just need a .csv output to upload to the data center.
**You must add a file path and name the desired csv file to the write.csv code**
```{r}
kcdatacenter <- acs_tribal %>%
  select(c(location,dataformat,data,age_group,timeframe,locationid)) %>%
  rename(Location=location,
         DataFormat=dataformat,
         Data=data,
         TimeFrame=timeframe,
         LocationId=locationid)

#SAVE OUTPUT DATASET TO UPLOAD TO DATA CENTER
write.csv(kcdatacenter,file="your path/goes here/nameyourcsv.csv",row.names=FALSE)
```